---
output: html_document
title: "MHP Special Project - Data prep"
author: "Anne Ju Laberge"
date: "2025"
format:
  html:
    
    code-fold: true
    output-fold: true
    code-tools: true
    code-link: true
    code-summary: "Show the code"
    theme: cosmo
    embed-resources: true
    self-contained: true
    toc-depth: 5
    toc: true
    toc-float: true
editor: visual
editor_options:
   
   chunk_output_type: console
execute:
  warning: false
  error: true
---

# 0. Data

Installing the package to have access to the MHP data (Just for the first time).

```{r For MHP data}

#install.packages("usethis")

#library(usethis)
#usethis::create_github_token() # to create a Token

#gitcreds::gitcreds_set()

#devtools::install_github("MaraHyenaProject/hyenadata", auth_token = gh::gh_token())

```

Load packages:

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

```

```{r Load MHP tables}
#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```

# 1. Create data

## 1.1. Communal den dataframe

I will be using the tblLandmarks to identified CDs' locations (GPS points). Maybe also tblSessions/tblLocationsPerSession to ID sessions at communal dens and then have the location with the time/year.

```{r Location data}
str(tblLandmarks)

CD_loc_data <- tblLandmarks |>
  subset(category == "cd" | category == "d") |>
  subset(region == "talek" | region == "serena") |>
  filter(landmark_id != "Car Park Den") |> # This coordinate is the same as utme, this is wrong... also wrong in the raw data... will have to look on the GPS itself
  select(landmark_id, utme, utmn, date) |>
  drop_na(utme, utmn)


CD_loc_data_date <- CD_loc_data |>
  separate(date, c("date1", "date2"), sep = " ", remove=FALSE) |>
  separate(date2, c("year", "month", "day"), sep = "-", remove = F) |>
  separate(date1, c("month1", "day1", "year1"), sep = "/", remove = F)

```

I gave up... The date column was too messed up so I exported it into a csv file, added the year column by hand and now I just reimport it here:

```{r Location data}
CD_loc_data <- read.csv2("data_files/CD_loc_data.csv")

CD_loc_data <- CD_loc_data |>
  select(-date)

```

```{r Session+Location data}

Sess_data <- tblSessions |>
  select(clan, session) |>
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra"))

CD_SessLoc_data <- tblLocationsPerSession |>
  subset(str_detect(landmark_id, "den") | str_detect(landmark_id, "Den"))

CD_data <- merge(Sess_data, CD_SessLoc_data, by = "session")

```

Great, now I have all my Den sessions in Talek. I need to fill the utme and utmn columns so all of them of a specific locations. I can try to do that with my CD_loc_data database created earlier here.

I want to match the landmark_id columns in both database and when it does, to add the utme and utmn values in the corresponding columns.

```{r Full DB}

CD_sess_data <- CD_data |>
  select(-c(3:5, 7:11))

# merge the utme and utmn columns
CD_data_full <- merge(CD_sess_data, CD_loc_data, by = "landmark_id")

str(CD_data_full)

```

```{r DB for mapping}

# First we need to turn the coordinates into numeric values
# here all the weird coordinates (the ones with multiple numbers that doesn't make sense) are transformed to NAs
CD_data_full$utme <- as.numeric(CD_data_full$utme)
CD_data_full$utmn <- as.numeric(CD_data_full$utmn)


#Just keeping the coordinates for the map
CD_for_map <- CD_data_full |>
  select(utme, utmn, landmark_id, clan, year) |>
  #create year column
  #mutate(year = year(date)) |> no need anymore
  
  # Getting rid of any year prior to the study time frame
  filter(year >= "1988") |> 
  
  #changing talek.w into talek
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  
  #Add a month column
  #mutate(month = month(date)) |> not gonna do that anymore
  
  #Put "_" in the landmark_id columns
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  #Keep only 1 distinct coordinates per den name
  distinct(landmark_id, year, .keep_all = T) |>
  
  #Getting rid of utmn coordinate with more or less than 7 numbers
  mutate(utmn = ifelse(nchar(as.character(utmn)) == 7, utmn, NA)) |>
  #Getting rid of utme coordinate with more or less than 7 numbers
  mutate(utme = ifelse(nchar(as.character(utme)) == 6, utme, NA)) |>
# for these two, we will change that eventuallly but for now we get rid of them
  
  #Getting rid of NAs
  drop_na(utme, utmn) |>
  filter(utme != 0) |>
  #removing the "." in the clan names
  mutate(clan = str_replace_all(clan, "\\.", ""))

CD_for_map$year <- as.character(CD_for_map$year)

str(CD_for_map)

```




## 1.1. All GPS locations

We will be using the GPS locations of all the points that hyenas were seen. We will use that to create KDEs and then create the random points to run the RSF.

So I have the table hyenas per session, which has a column for hyena IDs and session IDs. I will keep just these two rows, and filter to keep only serena's sessions and talek (not fig tree, pond, etc.).
I will also add the individual's clan with tlbHyenas.

```{r Sessions with hyenas}

DB_hyenas <- tblHyenasPerSession |>
  select(session, id, date) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session)) |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F)
  



# before adding the clan to the hyenas ids, I want to make sure each id have only 1 clan associated (see session S115986)

# before that, we will identify the ids that are duplicated because its hyenas born in one clan who immigrated in another one :

dupe_rows_indID <- tblClanMembership |>
  select(id, clan, status, notes) |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "unknown") |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "kcm") |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  ungroup() |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |>
  filter(id %in% id[duplicated(id)]) |>
  select(-notes) |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  slice(2) |>
  ungroup()


length(unique(dupe_rows_indID$id)) #20

```



Slicing the hyenas with the correct clan :

```{r Sessions with hyenas}

DB_ind_clan <- tblClanMembership |> # there are 3097 hyenas
  select(id, clan, status) |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |> # now we have 2061 hyenas
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |> 
  distinct() |>
  # to keep the second row of the duplicated ids
  group_by(id) |>
  filter(n() == 1 | row_number() == 2) |>
  ungroup()


length(unique(DB_ind_clan$id)) #1966 ind

dupes_diff_clan <- DB_ind_clan %>%
  group_by(id) %>%
  filter(n_distinct(clan) > 1) %>%
  ungroup()




```


```{r Sessions with hyenas}

# Then I want to add the clan for each individual 
DB_hyenas_clan <- DB_hyenas |>
  left_join(DB_ind_clan %>% select(id, clan), by = "id", relationship = "many-to-many") |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  #changing talek.w into talek
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  #removing the "." in the clan names
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  group_by(session, clan, year) |>
  summarise(
    hyenas = paste(unique(id), collapse = ","),
    .groups = "drop")

# then, let's get rid of the NAs in hyenas_id:
DB_hyenas_clan <- DB_hyenas_clan |>
  drop_na(hyenas) # No NA





# So, all this time I had a dataset with the sessions ID, clan, hyenas ID and date:

DB_session_hyenasID <- tblSessions |>
  select(clan, date, session, location, hyenas) |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-month, -day, -date) |>
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session))

# then, let's get rid of the NAs in hyenas_id:
DB_session_hyenasID <- DB_session_hyenasID |>
  drop_na(hyenas) # from 151719 to 122558 (difference of 29161)


```




I have more lines in DB_session_hyenasID (122558) than in DB_hyenas_clan (128 324). The difference is 5766. Let's see what are the differences:
```{r}

extra_rows <- DB_session_hyenasID |>
  distinct(session, year, hyenas, clan) |>
  anti_join(DB_hyenas_clan |>
              distinct(session, year, hyenas, clan),
            by = c("session", "year", "hyenas", "clan"))
#This is 6038


extra_sessions <- DB_session_hyenasID |>
  distinct(session) |>
  anti_join(DB_hyenas_clan %>% distinct(session), by = "session")
#this is 1340


extra_rows <- DB_session_hyenasID |>
  distinct(session, year, clan) |>
  anti_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#1791

common_rows <- DB_session_hyenasID %>%
  distinct(session, year, clan) %>%
  inner_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#120 767

```




Now that we have the sessions with all the hyenas that were seen in each of our 3 territories, we want to add the GPS location of those sessions.
We will use tblLocationsPerSession.

```{r Adding GPS location}

DB_all_locations <- DB_session_hyenasID |>
  left_join(tblLocationsPerSession %>% select(session, landmark_id, utme, utmn), by = "session", relationship = "many-to-many") |>
  #take the first row out
  group_by(session) |>
  slice(1) |>
  ungroup()

dupe_rows <- DB_session_hyenasID %>%
  filter(session %in% session[duplicated(session)])# no duplicates


```


Now we want to fill the NAs with the landmark ID GPS point. 

```{r Filling the NAs}

# first create a dataframe with all the coordinates for the landmarks:

DF_lndmk <- tblLandmarks |>
  select(landmark_id, utme, utmn) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "Marsh_Den" & (is.na(utme) | is.na(utmn))))


# should be 75271 rows:
DB_locations_for_MCPs <- DB_all_locations |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "~NA" & is.na(utme) & is.na(utmn))) |>
  filter(!(is.na(landmark_id) & is.na(utme) & is.na(utmn))) |>
  
  #changing the names that doesn't match:
  mutate(landmark_id = case_when(
    landmark_id == "Airstrip_D" ~ "Airstrip_Den",
    landmark_id == "Maushugana_D_11" ~ "Mashugana_Den_II",
    landmark_id == "Arba_Den" ~ "ARBA_Natal_Den",  
    landmark_id == "Banana_T" ~ "Banana_Tree",
    landmark_id == "upro_d" ~ "Uproar_Den",   
    landmark_id == "Uproar_D" ~ "Uproar_Den", 
    landmark_id == "cop_c" ~ "Copulation_Culvert", #verify
    landmark_id == "Coral_Rf" ~ "Coral_Reef",
    landmark_id == "Border_Lugga" ~ "South_Border_Lugga",
    landmark_id == "btn_Mystery_Spot_&_Talek_Gate_on_basecamp_plain" ~ "Mystery_Spot",
    landmark_id == "Camel_X" ~ "Camel_Crossing",
    landmark_id == "Century_X" ~ "Century_Crossing",
    landmark_id == "Crtn_Edge_Den" ~ "Croton_Edge_Communal_Den",
    landmark_id == "Dave_D" ~ "Dave_Den",
    landmark_id == "Fever_T_ND/Cd_Cplx" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "FFL_Wrap" ~ "Falafel_Wrap",
    landmark_id == "Hat_T" ~ "Hat_Tree",
    landmark_id == "K-6_Den" ~ "K6_Den",
    landmark_id == "Leopard_Sex_X" ~ "Leopard_Sex_Crossing",
    landmark_id == "MGTA_ND_13" ~ "MGTA_ND13",
    landmark_id == "Paula`s_Pond" ~ "Paula's_Pond",
    landmark_id == "Plant_D" ~ "Plantation_Den",
    landmark_id == "Pole_T" ~ "Pole_Tree",
    landmark_id == "Precipice_Den_3" ~ "Precipice_Den_III",
    landmark_id == "S_Lodge" ~ "Serena_Lodge",
    landmark_id == "Sbre_Gtwy_X" ~ "Sabre_Gateway_X",
    landmark_id == "Sherbert_D" ~ "Sherbert_Den",
    landmark_id == "Top_T_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "TT_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "Vulture_X" ~ "Vulture_Crossing",
    landmark_id == "Central_Park_CD2" ~ "Central_Park_Communal_Den",
    landmark_id == "Rbend_D" ~ "Riverbend_Den",
    landmark_id == "Leopard_X" ~ "Leopard_Crossing",
    landmark_id == "Talek_L_X" ~ "Talek_Lugga_Crossing",
    landmark_id == "T_Marsh" ~ "Talek_Marsh",
    landmark_id == "Border_T" ~ "Border_Tree",
    landmark_id == "Stupid_T" ~ "Stupid_Tree",
    landmark_id == "Dave_L_X" ~ "Dave's_Lugga_Crossing",
    landmark_id == "Top_FT_L" ~ "Fig_Tree_Lugga_-_Top",
    landmark_id == "Aardwolf_Rd_T" ~ "Aardwolf_Trees",
    landmark_id == "Diamond_X" ~ "Diamond_Crossing",
    landmark_id == "Gomer's_D" ~ "Gomer_den",
    landmark_id == "Lion_L" ~ "Lion_Lugga",
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing", 

    TRUE ~ landmark_id   # keep everything else unchanged
  )) |>
  
  # Filling up the NAs
  left_join(DF_lndmk |>
      select(landmark_id, utme_df2 = utme, utmn_df2 = utmn) |>
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2) |>
  
   #getting rid of landmark that I have no idea what they are..
  filter(landmark_id != "Martins_Den") |>
  filter(landmark_id != "Kit_Kat_Den_2") |>
  filter(landmark_id != "ccb") |>
  filter(landmark_id != "Precipice_Den_3") |>
  filter(landmark_id != "Albatross_Den_2") |>
  filter(landmark_id != "corner_of_baboon_culdesec") |># double check for Cul_de_sac
  filter(landmark_id != "Baboon_Culdesac") |>
  filter(landmark_id != "Dave_Den") |>
  filter(landmark_id != "Jackal_Hill") |>
  filter(landmark_id != "KCM_D") |>
  filter(landmark_id != "Main_DOC_X") |># we have that location for sure --> verify
  filter(landmark_id != "MB_X") |>
  filter(landmark_id != "Mteer_D" ) |>#Maybe 3 musketeer?? 
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "No_Name_Hill_Culdesac") |>
  filter(landmark_id != "Picasso_Shrub") |>
  filter(landmark_id != "Pond_Crossing") |>
  filter(landmark_id != "Top_???") |>
  filter(landmark_id != "Top_Bane_Lugga") |>
  filter(landmark_id != "Top_Bane_L") |>
  filter(landmark_id != "Eland_Hill") |>
  filter(landmark_id != "676") |>  # no utme and utmn for this in DF_lndmk
  filter(landmark_id != "13_Lugga_South") |>
  filter(landmark_id != "Base_of_Jackel_Hill_&_top_of_Den_One_Creek") |>
  filter(landmark_id != "Den_Six") |>
  filter(landmark_id != "HML_ND") |>
  filter(landmark_id != "JH_T") |>
  filter(landmark_id != "kindt") |>
  filter(landmark_id != "NCCPA") |>
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "Pilot's_D") |>
  filter(landmark_id != "Top_E_Fork_HS_L")

  

# important to remember those!! it's probably the sessions in 1988-2000 where the GPS locations was not associated to the sessions.
DB_NAs <- DB_all_locations |>
  filter(if_all(c(landmark_id, utme, utmn), is.na)) #51 267 obs

```


### 1.1.2. Cleaning coordinates

Now we have to deal with the weird coordinates that will create fuckall points.

ex: Reach_Out_and_Touch_Mound
```{r Weird coordinates}
DB_locations_for_MCPs_clean <- DB_locations_for_MCPs |>
  
  left_join(DF_lndmk %>%
      select(landmark_id, utme_new = utme, utmn_new = utmn) %>%
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  
  # fix utme/utmn when they have placeholder values (11111, 111111, 1111111)
  mutate(
    utme = if_else(utme %in% c("11111", "111111", "1111111"), utme_new, utme),
    utmn = if_else(utmn %in% c("11111", "111111", "1111111"), utmn_new, utmn)) |>
  
  # fix utme/utmn when utmn is "1111111"
  mutate(
    utme = if_else(utmn == "1111111", utme_new, utme),
    utmn = if_else(utmn == "1111111", utmn_new, utmn)) |>
  
  # fix utme/utmn when both are 0 for specific landmarks
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utme_new, utme),
    utmn = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utmn_new, utmn)) |>
  
  # fix specific 98... value for Fig_Tree_Lugga_-_Top
  mutate(
    utme = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utme_new, utme),
    utmn = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utmn_new, utmn)) |>
  
  # drop the helper columns
  select(-utme_new, -utmn_new) |>
  
  # change the airstrip_den utme and utmn value with a comma
  mutate(
    utme = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utme)),   # take everything after the comma
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>
  
 # change the session t99585 and t60813 and swap 4 and 7 in utme
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Green_Den") & grepl("^47", utme),
      sub("^47", "74", utme),
      utme)) |>
  
  #same for session t116873, t116658 and t54139 (change 5 and 7)
   mutate(
    utme = if_else(
      landmark_id %in% c("Croton_Island-East_Edge", "Cow_Tree", "Hi_Step_X") & grepl("^57", utme),
      sub("^57", "75", utme),
      utme)) |>

  #remove the last digit of a 7 digit utme
  mutate(
  utme = if_else(
    landmark_id %in% c("Big_Sky_Den", "The_Observatory", "BB_X") & nchar(utme) == 7,
    substr(utme, 1, 6),   # keep only first 6 digits
    utme)) |>
  
  #for session t117205, add a 4 after the first 7, and session t111704 add a 5 after the 7
  mutate(
    utme = if_else(
      landmark_id == "Laura_Tree" & utme == "79731",
      sub("^7", "74", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Sunshine_Tree" & utme == "70360",
      sub("^7", "75", utme),
      utme)) |>
  
  # remove the 2nd number 
  mutate(
  utme = if_else(
    session %in% c("t112413", "t51456"),
    paste0(substr(utme, 1, 1), substr(utme, 3, nchar(utme))),  # remove 2nd character
    utme)) |>
  
  # remove all that is after the coma for both utme and utmn
  mutate(
    across(
      c(utme, utmn),
      ~ if_else(
        landmark_id %in% c("Serena_Lodge", "Box_Crossing", "Surprise_X"),
        sub(",.*", "", .x),   # keep only before first comma
        .x))) |>
  
  # For this session "t49599" 
  # 745095 8906 983
  mutate(
    utme = if_else(
      session == "t49599",
      substr(utme, 1, 6),   # keep only the first 6 characters
      utme)) |>
  mutate(
    utmn = if_else(
      session == "t49599",
      paste0("9806", utmn),   # add 9806 before in utmn
      utmn)) |>
  
  # keep just after the 2nd comma (t22224)
  mutate(
    utme = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utme),   # keep everything after 2nd comma
      utme),
    utmn = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utmn),
      utmn)) |>
  
  #Get rid of the last digit
   mutate(
    utme = if_else(
      session %in% c("t47831", "t49436", "t48055", "t111641", "t72442", "t57018", "t120547", "t115642"),   # replace with your session IDs
      substr(utme, 1, nchar(utme) - 1),  # drop last character
      utme)) |>
  
  # Now these have only 5 digits in utme, i will just remove them
  filter(!session %in% c("t115558", "t49400", "t66664", "t74990")) |> #I will see later if I decide to look for the last digit or not 
  
  #take out the sessions with weird utmn (99 and 0)
  filter(session != "t99428") |>
  filter(session != "t54706") |>
  #take out the Low_Ranker's_Crossing cause utmn= utme
  filter(landmark_id != "Low_Ranker's_Crossing") |>
  
  # Keep only before the comma 
  mutate(
    utmn = if_else(
      landmark_id %in% c("Uproar_Den", "Fig_Tree_Lodge", "Virginia_Maria_Crossing", "Three_Sisters", "Green_Den"),  #t60159
      sub(",.*", "", utmn),   # keep before the first comma
      utmn)) |>
  
  # Now keep all after the comma
  mutate(
    utmn = if_else(
      landmark_id %in% c("Swakey_Tree"),
      sub("^[^,]*,", "", utmn),   # remove everything up to the first comma
      utmn)) |>
  
  # remove all utmn of 6 and 8
  filter(nchar(utmn) != 6) |>
  filter(nchar(utmn) != 8) |>

  #change the utmn starting with 6 or 7 to 9
  mutate(
    utmn = if_else(
      grepl("^[67]", utmn),   # starts with 6 or 7
      sub("^[67]", "9", utmn),
      utmn))
  
```





To get utme and utmn values of + or - than 6 numbers:
```{r Check}

DB_check_utme <- DB_locations_for_MCPs_clean |>
  filter(nchar(utme) != 6) # now it is 0

DB_check_utmn <- DB_locations_for_MCPs_clean |>
  filter(nchar(utmn) != 7) # now is 0

```




### 1.1.3. Outliers

Now we visualize to see if there are still weird outliers points:
```{r Visualize}
# first convert in sf object :
DB_locations_for_MCPs_sf <- st_as_sf(DB_locations_for_MCPs_clean, coords = c("utme", "utmn"), crs = 32736)

# map to visualize
mapview(DB_locations_for_MCPs_sf, xcol = "utme", ycol = "utmn")
# this is to see the outliers

```

The points that looks like outliers:
- t120888
- s58
- t55969
- t112359
- s6974
- s115706
- s117340 


We remove those outliers points:
```{r Outliers}

#in sf objects
DB_AllGPS_sf_clean <- DB_locations_for_MCPs_sf |>
  select(-hyenas) |>
  filter(!session %in% c("t120888", "s58", "t55969", "t112359", "s6974", "s115706", "s117340")) 


# normal dataframe
DB_AllGPS <- DB_locations_for_MCPs_clean |>
  select(-hyenas) |>
  filter(!session %in% c("t120888", "s58", "t55969", "t112359", "s6974", "s115706", "s117340"))
  
  
```



# 2. Save and Shapefiles

SHP file contains geometrical information (for example shape type and coordinates). DBF file contains additional attribute information (for example name or temperature)

We use St_as_sf() function from sf package to create an sf object directly from a data frame containing a column of type sfc (simple feature list-column). Here, the utme and utmn are coordinates and so can be transformed into an sfc column.

You can also create and sf object with st_sf, where you will have a sfc object stored in an r object and then its attributes and you can combien them into an sf object : "pts_sf \<- st_sf(pts_data, geometry = pts)" where pts = sfc of coordinates pts_data = attributes

## 2.1. Save data frame

```{r Save}

write_csv2(CD_for_map, "data_files/CD_for_map.csv")
write_csv2(DB_AllGPS, "data_files/DB_AllGPS.csv")

```




## 2.2. Shapefiles

This is the locations of all the hyenas seen in the 4 territories:
```{r All GPS}

# outlier:
DB_AllGPS_sf_clean <- DB_AllGPS_sf_clean |>
  filter(landmark_id != "Migration_Den")

# Saving it:
DB_AllGPS_sf <- sf::st_write(DB_AllGPS_sf_clean, "data_files/DB_AllGPS_sf.shp")

```






THis is with only the den locations:
```{r shp in UTM}
# Turning the points into coordinates to put in a map

#Convert the dataframe into a sf object:
CD_sf <- st_as_sf(CD_for_map, coords = c("utme", "utmn"), crs = 32736) #EPSG code is 32736, UTM Zone 36 South

#outlier
CD_sf <- CD_sf |>
  filter(landmark_id != "Migration_Den")

#Save the sf dataframe - can only do it once when it is not created before
CD_sf <- sf::st_write(CD_sf, "data_files/CD_sf.shp")

```

Now, CD_sf is a sf object because we have the feature geometrics (POINTS here) with their related attributes (year, landmark_id, clan, etc.).

```{r}

str(CD_sf)
# our sf object is a list of POINTS

# To extract the specific item of the list of points
CD_sf$geometry[[2]] # give: POINT (717702 9849474)

class(CD_sf$geometry[[2]]) # give : [1] "XY"    "POINT" "sfg"  

#To change the name of the "geometry column": (don'T use names() function, but st_geometry)
#st_geometry(CD_sf) <- "Den_location"

```


## 2.3. WGS84 transform

Also, to use leaflet (to make maps), we need to change the UTM coordinates into long lat : Here we transform the CD_sf into a WGS84 to use it with leaflet.


```{r All GSP}
# transform in WGS84
DB_AllGPS_sf_wgs84 <- st_transform(DB_AllGPS_sf, crs = 4326)

# outlier:
DB_AllGPS_sf_wgs84 <- DB_AllGPS_sf_wgs84 |>
  filter(landmark_id != "Migration_Den")

# saving it:
DB_AllGPS_sf_wgs84 <- sf::st_write(DB_AllGPS_sf_wgs84, "data_files/DB_AllGPS_sf_wgs84.shp")

```



```{r CD data}

CD_sf_wgs84 <- st_transform(CD_sf, crs = 4326) #4326 is the EPSG code for the WGS84 system, which uses degrees of latitude and longitude.


#outlier:
CD_sf_wgs84 <- CD_sf_wgs84 |>
  filter(landmark_id != "Migration_Den")

# save the lat long data points into a shapefile 
CD_sf_wgs84 <- sf::st_write(CD_sf_wgs84, "data_files/CD_sf_wgs84.shp")

```



# 3. Prepare RSF data

Before we can run an RSF, we need to prepare the database that will be used. The first step for that is to get our used GPS points, in this case, it is the den locations that were extracted and put in `CD_for_map` or `CD_sf` (CD = Communal Den). The second step is to create the clans' MCP (Minimum Convex Polygon). We will generate the 95% MCP and 100% MCP.

With the used points and each clans MCP, we will be able to generate random points in their respective territories. I will attribute 10 random points (available points) to the used points. To do so, I will take the clans' MCPs as the home range of the clans, and generate random points in the polygons created. This will create a database of used and available points. The used points are the den locations known, and the available are all the adjacent points in the territory.

I need to determine how many points I want to sample for each points used. For example, 5 points per used points.

The `random_points()` function will create this `case\_` variable that will be use in the RSF later.

## 3.1. MCP by clans

To make our RSF model, we need to create random points of potential locations of where the hyenas could have denned. To do that, we first need to know the extent of the clans' territory, and so we will calculate the MCP (Minimum Convex Polygon) for each clan. An MCP calculate the area that an animal uses. The MCP essentially draws a polygon around all the points where an animal has been observed.

Also, we are using the 95% MCP, so that it removes the 5% of outlier points.

### 3.1.1. Individual MCPs

Here is the MCPs for all hyenas locations
```{r Clan MCPs}

# Convert all points to sp (adehabitatHR works with sp)
CD_sp <- as(DB_AllGPS_sf_wgs84, "Spatial")

# Create MCPs grouped by clan
mcp_clan <- mcp(CD_sp[, "clan"], percent = 95)

# Convert back to sf
mcp_clan_sf <- st_as_sf(mcp_clan)

# Inspect
plot(st_geometry(mcp_clan_sf), col = NA, border = "red")

# change id for clan 
mcp_clan_sf <- mcp_clan_sf |>
  rename(clan = id)

```

Save
```{r Save}

# Save the dataframe
mcp_clan_sf <- sf::st_write(mcp_clan_sf, "data_files/mcp_clan_sf.shp")

```


### 3.1.2. Yearly MCPs
```{r Yearly MCPs}

# Create an ID that combines clan and year
CD_sp$clan_year <- paste(CD_sp$clan, CD_sp$year, sep = "_")

# Count points per clan-year
counts <- as.data.frame(table(CD_sp$clan_year))

# Keep only groups with >= 5 relocations
valid_ids <- counts$Var1[counts$Freq >= 5]

# Filter points
CD_sp_valid <- CD_sp[CD_sp$clan_year %in% valid_ids, ]

# Run MCP only on valid groups
mcp_clan_year <- mcp(CD_sp_valid[, "clan_year"], percent = 95)
mcp_clan_year_sf <- st_as_sf(mcp_clan_year)



# Split IDs back into clan and year
mcp_clan_year_sf <- mcp_clan_year_sf |>
  mutate(clan = sub("_(\\d+)$", "", row.names(mcp_clan_year_sf)),
         year = sub(".*_", "", row.names(mcp_clan_year_sf)))


```

```{r Excluded vs included}

# Count relocations per clan-year
counts <- as.data.frame(table(CD_sp$clan_year)) %>%
  rename(clan_year = Var1, n_points = Freq)

# Split clan and year into separate columns
counts <- counts %>%
  mutate(
    clan = sub("_(\\d+)$", "", clan_year),
    year = sub(".*_", "", clan_year)
  )

# Show which groups are valid / invalid
counts <- counts %>%
  mutate(valid = ifelse(n_points >= 5, "yes", "no"))

# View excluded groups
excluded <- counts %>% filter(valid == "no")

# View valid groups
included <- counts %>% filter(valid == "yes")

excluded
included

```


Save
```{r Save}

#  Save the dataframe
mcp_clan_year_sf <- sf::st_write(mcp_clan_year_sf, "data_files/mcp_clan_year_sf.shp")

```


### 3.1.3. Plots

```{r Dataframe}
# Transform into a dataframe:
clan_MCPs_dataframe <- as.data.frame(mcp_clan_sf)

clan_MCPs_dataframe <- rownames_to_column(clan_MCPs_dataframe, "clan")

```


```{r Single MCP}
MCPs_clan <- unique(clan_MCPs_dataframe$clan)

MCPs_plot <- AllGPS_leaflet |>
  addPolygons(data = mcp_clan_sf,
              color = "red",
              fillColor = "none",
              weight = 2,
              fillOpacity = 0.05)
  
MCPs_plot

```


```{r Yearly MCPs}

# Start with an empty leaflet map
yearly_MCPs_plot <- leaflet() |> addTiles()

# Define your clan palette
pal2 <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

# Get all years
years2 <- sort(unique(DB_AllGPS_sf_wgs84$year))

# Loop over years to add both points and MCPs under the same baseGroup (year)
for (yr in years2) {
  
  # Points for this year
  points_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
  # MCP polygons for this year
  mcp_year <- mcp_clan_year_sf |> filter(year == yr)

  # Add MCP polygons
  yearly_MCPs_plot <- yearly_MCPs_plot |>
    addPolygons(data = mcp_year,
                color = "red",
                fillColor = "none",
                weight = 2,
                fillOpacity = 0.05,
                popup = ~paste("Year:", year, "<br>", "Clan:", clan),
                group = paste0("Year ", yr))  # same group = year
  
  # Add points
  yearly_MCPs_plot <- yearly_MCPs_plot |>
    addCircleMarkers(data = points_year,
                     color = ~pal2(clan),
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan, "<br>", "Session:", session, "<br>", "Landmark:", lndmrk_),
                     group = paste0("Year ", yr))  # group = year
}

# Add **radio button style toggle** for years
yearly_MCPs_plot <- yearly_MCPs_plot |>
  addLayersControl(
    baseGroups = paste0("Year ", years2),  # radio button style
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  addLegend("bottomright", pal = pal2, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)


yearly_MCPs_plot


```


Save
```{r}

# Save the map
htmlwidgets::saveWidget(yearly_MCPs_plot, file = "Maps/yearly_MCPs_plot.html", selfcontained = TRUE)
```


## 3.2. Random points

```{r Random_points}

#Convert your sf object to a data frame with coordinates: 

CD_sf_wgs84_df <- CD_sf_wgs84 |> mutate(long = sf::st_coordinates(geometry)[,1], lat = sf::st_coordinates(geometry)[,2]) 

#Change the CD_sf_wgs84 into a track_xy object 

presence_track <- make_track(CD_sf_wgs84_df, long, lat, crs = 4326, clan = clan, year = year) 


#serenan 
presence_track_serenan <- presence_track |> subset(clan == "serenan") 

#serenas 
presence_track_serenas <- presence_track |> subset(clan == "serenas") 

#hz 
presence_track_hz <- presence_track |> subset(clan == "happyzebra") 

#talek 
presence_track_talek <- presence_track |> subset(clan == "talek") 


# make MCP valid:
mcp_clan_sf <- st_make_valid(mcp_clan_sf)


#creating the dataset to use in RSF after (for all clans) 
rp_clans_MCP <- random_points(mcp_clan_sf, n=4000, presence = presence_track) 

#Visualising 
plot(rp_clans_MCP)

```

```{r}
# Function to generate 10 random points per used point
generate_randoms <- function(pres_row, mcp_polygons, n = 10) {
  # Extract clan from the presence row
  clan_i <- pres_row$clan
  
  # Subset MCP polygon for this clan
  mcp_i <- mcp_polygons %>% filter(clan == clan_i)
  
  # Generate n random points inside this MCP
  rand_pts <- random_points(mcp_i, n = n, presence = pres_row)
  
  # Convert to sf
  rand_sf <- st_as_sf(rand_pts, coords = c("x_", "y_"), crs = st_crs(mcp_polygons))
  
  # Add back metadata
  rand_sf <- rand_sf %>%
    mutate(clan = clan_i,
           year = pres_row$year,  # still keep year from the presence row
           case_ = 0)
  
  return(rand_sf)
}

# Apply over all presence points
CD_rp_data <- map_dfr(1:nrow(presence_track), ~ generate_randoms(presence_track[.x, ], mcp_clan_sf, n = 10))

# Combine presences (case_ = 1) and randoms (case_ = 0)
CD_all_data <- bind_rows(
  presence_track %>%
    st_as_sf(coords = c("x_", "y_"), crs = st_crs(mcp_clan_sf)) %>%
    mutate(case_ = 1),
  CD_rp_data)




```


### 3.2.1. Plots
```{r}
# visualizing
pal_case <- colorFactor(palette = c("red", "blue"),
                        domain = CD_all_data$case_)

rp_CD_map <-
leaflet() |>
  addTiles() |>

  # Then plot random (case_ == 0) in blue (will appear on top)
  addCircleMarkers(
    data = CD_all_data[CD_all_data$case_ == 0, ],
    color = "blue", radius = 3, fillOpacity = 0.6,
    group = "Random"
  ) |>
   # First plot used (case_ == 1) in red
  addCircleMarkers(
    data = CD_all_data[CD_all_data$case_ == 1, ],
    color = "red", radius = 3, fillOpacity = 0.6,
    group = "Used"
  ) |>
  addLayersControl(
    overlayGroups = c("Used", "Random"),
    options = layersControlOptions(collapsed = FALSE)
  )



```


```{r}
# Save the map
htmlwidgets::saveWidget(rp_CD_map, file = "Maps/rp_CD_map.html", selfcontained = TRUE)

```




### Save

```{r}

CD_all_data <- sf::st_write(CD_all_data, "data_files/CD_all_data.shp")

```







