---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Data preparation

# 0. Libraries et al.

Installing the package to have access to the MHP data (Just for the first time).

```{r For MHP data}

#install.packages("usethis")

#library(usethis)
#usethis::create_github_token() # to create a Token

#gitcreds::gitcreds_set()

#devtools::install_github("MaraHyenaProject/hyenadata", auth_token = gh::gh_token())

```

Load packages:

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to play with date columns
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

# For RSF
library(amt) # for the RSF
library(adehabitatHR) # for mcp
library(brms)
library(glmmTMB) # to make RSF models

```

```{r Load MHP tables}

#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```

```{r}
#write_csv2(tblSessions, "data_files/tblSessions.csv")
#write_csv2(tblLocationsPerSession, "data_files/tblLocationsPerSession.csv")

#write_csv2(tblLifeHistory.wide, "data_files/tblLifeHistory.wide.csv")
```


# 1. Create data

## 1.1. Communal dens (CD) dataframe

I will be using the tblLandmarks to identified CDs' locations (GPS points). Maybe also tblSessions/tblLocationsPerSession to ID sessions at communal dens and then have the location with the time/year.

```{r Location data}
str(tblLandmarks)

CD_loc_data <- tblLandmarks |>
  subset(category == "cd" | category == "d") |>
  subset(region == "talek" | region == "serena") |>
  filter(landmark_id != "Car Park Den") |> # This coordinate is the same as utme, this is wrong... also wrong in the raw data... will have to look on the GPS itself
  select(landmark_id, utme, utmn, date) |>
  drop_na(utme, utmn)


CD_loc_data_date <- CD_loc_data |>
  separate(date, c("date1", "date2"), sep = " ", remove=FALSE) |>
  separate(date2, c("year", "month", "day"), sep = "-", remove = F) |>
  separate(date1, c("month1", "day1", "year1"), sep = "/", remove = F)

```

I gave up... The date column was too messed up so I exported it into a csv file, added the year column by hand and now I just reimport it here:

```{r Location data}
CD_loc_data <- read.csv2("data_files/CD_loc_data.csv")

CD_loc_data <- CD_loc_data |>
  select(-date)

```

```{r Session+Location data}

Sess_data <- tblSessions |>
  select(clan, session) |>
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra"))

CD_SessLoc_data <- tblLocationsPerSession |>
  subset(str_detect(landmark_id, "den") | str_detect(landmark_id, "Den"))

CD_data <- merge(Sess_data, CD_SessLoc_data, by = "session")

```

## 1.2. CD monthly - tblHyenasPerSession

I will use tblSessions to filter out the communal dens sessions and get the month and year of use of the different CD systems.

Also, we have to make sure that the CD that are recorded in a session are active communal dens. I think that using the `location` column makes it an active den, but to verify... 


```{r with tblHyenasPerSession}
# filtering out the right clans and sessions at a CD 
HyenasPerSession_data <- tblHyenasPerSession |>
  select(id, session, date)  |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session))

length(unique(HyenasPerSession_data$session)) # 128 200

Loc_HyenasPerSess <- tblLocationsPerSession |> 
  filter(session %in% HyenasPerSession_data$session) |>
  select(session, landmark_id, utme, utmn)

length(unique(Loc_HyenasPerSess$session)) #77178

#The difference here is of 51022 obs... thats kind of a lot...
```

### 1.2.1. Verify sessions

Here I will get the difference in sessions (the 51022 sessions):
```{r Difference in sessions}

HyenasPerSession_uniqueSess <- HyenasPerSession_data |>
  distinct(session, .keep_all = T) # 128200

Loc_HyenasPerSess_uniqueSess <- Loc_HyenasPerSess |>
  distinct(session, .keep_all = T) # 77178

# the session different:
HyenasPerSession_uniqueSess_diff <- HyenasPerSession_uniqueSess |>
   filter(!session %in% Loc_HyenasPerSess_uniqueSess$session) #51022

# So it seems that it is all the sessions before 2002 (including 2002), and then some sessions throughout the whole time frame.
#Maybe they are session outside of our desired 4 clans

```

Now, I am going to use tblSessions (hopefully all the same sessions are in this and HyenasPerSession_uniqueSess) to add the clan column:
```{r Adding clan}

HyenasPerSession_uniqueSess_clan <- tblSessions |>
  select(clan, session, location, date) |>
  filter(session %in% HyenasPerSession_uniqueSess$session) #128200 --> good


#now we remerge the whole dataframe, not just the unique session:
HyenasPerSession_data_clan <- HyenasPerSession_data |>
  left_join(HyenasPerSession_uniqueSess_clan %>% select(session, clan), by = "session") |>
  #now we keep only the clans we are interested in, so that will keep only the right session :
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra")) |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  mutate(clan = str_replace_all(clan, "talek.w", "talek"))


length(unique(HyenasPerSession_data_clan$session)) # 122547


#Here we filter the location per session:
Loc_HyenasPerSess_clan <- tblLocationsPerSession |> 
  filter(session %in% HyenasPerSession_data_clan$session) |>
  select(session, landmark_id, utme, utmn)

length(unique(Loc_HyenasPerSess_clan$session)) # 71584

#Now the difference is 50963 sessions


```


Here, we redo the same uniqueSess process to see which sessions are not in LocationPerSession:
```{r}

HyenasPerSession_clan_uniqueSess <- HyenasPerSession_data_clan |>
  distinct(session, .keep_all = T) # 122547

Loc_HyenasPerSess_clan_uniqueSess <- Loc_HyenasPerSess_clan |>
  distinct(session, .keep_all = T) # 71584

# the session different:
HyenasPerSession_clan_uniqueSess_diff <- HyenasPerSession_clan_uniqueSess |>
   filter(!session %in% Loc_HyenasPerSess_clan_uniqueSess$session) #50963

```
Ok, so it is just the same thing... Still has everything from before 2002, then some random sessions throughout the whole study period.


### 1.2.2. Filter CD sess
Here we keep only the den sessions. Will do that with tblSessions
```{r}

CD_Sess <- tblSessions |>   # 26911
  select(session, location, date) |>
  filter(location == "d") |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session))

HyenasPerSession_CD_data <- HyenasPerSession_data_clan |>
  filter(session %in% CD_Sess$session)


#now the location data:
#Here we filter the location per session:
Loc_HyenasPerSess_CD <- tblLocationsPerSession |> 
  filter(session %in% HyenasPerSession_CD_data$session) |>
  select(session, landmark_id, utme, utmn)


```

Redo verification of the difference:
```{r}
HyenasPerSession_CD_data_uniqueSess <- HyenasPerSession_CD_data |>
  distinct(session, .keep_all = T) # 24085

Loc_HyenasPerSess_CD_uniqueSess <- Loc_HyenasPerSess_CD |>
  distinct(session, .keep_all = T) # 16453

# the session different:
HyenasPerSession_CD_uniqueSess_diff <- HyenasPerSession_CD_data_uniqueSess |>
   filter(!session %in% Loc_HyenasPerSess_CD_uniqueSess$session) #50963

# Here we have a difference of 7632 sessions, also possibly all from before 2002 and some throughout the time serie

```


### 1.2.3. Merge 


```{r}

HyenasLocPerSession_CD <- merge(Loc_HyenasPerSess_CD_uniqueSess, HyenasPerSession_CD_data, by = "session")

```


```{r}

HyenasLocPerSession_CD <- HyenasLocPerSession_CD |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  select(-utme, -utmn)

length(unique(HyenasLocPerSession_CD$session)) 
# 16453 sessions
```

















## 1.3. Filling utme/utmn

I will just make fit landmark_id from the tblLandmarks with the newly created CD_monthly_data.

I will use month as a proxy for season for den location, to test if the hyenas change their den location selection depending on season or not.

```{r}

Landmarks <- tblLandmarks |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_"))

# now we fill the coordinates column:

```


```{r}
#Just keeping the coordinates for the map
HyenasLocPerSession_CD_clean <- HyenasLocPerSession_CD |>
  #changing talekw into talek
  mutate(clan = str_replace_all(clan, "talekw", "talek")) |>
  
  # change the non matching landmark_id names
  mutate(landmark_id = case_when(
    landmark_id == "Airstrip_D" ~ "Airstrip_Den",
    landmark_id == "Maushugana_D_11" ~ "Mashugana_Den_II",
    landmark_id == "Arba_Den" ~ "ARBA_Natal_Den",  
    landmark_id == "upro_d" ~ "Uproar_Den",   
    landmark_id == "Uproar_D" ~ "Uproar_Den", 
    landmark_id == "Precipice_Den_3" ~ "Precipice_Den_III",
    landmark_id == "Fever_T_ND/Cd_Cplx" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "Crtn_Edge_Den" ~ "Croton_Edge_Communal_Den",
    landmark_id == "Central_Park_CD2" ~ "Central_Park_Communal_Den",
    landmark_id == "Leopard_Sex_X" ~ "Leopard_Sex_Crossing",
    landmark_id == "Rbend_D" ~ "Riverbend_Den",
    landmark_id == "Plant_D" ~ "Plantation_Den",
    landmark_id == "K-6_Den" ~ "K6_Den",
    landmark_id == "Albatross_Den_2" ~ "Albatross_Den",
    landmark_id == "Pdise_D" ~ "Paradise_Den",
    landmark_id == "Lala_D" ~ "Lala_Den",
    landmark_id == "Roadside_D" ~ "Roadside_Den",
    landmark_id == "Aardwolf_CD" ~ "Aardwolf_Communal_Den",
    landmark_id == "Banana_T_D" ~ "Banana_Tree_Communal_Den",
    landmark_id == "Alt_Aard_CD" ~ "Alternate_Aardvark_Communal_Den",
    landmark_id == "Barbie_Party_D" ~ "Barbie_Party_Den",
    landmark_id == "Base_D" ~ "Base_Natal_Den/Communal_Den",
    landmark_id == "BLVD_D" ~ "Boulevard_Den",
    landmark_id == "Cliff_D" ~ "Cliff_Den",
    landmark_id == "Croc_C_D" ~ "Crocodile_Camp_Den",
    landmark_id == "Crtn_Edge_CD" ~ "Croton_Edge_Communal_Den",
    landmark_id == "Dave's_D" ~ "Dave_Den",
    landmark_id == "Dave_D" ~ "Dave_Den",
    landmark_id == "Demo_D" ~ "Democracy_Den",
    landmark_id == "Dick_D" ~ "Dicks_Den",
    landmark_id == "DOC_D" ~ "DOC_Communal_Den",
    landmark_id == "DOC_D2" ~ "DOC_Den_2",
    landmark_id == "Fever_T_ND/CD" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "Fever_T_ND/CD_Cplex" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "	
near_Fever_T_ND/CD_Cplex_" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "Gardenia_D" ~ "Gardenia_Den",
    landmark_id == "Green_D" ~ "Green_Den",
    landmark_id == "Hday_CD" ~ "Holiday_Communal_Den",
    landmark_id == "Hday_D" ~ "Holiday_Communal_Den",
    landmark_id == "Milkyway_D" ~ "Milkyway_Den",
    landmark_id == "Nomadic_D" ~ "Nomadic_Den",
    landmark_id == "Olos_WB_ND/CD" ~ "Olosogon_Clearing_Communal_Den_Complex",
    landmark_id == "Ranger_D" ~ "Ranger_Den",
    landmark_id == "Res_D" ~ "Resurrection_Main_Den",
    landmark_id == "Res_M_D" ~ "Resurrection_Main_Den",
    landmark_id == "Shakespeare_D" ~ "Shakespeare_Den",
    landmark_id == "Singizi_D" ~ "Singizi_Den",
    landmark_id == "Sun_D" ~ "Sun_Den",
    landmark_id == "Waverly_D" ~ "Waverly_Den",
    landmark_id == "Wicked_D" ~ "Wicked_Den",
    landmark_id == "Yoplait_D" ~ "Yoplait_Den",



    TRUE ~ landmark_id   # keep everything else unchanged
  )) |>

 #getting rid of landmark that I have no idea what they are..
  filter(landmark_id != "Martins_Den") |>
  filter(landmark_id != "Kit_Kat_Den_2") |>
  filter(landmark_id != "Pilot's_D") |>
  filter(landmark_id != "KCM_D") |>
  filter(landmark_id != "~NA") |>
  filter(landmark_id != "Mteer_D" ) |> #Maybe 3 musketeer?? 
  filter(landmark_id != "NCM_D")



```




Joining utme and utmn columns:
```{r}
HyenasLocPerSession_CD_coord <- HyenasLocPerSession_CD_clean |>
left_join(Landmarks %>% select(landmark_id, utme, utmn),
            by = "landmark_id", relationship = "many-to-many")
```



#### 1.3.1. Cleaning coordinates
```{r}

HyenasLocPerSession_CD_coord_clean <- HyenasLocPerSession_CD_coord  |>

  # change the airstrip_den utme and utmn value with a comma
  mutate(
    utme = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utme)),   # take everything after the comma
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>

   # remove all that is after the coma for both utme and utmn
  mutate(
    across(
      c(utme, utmn),
      ~ if_else(
        landmark_id %in% c("Green_Den", "Governor's_Camp", "Box_Crossing"),
        sub(",.*", "", .x),   # keep only before first comma
        .x))) |>
  mutate(
    across(
      c(utme, utmn),
      ~ if_else(
        landmark_id %in% c("OK_Den"),
        sub("^[^,]*,[^,]*,", "", .x),   # remove everything up to 2nd comma
        .x))) |>
  
  # Keep only before the comma for utmn
  mutate(
    utmn = if_else(
      landmark_id %in% c("Uproar_Den", "Banana_Tree_Communal_Den", "Bloody_Awful_Den", "Three_Sisters"),
      sub(",.*", "", utmn),
      utmn)) |>
  drop_na(utme)

```



```{r}
CD_check_utme <- HyenasLocPerSession_CD_coord_clean |>
  filter(nchar(utme) != 6) # now it is 0

CD_check_utmn <- HyenasLocPerSession_CD_coord_clean |>
  filter(nchar(utmn) != 7) # now is 0
```



## 1.4. Active CD - tblLifeHistory.wide

Here I will find a way to filter out the active dens from the non active. 

Using the hyenas ID, I want to identify the individuals that were cubs at a given year. Then, if they were cubs and seen at a den, I will consider the den as a communal den that year. So first, I want to create a dataframe with all individual IDs and their data of birth (DOB) including the year.


```{r}

Life_history_data <- tblLifeHistory.wide |>
  select(id, dob, dengrad) |>
  mutate(dengrad = as.character(dengrad),
         dengrad = ifelse(id == "bard", "2018-02-20", dengrad), 
         dengrad = ifelse(id == "lgnd", "2017-01-19", dengrad),
         dengrad = as.Date(dengrad)) |>
  mutate(dengrad = ifelse(is.na(dengrad), dob + years(1), dengrad),
         dengrad = as.Date(dengrad)) |># to make sure to keep it in yy-mm-dd
  drop_na(dob)

```


### 1.4.1. ActiveCD_data
```{r}

ActiveCD_PerHyenas_data <- HyenasLocPerSession_CD_coord_clean |>
  #keep only the landmark_id with den in it
  subset(str_detect(landmark_id, "den") | str_detect(landmark_id, "Den")) |>
  left_join(Life_history_data %>% select(id, dob, dengrad), by = "id") %>%
  filter(date >= dob & date <= dengrad)



ActiveCD_data <- ActiveCD_PerHyenas_data |>
  select(-id) |>
  distinct(session, .keep_all = T)


length(unique(ActiveCD_data$landmark_id))
#355 dens!!
```



### 1.4.2. Active CD year
```{r}
ActiveCD_year <- ActiveCD_data |>
  select(-month, -day, -dob, -dengrad) |>
  distinct(landmark_id, year, .keep_all = T)
# 628 dens used in total 
```



### 1.4.3. Active CD unique
```{r}
ActiveCD_unique <- ActiveCD_data |>
  select(-date, -year, -month, -day) |>
  distinct(landmark_id, .keep_all = T)
#355 dens in total (no reuse)
```





## 1.5. All GPS locations

We will be using the GPS locations of all the points that hyenas were seen. We will use that to create KDEs and then create the random points to run the RSF.

So I have the table hyenas per session, which has a column for hyena IDs and session IDs. I will keep just these two rows, and filter to keep only serena's sessions and talek (not fig tree, pond, etc.).
I will also add the individual's clan with tlbHyenas.

```{r Sessions with hyenas}

DB_hyenas <- tblHyenasPerSession |>
  select(session, id, date) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session)) |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F)
  



# before adding the clan to the hyenas ids, I want to make sure each id have only 1 clan associated (see session S115986)

# before that, we will identify the ids that are duplicated because its hyenas born in one clan who immigrated in another one :

dupe_rows_indID <- tblClanMembership |>
  select(id, clan, status, notes) |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "unknown") |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "kcm") |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  ungroup() |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |>
  filter(id %in% id[duplicated(id)]) |>
  select(-notes) |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  slice(2) |>
  ungroup()


length(unique(dupe_rows_indID$id)) #20

```



Slicing the hyenas with the correct clan :

```{r Sessions with hyenas}

DB_ind_clan <- tblClanMembership |> # there are 3097 hyenas
  select(id, clan, status) |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |> # now we have 2061 hyenas
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |> 
  distinct() |>
  # to keep the second row of the duplicated ids
  group_by(id) |>
  filter(n() == 1 | row_number() == 2) |>
  ungroup()


length(unique(DB_ind_clan$id)) #1966 ind

dupes_diff_clan <- DB_ind_clan %>%
  group_by(id) %>%
  filter(n_distinct(clan) > 1) %>%
  ungroup()




```


```{r Sessions with hyenas}

# Then I want to add the clan for each individual 
DB_hyenas_clan <- DB_hyenas |>
  left_join(DB_ind_clan %>% select(id, clan), by = "id", relationship = "many-to-many") |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  #changing talek.w into talek
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  #removing the "." in the clan names
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  group_by(session, clan, year) |>
  summarise(
    hyenas = paste(unique(id), collapse = ","),
    .groups = "drop")

# then, let's get rid of the NAs in hyenas_id:
DB_hyenas_clan <- DB_hyenas_clan |>
  drop_na(hyenas) # No NA





# So, all this time I had a dataset with the sessions ID, clan, hyenas ID and date:

DB_session_hyenasID <- tblSessions |>
  select(clan, date, session, location, hyenas) |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-month, -day, -date) |>
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session))

# then, let's get rid of the NAs in hyenas_id:
DB_session_hyenasID <- DB_session_hyenasID |>
  drop_na(hyenas) # from 151719 to 122558 (difference of 29161)


```




I have more lines in DB_session_hyenasID (122558) than in DB_hyenas_clan (128 324). The difference is 5766. Let's see what are the differences:
```{r}

extra_rows <- DB_session_hyenasID |>
  distinct(session, year, hyenas, clan) |>
  anti_join(DB_hyenas_clan |>
              distinct(session, year, hyenas, clan),
            by = c("session", "year", "hyenas", "clan"))
#This is 6038


extra_sessions <- DB_session_hyenasID |>
  distinct(session) |>
  anti_join(DB_hyenas_clan %>% distinct(session), by = "session")
#this is 1340


extra_rows <- DB_session_hyenasID |>
  distinct(session, year, clan) |>
  anti_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#1791

common_rows <- DB_session_hyenasID %>%
  distinct(session, year, clan) %>%
  inner_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#120 767

```




Now that we have the sessions with all the hyenas that were seen in each of our 3 territories, we want to add the GPS location of those sessions.
We will use tblLocationsPerSession.

```{r Adding GPS location}

DB_all_locations <- DB_session_hyenasID |>
  left_join(tblLocationsPerSession %>% select(session, landmark_id, utme, utmn), by = "session", relationship = "many-to-many") |>
  #take the first row out
  group_by(session) |>
  slice(1) |>
  ungroup()

dupe_rows <- DB_session_hyenasID %>%
  filter(session %in% session[duplicated(session)])# no duplicates


```


Now we want to fill the NAs with the landmark ID GPS point. 

```{r Filling the NAs}

# first create a dataframe with all the coordinates for the landmarks:

DF_lndmk <- tblLandmarks |>
  select(landmark_id, utme, utmn) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "Marsh_Den" & (is.na(utme) | is.na(utmn))))


# should be 75271 rows:
DB_locations_for_MCPs <- DB_all_locations |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "~NA" & is.na(utme) & is.na(utmn))) |>
  filter(!(is.na(landmark_id) & is.na(utme) & is.na(utmn))) |>
  
  #changing the names that doesn't match:
  mutate(landmark_id = case_when(
    landmark_id == "Airstrip_D" ~ "Airstrip_Den",
    landmark_id == "Maushugana_D_11" ~ "Mashugana_Den_II",
    landmark_id == "Arba_Den" ~ "ARBA_Natal_Den",  
    landmark_id == "Banana_T" ~ "Banana_Tree",
    landmark_id == "upro_d" ~ "Uproar_Den",   
    landmark_id == "Uproar_D" ~ "Uproar_Den", 
    landmark_id == "cop_c" ~ "Copulation_Culvert", #verify
    landmark_id == "Coral_Rf" ~ "Coral_Reef",
    landmark_id == "Border_Lugga" ~ "South_Border_Lugga",
    landmark_id == "btn_Mystery_Spot_&_Talek_Gate_on_basecamp_plain" ~ "Mystery_Spot",
    landmark_id == "Camel_X" ~ "Camel_Crossing",
    landmark_id == "Century_X" ~ "Century_Crossing",
    landmark_id == "Crtn_Edge_Den" ~ "Croton_Edge_Communal_Den",
    landmark_id == "Dave_D" ~ "Dave_Den",
    landmark_id == "Fever_T_ND/Cd_Cplx" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "FFL_Wrap" ~ "Falafel_Wrap",
    landmark_id == "Hat_T" ~ "Hat_Tree",
    landmark_id == "K-6_Den" ~ "K6_Den",
    landmark_id == "Leopard_Sex_X" ~ "Leopard_Sex_Crossing",
    landmark_id == "MGTA_ND_13" ~ "MGTA_ND13",
    landmark_id == "Paula`s_Pond" ~ "Paula's_Pond",
    landmark_id == "Plant_D" ~ "Plantation_Den",
    landmark_id == "Pole_T" ~ "Pole_Tree",
    landmark_id == "Precipice_Den_3" ~ "Precipice_Den_III",
    landmark_id == "S_Lodge" ~ "Serena_Lodge",
    landmark_id == "Sbre_Gtwy_X" ~ "Sabre_Gateway_X",
    landmark_id == "Sherbert_D" ~ "Sherbert_Den",
    landmark_id == "Top_T_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "TT_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "Vulture_X" ~ "Vulture_Crossing",
    landmark_id == "Central_Park_CD2" ~ "Central_Park_Communal_Den",
    landmark_id == "Rbend_D" ~ "Riverbend_Den",
    landmark_id == "Leopard_X" ~ "Leopard_Crossing",
    landmark_id == "Talek_L_X" ~ "Talek_Lugga_Crossing",
    landmark_id == "T_Marsh" ~ "Talek_Marsh",
    landmark_id == "Border_T" ~ "Border_Tree",
    landmark_id == "Stupid_T" ~ "Stupid_Tree",
    landmark_id == "Dave_L_X" ~ "Dave's_Lugga_Crossing",
    landmark_id == "Top_FT_L" ~ "Fig_Tree_Lugga_-_Top",
    landmark_id == "Aardwolf_Rd_T" ~ "Aardwolf_Trees",
    landmark_id == "Diamond_X" ~ "Diamond_Crossing",
    landmark_id == "Gomer's_D" ~ "Gomer_den",
    landmark_id == "Lion_L" ~ "Lion_Lugga",
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing", 

    TRUE ~ landmark_id   # keep everything else unchanged
  )) |>
  
  # Filling up the NAs
  left_join(DF_lndmk |>
      select(landmark_id, utme_df2 = utme, utmn_df2 = utmn) |>
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2) |>
  
   #getting rid of landmark that I have no idea what they are..
  filter(landmark_id != "Martins_Den") |>
  filter(landmark_id != "Kit_Kat_Den_2") |>
  filter(landmark_id != "ccb") |>
  filter(landmark_id != "Precipice_Den_3") |>
  filter(landmark_id != "Albatross_Den_2") |>
  filter(landmark_id != "corner_of_baboon_culdesec") |># double check for Cul_de_sac
  filter(landmark_id != "Baboon_Culdesac") |>
  filter(landmark_id != "Dave_Den") |>
  filter(landmark_id != "Jackal_Hill") |>
  filter(landmark_id != "KCM_D") |>
  filter(landmark_id != "Main_DOC_X") |># we have that location for sure --> verify
  filter(landmark_id != "MB_X") |>
  filter(landmark_id != "Mteer_D" ) |>#Maybe 3 musketeer?? 
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "No_Name_Hill_Culdesac") |>
  filter(landmark_id != "Picasso_Shrub") |>
  filter(landmark_id != "Pond_Crossing") |>
  filter(landmark_id != "Top_???") |>
  filter(landmark_id != "Top_Bane_Lugga") |>
  filter(landmark_id != "Top_Bane_L") |>
  filter(landmark_id != "Eland_Hill") |>
  filter(landmark_id != "676") |>  # no utme and utmn for this in DF_lndmk
  filter(landmark_id != "13_Lugga_South") |>
  filter(landmark_id != "Base_of_Jackel_Hill_&_top_of_Den_One_Creek") |>
  filter(landmark_id != "Den_Six") |>
  filter(landmark_id != "HML_ND") |>
  filter(landmark_id != "JH_T") |>
  filter(landmark_id != "kindt") |>
  filter(landmark_id != "NCCPA") |>
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "Pilot's_D") |>
  filter(landmark_id != "Top_E_Fork_HS_L")

  

# important to remember those!! it's probably the sessions in 1988-2000 where the GPS locations was not associated to the sessions.
DB_NAs <- DB_all_locations |>
  filter(if_all(c(landmark_id, utme, utmn), is.na)) #51 267 obs

```


### 1.5.1. Cleaning coordinates

Now we have to deal with the weird coordinates that will create fuckall points.

ex: Reach_Out_and_Touch_Mound
```{r Weird coordinates}
DB_locations_for_MCPs_clean <- DB_locations_for_MCPs |>
  
  left_join(DF_lndmk %>%
      select(landmark_id, utme_new = utme, utmn_new = utmn) %>%
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  
  # fix utme/utmn when they have placeholder values (11111, 111111, 1111111)
  mutate(
    utme = if_else(utme %in% c("11111", "111111", "1111111"), utme_new, utme),
    utmn = if_else(utmn %in% c("11111", "111111", "1111111"), utmn_new, utmn)) |>
  
  # fix utme/utmn when utmn is "1111111"
  mutate(
    utme = if_else(utmn == "1111111", utme_new, utme),
    utmn = if_else(utmn == "1111111", utmn_new, utmn)) |>
  
  # fix utme/utmn when both are 0 for specific landmarks
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utme_new, utme),
    utmn = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utmn_new, utmn)) |>
  
  # fix specific 98... value for Fig_Tree_Lugga_-_Top
  mutate(
    utme = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utme_new, utme),
    utmn = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utmn_new, utmn)) |>
  
  # drop the helper columns
  select(-utme_new, -utmn_new) |>
  
  # change the airstrip_den utme and utmn value with a comma
  mutate(
    utme = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utme)),   # take everything after the comma
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>
  
 # change the session t99585 and t60813 and swap 4 and 7 in utme
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Green_Den") & grepl("^47", utme),
      sub("^47", "74", utme),
      utme)) |>
  
  #same for session t116873, t116658 and t54139 (change 5 and 7)
   mutate(
    utme = if_else(
      landmark_id %in% c("Croton_Island-East_Edge", "Cow_Tree", "Hi_Step_X") & grepl("^57", utme),
      sub("^57", "75", utme),
      utme)) |>

  #remove the last digit of a 7 digit utme
  mutate(
  utme = if_else(
    landmark_id %in% c("Big_Sky_Den", "The_Observatory", "BB_X") & nchar(utme) == 7,
    substr(utme, 1, 6),   # keep only first 6 digits
    utme)) |>
  
  #for session t117205, add a 4 after the first 7, and session t111704 add a 5 after the 7
  mutate(
    utme = if_else(
      landmark_id == "Laura_Tree" & utme == "79731",
      sub("^7", "74", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Sunshine_Tree" & utme == "70360",
      sub("^7", "75", utme),
      utme)) |>
  
  # remove the 2nd number 
  mutate(
  utme = if_else(
    session %in% c("t112413", "t51456"),
    paste0(substr(utme, 1, 1), substr(utme, 3, nchar(utme))),  # remove 2nd character
    utme)) |>
  
  # remove all that is after the coma for both utme and utmn
  mutate(
    across(
      c(utme, utmn),
      ~ if_else(
        landmark_id %in% c("Serena_Lodge", "Box_Crossing", "Surprise_X"),
        sub(",.*", "", .x),   # keep only before first comma
        .x))) |>
  
  # For this session "t49599" 
  # 745095 8906 983
  mutate(
    utme = if_else(
      session == "t49599",
      substr(utme, 1, 6),   # keep only the first 6 characters
      utme)) |>
  mutate(
    utmn = if_else(
      session == "t49599",
      paste0("9806", utmn),   # add 9806 before in utmn
      utmn)) |>
  
  # keep just after the 2nd comma (t22224)
  mutate(
    utme = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utme),   # keep everything after 2nd comma
      utme),
    utmn = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utmn),
      utmn)) |>
  
  #Get rid of the last digit
   mutate(
    utme = if_else(
      session %in% c("t47831", "t49436", "t48055", "t111641", "t72442", "t57018", "t120547", "t115642"),   # replace with your session IDs
      substr(utme, 1, nchar(utme) - 1),  # drop last character
      utme)) |>
  
  # Now these have only 5 digits in utme, i will just remove them
  filter(!session %in% c("t115558", "t49400", "t66664", "t74990")) |> #I will see later if I decide to look for the last digit or not 
  
  #take out the sessions with weird utmn (99 and 0)
  filter(session != "t99428") |>
  filter(session != "t54706") |>
  #take out the Low_Ranker's_Crossing cause utmn= utme
  filter(landmark_id != "Low_Ranker's_Crossing") |>
  
  # Keep only before the comma 
  mutate(
    utmn = if_else(
      landmark_id %in% c("Uproar_Den", "Fig_Tree_Lodge", "Virginia_Maria_Crossing", "Three_Sisters", "Green_Den"),  #t60159
      sub(",.*", "", utmn),   # keep before the first comma
      utmn)) |>
  
  # Now keep all after the comma
  mutate(
    utmn = if_else(
      landmark_id %in% c("Swakey_Tree"),
      sub("^[^,]*,", "", utmn),   # remove everything up to the first comma
      utmn)) |>
  
  # remove all utmn of 6 and 8
  filter(nchar(utmn) != 6) |>
  filter(nchar(utmn) != 8) |>

  #change the utmn starting with 6 or 7 to 9
  mutate(
    utmn = if_else(
      grepl("^[67]", utmn),   # starts with 6 or 7
      sub("^[67]", "9", utmn),
      utmn))
  
```





To get utme and utmn values of + or - than 6 numbers:
```{r Check}

DB_check_utme <- DB_locations_for_MCPs_clean |>
  filter(nchar(utme) != 6) # now it is 0

DB_check_utmn <- DB_locations_for_MCPs_clean |>
  filter(nchar(utmn) != 7) # now is 0

```




### 1.5.2. Outliers

Now we visualize to see if there are still weird outliers points:
```{r Visualize}
# first convert in sf object :
DB_locations_for_MCPs_sf <- st_as_sf(DB_locations_for_MCPs_clean, coords = c("utme", "utmn"), crs = 32736)

# map to visualize
mapview(DB_locations_for_MCPs_sf, xcol = "utme", ycol = "utmn")
# this is to see the outliers

```

The points that looks like outliers:
- t120888
- s58
- t55969
- t112359
- s6974
- s115706
- s117340 


We remove those outliers points:
```{r Outliers}

#in sf objects
DB_AllGPS_sf_clean <- DB_locations_for_MCPs_sf |>
  select(-hyenas) |>
  filter(!session %in% c("t120888", "s58", "t55969", "t112359", "s6974", "s115706", "s117340")) 


# normal dataframe
DB_AllGPS <- DB_locations_for_MCPs_clean |>
  select(-hyenas) |>
  filter(!session %in% c("t120888", "s58", "t55969", "t112359", "s6974", "s115706", "s117340"))
  
  
```



# 2. Save and Shapefiles

SHP file contains geometrical information (for example shape type and coordinates). DBF file contains additional attribute information (for example name or temperature)

We use St_as_sf() function from sf package to create an sf object directly from a data frame containing a column of type sfc (simple feature list-column). Here, the utme and utmn are coordinates and so can be transformed into an sfc column.

You can also create and sf object with st_sf, where you will have a sfc object stored in an r object and then its attributes and you can combien them into an sf object : "pts_sf \<- st_sf(pts_data, geometry = pts)" where pts = sfc of coordinates pts_data = attributes

## 2.1. Save data frame

Active Den data:
```{r Save}
write_csv2(ActiveCD_data, "data_files/ActiveCD_data.csv") # Active CD per session with date
write.csv2(ActiveCD_unique, "data_files/ActiveCD_unique.csv") # the 355 unique CD used through the years
write_csv2(ActiveCD_year, "data_files/ActiveCD_year.csv")
```



All GPS location data, not just CD:
```{r Save}
write_csv2(DB_AllGPS, "data_files/DB_AllGPS.csv")
```



Hyena data for CD sessions:
```{r Save}
#For fiona:
write_csv2(ActiveCD_PerHyenas_data, "data_files/ActiveCD_PerHyenas_data.csv") #active CD with a hyena id column

write_csv2(ActiveCD_year, "data_files/ActiveCD_year.csv") #All CD used in one year (no month, no hyena id)
write_csv2(ActiveCD_unique, "data_files/ActiveCD_unique.csv") #the 355 unique den locations

```




## 2.2. Shapefiles

This is the locations of all the hyenas seen in the 4 territories:
```{r All GPS}

# outlier:
DB_AllGPS_sf <- DB_AllGPS_sf_clean |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den", "Mausoleun_at_Halicarnassus_Den")) 


# Saving it:
DB_AllGPS_sf <- sf::st_write(DB_AllGPS_sf, "data_files/DB_AllGPS_sf.shp", delete_dsn = TRUE)

```



Active CD hyena data:
```{r shp in UTM}

#Convert the dataframe into a sf object:
ActiveCD_sf <- st_as_sf(ActiveCD_data, coords = c("utme", "utmn"), crs = 32736) #EPSG code is 32736, UTM Zone 36 South

#outlier
ActiveCD_sf <- ActiveCD_sf |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

#Save the sf dataframe - can only do it once when it is not created before
ActiveCD_sf <- sf::st_write(ActiveCD_sf, "data_files/ActiveCD_sff.shp", delete_dsn = TRUE)

```
Now, ActiveCD_sf is a sf object because we have the feature geometrics (POINTS here) with their related attributes (year, landmark_id, clan, etc.).



```{r}
#Convert the dataframe into a sf object:
ActiveCD_year_sf <- st_as_sf(ActiveCD_year, coords = c("utme", "utmn"), crs = 32736) #EPSG code is 32736, UTM Zone 36 South

#outlier
ActiveCD_year_sf <- ActiveCD_year_sf |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

#Save the sf dataframe - can only do it once when it is not created before
ActiveCD_year_sf <- sf::st_write(ActiveCD_year_sf, "data_files/ActiveCD_year_sf.shp", delete_dsn = TRUE)

```


```{r}
#Convert the dataframe into a sf object:
ActiveCD_unique_sf <- st_as_sf(ActiveCD_unique, coords = c("utme", "utmn"), crs = 32736) #EPSG code is 32736, UTM Zone 36 South

#outlier
ActiveCD_unique_sf <- ActiveCD_unique_sf |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

#Save the sf dataframe - can only do it once when it is not created before
ActiveCD_unique_sf <- sf::st_write(ActiveCD_unique_sf, "data_files/ActiveCD_unique_sf.shp", delete_dsn = TRUE)

```





## 2.3. WGS84 transform

Also, to use leaflet (to make maps), we need to change the UTM coordinates into long lat : Here we transform the CD_sf into a WGS84 to use it with leaflet.


```{r All GSP}
# transform in WGS84
DB_AllGPS_sf_wgs84 <- st_transform(DB_AllGPS_sf, crs = 4326)

# outlier:
DB_AllGPS_sf_wgs84 <- DB_AllGPS_sf_wgs84 |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den", "Mausoleun_at_Halicarnassus_Den")) 

# saving it:
DB_AllGPS_sf_wgs84 <- sf::st_write(DB_AllGPS_sf_wgs84, "data_files/DB_AllGPS_sf_wgs84.shp", delete_dsn = TRUE)

```



ActiveCD data:
```{r}

ActiveCD_wgs84 <- st_transform(ActiveCD_sf, crs = 4326)

ActiveCD_wgs84 <- ActiveCD_wgs84 |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

ActiveCD_wgs84 <- sf::st_write(ActiveCD_wgs84, "data_files/ActiveCD_wgs84.shp", delete_dsn = TRUE)

```


```{r}
ActiveCD_year_wgs84 <- st_transform(ActiveCD_year_sf, crs = 4326)

ActiveCD_year_wgs84 <- ActiveCD_year_wgs84 |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

ActiveCD_year_wgs84 <- sf::st_write(ActiveCD_year_wgs84, "data_files/ActiveCD_year_wgs84.shp", delete_dsn = TRUE)

```


```{r}
ActiveCD_unique_wgs84 <- st_transform(ActiveCD_unique_sf, crs = 4326)

ActiveCD_unique_wgs84 <- ActiveCD_unique_wgs84 |>
  filter(!landmark_id %in% c("Migration_Den", "17km_Den", "Heathrow_Den", "Runway_Den", "Sleepy_Den",  "Mausoleun_at_Halicarnassus_Den")) 

ActiveCD_unique_wgs84 <- sf::st_write(ActiveCD_unique_wgs84, "data_files/ActiveCD_unique_wgs84.shp", delete_dsn = TRUE)

```




# 3. Data visualisation

## 3.1. Maps

Now let's try to create a map...

### i) Basemap (mapview)

Mapview is just to see real quick what the points look like.

```{r Mapviews}

#According to year
clan_yearly_CD_mapview <- mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "year")
htmlwidgets::saveWidget(clan_yearly_CD_mapview@map, "Maps/clan_yearly_CD_mapview.html", selfcontained = TRUE)


# According to clan 
clan_CD_mapview <- mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "clan")
htmlwidgets::saveWidget(clan_CD_mapview@map, file = "Maps/clan_CD_mapview.html", selfcontained = TRUE)


```

### ii) Basemap year (leaflet)

Leaflet is a bit more complex than mapview and can get you more options for mapping.

We will create a basemap for starter: - Leaflet tile : draws the map tiles. There are numerous types of map tiles. - Layers Control: Leaflet has a nice little control that allows your users to control which layers they see on your map

```{r Leaflet per year}

pal <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$clan)

CD_peryear_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years <- unique(CD_sf_wgs84$year)

for (yr in years) {
  data_year <- CD_sf_wgs84 |> filter(year == yr)
  
CD_peryear_mapleaflet <- CD_peryear_mapleaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal(clan),
                     group = yr,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan))
}

#Add toggle control for years
CD_peryear_mapleaflet <- CD_peryear_mapleaflet |>
  
  addLayersControl(
    baseGroups = years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal, values = CD_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

CD_peryear_mapleaflet  # display the map

```

Saving as a html file

```{r}

htmlwidgets::saveWidget(CD_peryear_mapleaflet, file = "Maps/CD_peryear_mapleaflet.html", selfcontained = TRUE)

# Saving it as a R object to reuse in Covariables:
saveRDS(CD_peryear_mapleaflet, "Maps/CD_peryear_mapleaflet.rds")

```

### iii) Basemap clan (leaflet)

This is the same map, but we invert the clan and year variables

```{r Leaflet per clan}

pal_year <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$year)

CD_perclan_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
clans <- unique(CD_sf_wgs84$clan)

for (cl in clans) {
  data_clan <- CD_sf_wgs84 |> filter(clan == cl)
  
CD_perclan_mapleaflet <- CD_perclan_mapleaflet |>
    addCircleMarkers(data = data_clan,
                     color = ~pal_year(year),
                     group = cl,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year,"<br>",
                                    "Clan:", clan, "<br>", 
                                    "Name:", landmark_id))
}

#Add toggle control for years
CD_perclan_mapleaflet <- CD_perclan_mapleaflet |>
  
  addLayersControl(
    baseGroups = clans,
    overlayGroups = c("Clan"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal_year, values = CD_sf_wgs84$year,
            title = "Year", opacity = 1)
  

CD_perclan_mapleaflet  # display the map

```

Save

```{r}

htmlwidgets::saveWidget(CD_perclan_mapleaflet, file = "Maps/CD_perclan_mapleaflet.html", selfcontained = TRUE)

```

### iv) All GPS points map

```{r}

pal2 <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

AllGPS_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years2 <- unique(DB_AllGPS_sf_wgs84$year)

for (yr in years2) {
  data_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
AllGPS_mapleaflet <- AllGPS_mapleaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal2(clan),
                     group = yr,
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan, "<br>", "session", "<br>", "lndmrk_"))
}

#Add toggle control for years
AllGPS_mapleaflet <- AllGPS_mapleaflet |>
  
  addLayersControl(
    baseGroups = years2,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal2, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

AllGPS_mapleaflet  # display the map

```


```{r Save}
#save
htmlwidgets::saveWidget(AllGPS_mapleaflet, file = "Maps/AllGPS_mapleaflet.html", selfcontained = TRUE)

```


### v) Per Location

```{r Location map}
pal3 <- colorFactor(c("red", "blue", "yellow", "green", "pink", "black", "orange"), domain = DB_AllGPS_sf_wgs84$location)

AllGPS_perlocation_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years2 <- unique(DB_AllGPS_sf_wgs84$year)

for (yr in years2) {
  data_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
AllGPS_perlocation_mapleaflet <- AllGPS_perlocation_mapleaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal3(location),
                     group = yr,
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", 
                                    "Clan:", clan, "<br>", 
                                    "Session", session, "<br>", 
                                    "Landmark", landmark_id))
}

#Add toggle control for years
AllGPS_perlocation_mapleaflet <- AllGPS_perlocation_mapleaflet |>
  
  addLayersControl(
    baseGroups = years2,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal3, values = DB_AllGPS_sf_wgs84$location,
            title = "Location", opacity = 1)
  

AllGPS_perlocation_mapleaflet  # display the map
```


```{r Save}
#save
htmlwidgets::saveWidget(AllGPS_perlocation_mapleaflet, file = "Maps/AllGPS_perlocation_mapleaflet.html", selfcontained = TRUE)

```


```{r Location}

pal_clan <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

AllGPS_location_perclan_mapleaflet <- leaflet() |>
  addTiles()

loc <- unique(DB_AllGPS_sf_wgs84$location)

for (lc in loc) {
  data_loc <- DB_AllGPS_sf_wgs84 |> filter(location == lc)
  
AllGPS_location_perclan_mapleaflet <- AllGPS_location_perclan_mapleaflet |>
    addCircleMarkers(data = data_loc,
                     color = ~pal_clan(clan),
                     group = lc,
                     radius = 1,
                     fillOpacity = 1,
                     popup = ~paste("Year:", year,"<br>",
                                    "Clan:", clan, "<br>", 
                                    "Name:", landmark_id))
}

#Add toggle control for years
AllGPS_location_perclan_mapleaflet <- AllGPS_location_perclan_mapleaflet |>
  
  addLayersControl(
    baseGroups = loc,
    overlayGroups = c("Location"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal_clan, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

AllGPS_location_perclan_mapleaflet  # display the map
```


```{r Save}
#save
htmlwidgets::saveWidget(AllGPS_location_perclan_mapleaflet, file = "Maps/AllGPS_location_perclan_mapleaflet.html", selfcontained = TRUE)

```


### vi) Monthly CD

```{r}

pal <- colorFactor(palette = "viridis", domain = CD_monthly_wgs84$clan)

CD_monthlyPerYear_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years <- unique(CD_monthly_wgs84$year)

for (yr in years) {
  data_year <- CD_monthly_wgs84 |> filter(year == yr)
  
CD_monthlyPerYear_mapleaflet <- CD_monthlyPerYear_mapleaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal(clan),
                     group = yr,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan))
}

#Add toggle control for years
CD_monthlyPerYear_mapleaflet <- CD_monthlyPerYear_mapleaflet |>
  
  addLayersControl(
    baseGroups = years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal, values = CD_monthly_wgs84$clan,
            title = "Clan", opacity = 1)
  

CD_monthlyPerYear_mapleaflet  # display the map


```


```{r}
#save
htmlwidgets::saveWidget(CD_monthlyPerYear_mapleaflet, file = "Maps/CD_monthlyPerYear_mapleaflet.html", selfcontained = TRUE)
```



### vii) Active CD
with active CD dataframe:
```{r}

pal <- colorFactor(palette = "viridis", domain = ActiveCD_wgs84$clan)

ActiveCD_monthlyPerYear_mapleaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years <- unique(ActiveCD_wgs84$year)

for (yr in years) {
  data_year <- ActiveCD_wgs84 |> filter(year == yr)
  
ActiveCD_monthlyPerYear_mapleaflet <- ActiveCD_monthlyPerYear_mapleaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal(clan),
                     group = yr,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan))
}

#Add toggle control for years
ActiveCD_monthlyPerYear_mapleaflet <- ActiveCD_monthlyPerYear_mapleaflet |>
  
  addLayersControl(
    baseGroups = years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal, values = ActiveCD_wgs84$clan,
            title = "Clan", opacity = 1)
  

ActiveCD_monthlyPerYear_mapleaflet  # display the map


```


```{r}
#save
htmlwidgets::saveWidget(ActiveCD_monthlyPerYear_mapleaflet, file = "Maps/ActiveCD_monthlyPerYear_mapleaflet.html", selfcontained = TRUE)
```








## 3.2. Tables 

Calculate the sum of CD per year and clan.

### iii) ActiveCD_data
```{r}
ActiveCD_data$year <- as.character(ActiveCD_data$year)

ActiveCD_data_tab <- ActiveCD_data |>
  group_by(year, month, clan) |>
  summarise(CD_num = n()) |>
  arrange(year, month, clan)

ActiveCD_data_tab
```

```{r}

ActiveCD_data_year <- readr::read_csv2("data_files/ActiveCD_data_year.csv")

ActiveCD_unique_tab <- ActiveCD_data_year |>
  distinct(landmark_id, utme, utmn, clan)

write_csv2(ActiveCD_unique_tab, "Tables/ActiveCD_unique_tab.csv")


```



# 4. Prepare RSF data

Before we can run an RSF, we need to prepare the database that will be used. The first step for that is to get our used GPS points, in this case, it is the den locations that were extracted and put in `CD_for_map` or `CD_sf` (CD = Communal Den). The second step is to create the clans' MCP (Minimum Convex Polygon). We will generate the 95% MCP and 100% MCP.

With the used points and each clans MCP, we will be able to generate random points in their respective territories. I will attribute 10 random points (available points) to the used points. To do so, I will take the clans' MCPs as the home range of the clans, and generate random points in the polygons created. This will create a database of used and available points. The used points are the den locations known, and the available are all the adjacent points in the territory.

I need to determine how many points I want to sample for each points used. For example, 5 points per used points.

The `random_points()` function will create this `case\_` variable that will be use in the RSF later.



## 4.1. MCP by clans

To make our RSF model, we need to create random points of potential locations of where the hyenas could have denned. To do that, we first need to know the extent of the clans' territory, and so we will calculate the MCP (Minimum Convex Polygon) for each clan. An MCP calculate the area that an animal uses. The MCP essentially draws a polygon around all the points where an animal has been observed.

Also, we are using the 95% MCP, so that it removes the 5% of outlier points.

### 4.1.1. Individual MCPs

Here is the MCPs for all hyenas locations
```{r sp}
# Convert all points to sp (adehabitatHR works with sp)
AllGPS_sp <- as(DB_AllGPS_sf_wgs84, "Spatial")
```


```{r Clan 95 MCPs}
# Create MCPs grouped by clan
mcp95_clan <- mcp(AllGPS_sp[, "clan"], percent = 95)

# Convert back to sf
mcp95_clan_sf <- st_as_sf(mcp95_clan)

# Inspect
plot(st_geometry(mcp95_clan_sf), col = NA, border = "red")

# change id for clan 
mcp95_clan_sf <- mcp95_clan_sf |>
  rename(clan = id)

```

```{r Clans 100 MCP}

# Create MCPs grouped by clan
mcp100_clan <- mcp(AllGPS_sp[, "clan"], percent = 100)

# Convert back to sf
mcp100_clan_sf <- st_as_sf(mcp100_clan)

# Inspect
plot(st_geometry(mcp100_clan_sf), col = NA, border = "red")

# change id for clan 
mcp100_clan_sf <- mcp100_clan_sf |>
  rename(clan = id)

```
I don't think we will be using the 100% MCP... Gives way too large MCPs.



Save
```{r Save}

# Save the dataframe
mcp95_clan_sf <- sf::st_write(mcp95_clan_sf, "data_files/mcp95_clan_sf.shp", delete_dsn = TRUE)

```



### 4.1.2. Yearly MCPs

Here is to make yearly MCPs per clans.
```{r Yearly 95% MCPs}

# Create an ID that combines clan and year
AllGPS_sp$clan_year <- paste(AllGPS_sp$clan, AllGPS_sp$year, sep = "_")

# Count points per clan-year
counts <- as.data.frame(table(AllGPS_sp$clan_year))

# Keep only groups with >= 5 relocations
valid_ids <- counts$Var1[counts$Freq >= 5]

# Filter points
AllGPS_sp_valid <- AllGPS_sp[AllGPS_sp$clan_year %in% valid_ids, ]

# Run MCP only on valid groups
mcp95_clan_year <- mcp(AllGPS_sp_valid[, "clan_year"], percent = 95)
mcp95_clan_year_sf <- st_as_sf(mcp95_clan_year)



# Split IDs back into clan and year
mcp95_clan_year_sf <- mcp95_clan_year_sf |>
  mutate(clan = sub("_(\\d+)$", "", row.names(mcp95_clan_year_sf)),
         year = sub(".*_", "", row.names(mcp95_clan_year_sf)))


```

```{r Excluded vs included}

# Count relocations per clan-year
counts <- as.data.frame(table(AllGPS_sp$clan_year)) |>
  rename(clan_year = Var1, n_points = Freq)

# Split clan and year into separate columns
counts <- counts |>
  mutate(
    clan = sub("_(\\d+)$", "", clan_year),
    year = sub(".*_", "", clan_year))

# Show which groups are valid / invalid
counts <- counts |>
  mutate(valid = ifelse(n_points >= 5, "yes", "no"))

# View excluded groups
excluded <- counts |>
  filter(valid == "no")

# View valid groups
included <- counts |>
  filter(valid == "yes")

excluded
included

```


Save
```{r Save}

#  Save the dataframe
mcp95_clan_year_sf <- sf::st_write(mcp95_clan_year_sf, "data_files/mcp95_clan_year_sf.shp", delete_dsn = T)

```



### 4.1.3. Plots


```{r Single MCP map}

AllGPS_MCPs95_mapleaflet <- AllGPS_mapleaflet |>
  addPolygons(data = mcp95_clan_sf,
              color = "red",
              fillColor = "none",
              weight = 2,
              fillOpacity = 0.05)
  
AllGPS_MCPs95_mapleaflet
```


```{r Save}
htmlwidgets::saveWidget(AllGPS_MCPs95_mapleaflet, file = "Maps/AllGPS_MCPs95_mapleaflet.html", selfcontained = TRUE)

```


```{r Yearly MCPs}

# Start with an empty leaflet map
AllGPS_yearly_95MCPs_mapleaflet <- leaflet() |> addTiles()

# Define your clan palette
pal2 <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

# Get all years
years2 <- sort(unique(DB_AllGPS_sf_wgs84$year))

# Loop over years to add both points and MCPs under the same baseGroup (year)
for (yr in years2) {
  
  # Points for this year
  points_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
  # MCP polygons for this year
  mcp_year <- mcp95_clan_year_sf |> filter(year == yr)

  
  # Add points
  AllGPS_yearly_95MCPs_mapleaflet <- AllGPS_yearly_95MCPs_mapleaflet |>
    addCircleMarkers(data = points_year,
                     color = ~pal2(clan),
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>",
                                    "Clan:", clan, "<br>", 
                                    "Session:", session, "<br>",
                                    "Landmark:", landmark_id),
                     group = paste0("Year ", yr))  # group = year
  
  # Add MCP polygons
  AllGPS_yearly_95MCPs_mapleaflet <- AllGPS_yearly_95MCPs_mapleaflet |>
    addPolygons(data = mcp_year,
                color = "red",
                fillColor = "none",
                weight = 2,
                fillOpacity = 0.05,
                popup = ~paste("Year:", year, "<br>",
                               "Clan:", clan),
                group = paste0("Year ", yr))  # same group = year
  
}

# Add **radio button style toggle** for years
AllGPS_yearly_95MCPs_mapleaflet <- AllGPS_yearly_95MCPs_mapleaflet |>
  addLayersControl(
    baseGroups = paste0("Year ", years2),  # radio button style
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  addLegend("bottomright", pal = pal2, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)


AllGPS_yearly_95MCPs_mapleaflet


```


Save
```{r}

# Save the map
htmlwidgets::saveWidget(AllGPS_yearly_95MCPs_mapleaflet, file = "Maps/AllGPS_yearly_95MCPs_mapleaflet.html", selfcontained = TRUE)
```


## 4.2. Random points

Great, now that we have our 95% MCPs based on all GPS points of the hyenas and the GPS locations of the communal dens, we can proceed to generate random points and put all that in a dataframe that will be use in our RSF (See covariables quarto to see all the covariables added to that dataframe).

```{r presence track}
#Convert your sf object to a data frame with coordinates: 
ActiveCD_wgs84_df <- ActiveCD_wgs84 |> 
  mutate(long = sf::st_coordinates(geometry)[,1], lat = sf::st_coordinates(geometry)[,2]) 

#Change the CD_sf_wgs84_df into a track_xy object 
presence_track <- make_track(ActiveCD_wgs84_df, long, lat, crs = 4326, clan = clan, year = year, landmark_id = landmark_id, month = month, session = session)
```


### 4.2.1. General MCP
This one is with the general MCP

```{r 10 random per used}

# Function to generate 10 random points per used
generate_randoms <- function(pres_row, mcp_polygons, n = 10) {
  clan_i        <- pres_row$clan
  year_i        <- pres_row$year
  month_i       <- pres_row$month
  session_i     <- pres_row$session
  landmark_id_i <- pres_row$landmark_id
  
  mcp_i <- mcp_polygons %>% filter(clan == clan_i)
  
  if (nrow(mcp_i) == 0) return(NULL)
  
  # Use sf::st_sample directly to get n points inside polygon
  rand_geom <- st_sample(mcp_i, size = n, exact = TRUE)
  
  rand_sf <- st_sf(
    geometry = rand_geom,
    clan = clan_i,
    year = year_i,
    month = month_i,
    session = session_i,
    landmark_id = landmark_id_i,
    case_ = 0,
    crs = st_crs(mcp_polygons)
  )
  
  return(rand_sf)
}

mcp95_clan_sf <- mcp95_clan_sf |>
  st_make_valid()

RandomPointsCD_data <- map_dfr(
  1:nrow(presence_track),
  ~ generate_randoms(presence_track[.x, ], mcp95_clan_sf, n = 10))

# Combine used and random points
RandomPointsCD_data <- bind_rows(
  presence_track |>
    select(-x_, -y_) |>
    mutate(case_ = 1),
  RandomPointsCD_data)

# Clean up naming
RandomPointsCD_data <- RandomPointsCD_data |>
  rename(case = case_)

```

Save
```{r Save}
RandomPointsCD_data <- sf::st_write(RandomPointsCD_data, "data_files/RandomPointsCD_data.shp", delete_dsn = T)
```


### 4.2.2. Yearly MCPs
Now this one is with the yearly MCP
```{r}
# Function to generate random points per presence row
generate_randoms <- function(pres_row, mcp_polygons, n = 10) {
  # Extract identifiers from the presence row
  clan_i       <- pres_row$clan
  year_i       <- pres_row$year
  month_i      <- pres_row$month
  session_i    <- pres_row$session
  landmark_id_i <- pres_row$landmark_id
  
  # Subset to the MCP polygon for the *same clan and year*
  mcp_i <- mcp_polygons %>%
    filter(clan == clan_i, year == year_i)
  
  # If no polygon found (e.g. missing MCP for that year/clan), skip
  if (nrow(mcp_i) == 0) return(NULL)
  
  # Sample n random points inside this year's MCP
  rand_geom <- st_sample(mcp_i, size = n, exact = TRUE)
  
  # Wrap sampled points into an sf object with metadata
  rand_sf <- st_sf(
    clan        = clan_i,
    year        = year_i,
    month       = month_i,
    session     = session_i,
    landmark_id = landmark_id_i,
    case_       = 0,                 # available points
    geometry    = rand_geom,
    crs         = st_crs(mcp_polygons)
  )
  
  return(rand_sf)
}


mcp95_clan_year_sf <- mcp95_clan_year_sf |>
  st_make_valid()

# Apply over all presence points
RandomPointsCD_yearly_data <- map_dfr(
  1:nrow(presence_track),
  ~ generate_randoms(presence_track[.x, ], mcp95_clan_year_sf, n = 10))

# Combine used and random points
RandomPointsCD_yearly_data <- bind_rows(
  presence_track |>
    select(-x_, -y_) |>
    mutate(case_ = 1),
  RandomPointsCD_yearly_data)

# Clean up naming
RandomPointsCD_yearly_data <- RandomPointsCD_yearly_data |>
  rename(case = case_)



```

```{r Save}

RandomPointsCD_yearly_data <- sf::st_write(RandomPointsCD_yearly_data, "data_files/RandomPointsCD_year_data.shp", delete_dsn = T)

```




### 4.2.3. Plots

Using the yearly data points:
```{r}

RandomPointsCD_mapleaflet <- leaflet() |> 
  addTiles() |>

  # Then plot random (case_ == 0) in blue (will appear on top)
  addCircleMarkers(
    data = RandomPointsCD_yearly_data[RandomPointsCD_yearly_data$case == 0, ],
    color = "blue", radius = 1, fillOpacity = 0.8,
    group = "Random"
  ) |>
   # First plot used (case_ == 1) in red
  addCircleMarkers(
    data = RandomPointsCD_yearly_data[RandomPointsCD_yearly_data$case == 1, ],
    color = "red", radius = 1, fillOpacity = 0.8,
    group = "Used"
  ) |>
  addLayersControl(
    overlayGroups = c("Used", "Random"),
    options = layersControlOptions(collapsed = FALSE)
  )

RandomPointsCD_mapleaflet

```


```{r}
# Save the map
htmlwidgets::saveWidget(RandomPointsCD_mapleaflet, file = "Maps/RandomPointsCD_mapleaflet.html", selfcontained = TRUE)

```


Adding to the AllGPS
```{r}

# Adding the random point in the AllGPS_yearly_95MCPs_mapleaflet
# New map object
CD_yearly_95MCPs_withRandom_mapleaflet <- leaflet() |> addTiles()

pal2 <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)
years2 <- sort(unique(DB_AllGPS_sf_wgs84$year))


for (yr in years2) {
  
  points_year  <- ActiveCD_wgs84 |> filter(year == yr)
  mcp_year     <- mcp95_clan_year_sf |> filter(year == yr)
  random_year  <- RandomPointsCD_yearly_data |> filter(year == yr)
  
  # 1. Random points (back layer)
  CD_yearly_95MCPs_withRandom_mapleaflet <- CD_yearly_95MCPs_withRandom_mapleaflet |>
    addCircleMarkers(data = random_year,
                     color = "red",
                     radius = 1,
                     fillOpacity = 0.6,
                     popup = ~paste("Year:", year, "<br>",
                                    "Random point"),
                     group = paste0("Year ", yr))
  
  # 2. Used clan points (middle layer)
  CD_yearly_95MCPs_withRandom_mapleaflet <- CD_yearly_95MCPs_withRandom_mapleaflet |>
    addCircleMarkers(data = points_year,
                     color = ~pal2(clan),
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>",
                                    "Clan:", clan, "<br>", 
                                    "Landmark:", landmark_id, "<br>",
                                    "Month:", month),
                     group = paste0("Year ", yr))
  
  # 3. MCP polygons (top layer)
  CD_yearly_95MCPs_withRandom_mapleaflet <- CD_yearly_95MCPs_withRandom_mapleaflet |>
    addPolygons(data = mcp_year,
                color = "red",
                fillColor = "none",
                weight = 2,
                fillOpacity = 0.05,
                popup = ~paste("Year:", year, "<br>",
                               "Clan:", clan),
                group = paste0("Year ", yr))
}

# Add controls + legend
CD_yearly_95MCPs_withRandom_mapleaflet <- CD_yearly_95MCPs_withRandom_mapleaflet |>
  addProviderTiles(providers$Esri.WorldImagery) |>

  addLayersControl(
    baseGroups = paste0("Year ", years2),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  addLegend("bottomright", pal = pal2, values = ActiveCD_wgs84$clan,
            title = "Clan", opacity = 1)

CD_yearly_95MCPs_withRandom_mapleaflet

```


```{r}
# Save the map
htmlwidgets::saveWidget(CD_yearly_95MCPs_withRandom_mapleaflet, file = "Maps/CD_yearly_95MCPs_withRandom_mapleaflet.html", selfcontained = TRUE)

```






