---
output: html_document
title: "MHP Special Project - Data prep"
author: "Anne Ju Laberge"
date: "2025"
format:
  html:
    
    code-fold: true
    output-fold: true
    code-tools: true
    code-link: true
    code-summary: "Show the code"
    theme: cosmo
    embed-resources: true
    self-contained: true
    toc-depth: 5
    toc: true
    toc-float: true
editor: visual
editor_options:
   
   chunk_output_type: console
execute:
  warning: false
  error: true
---

#0. Data

Installing the package to have access to the MHP data (Just for the first time).

```{r For MHP data}

#install.packages("usethis")

#library(usethis)
#usethis::create_github_token() # to create a Token

#gitcreds::gitcreds_set()

#devtools::install_github("MaraHyenaProject/hyenadata", auth_token = gh::gh_token())

```

Load packages:

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

```

```{r Load MHP tables}
#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```

#1. Create db

I will be using the tblLandmarks to identified CDs' locations (GPS points). Maybe also tblSessions/tblLocationsPerSession to ID sessions at communal dens and then have the location with the time/year.

```{r Location data}
str(tblLandmarks)

CD_loc_data <- tblLandmarks |>
  subset(category == "cd" | category == "d") |>
  subset(region == "talek" | region == "serena") |>
  filter(landmark_id != "Car Park Den") |> # This coordinate is the same as utme, this is wrong... also wrong in the raw data... will have to look on the GPS itself
  select(landmark_id, utme, utmn, date) |>
  drop_na(utme, utmn)

#write_csv2(CD_loc_data, "data_files/CD_loc_data.csv")
CD_loc_data_date <- CD_loc_data |>
  separate(date, c("date1", "date2"), sep = " ", remove=FALSE) |>
  separate(date2, c("year", "month", "day"), sep = "-", remove = F) |>
  separate(date1, c("month1", "day1", "year1"), sep = "/", remove = F)

```

I gave up... The date column was too messed up so I exported it into a csv file, added the year column by hand and now I just reimport it here:

```{r Location data}
CD_loc_data <- read.csv2("data_files/CD_loc_data.csv")

CD_loc_data <- CD_loc_data |>
  select(-date)

```

```{r Session+Location data}

Sess_data <- tblSessions |>
  select(clan, session) |>
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra"))

CD_SessLoc_data <- tblLocationsPerSession |>
  subset(str_detect(landmark_id, "den") | str_detect(landmark_id, "Den"))

CD_data <- merge(Sess_data, CD_SessLoc_data, by = "session")

```

Great, now I have all my Den sessions in Talek. I need to fill the utme and utmn columns so all of them of a specific locations. I can try to do that with my CD_loc_data database created earlier here.

I want to match the landmark_id columns in both database and when it does, to add the utme and utmn values in the corresponding columns.

```{r Full DB}

CD_sess_data <- CD_data |>
  select(-c(3:5, 7:11))

# merge the utme and utmn columns
CD_data_full <- merge(CD_sess_data, CD_loc_data, by = "landmark_id")

str(CD_data_full)

```

```{r DB for mapping}

# First we need to turn the coordinates into numeric values
# here all the weird coordinates (the ones with multiple numbers that doesn't make sense) are transformed to NAs
CD_data_full$utme <- as.numeric(CD_data_full$utme)
CD_data_full$utmn <- as.numeric(CD_data_full$utmn)


#Just keeping the coordinates for the map
CD_for_map <- CD_data_full |>
  select(utme, utmn, landmark_id, clan, year) |>
  #create year column
  #mutate(year = year(date)) |> no need anymore
  
  # Getting rid of any year prior to the study time frame
  filter(year >= "1988") |> 
  
  #changing talek.w into talek
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  
  #Add a month column
  #mutate(month = month(date)) |> not gonna do that anymore
  
  #Put "_" in the landmark_id columns
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  #Keep only 1 distinct coordinates per den name
  distinct(landmark_id, year, .keep_all = T) |>
  
  #Getting rid of utmn coordinate with more or less than 7 numbers
  mutate(utmn = ifelse(nchar(as.character(utmn)) == 7, utmn, NA)) |>
  #Getting rid of utme coordinate with more or less than 7 numbers
  mutate(utme = ifelse(nchar(as.character(utme)) == 6, utme, NA)) |>
# for these two, we will change that eventuallly but for now we get rid of them
  
  #Getting rid of NAs
  drop_na(utme, utmn) |>
  filter(utme != 0)

CD_for_map$year <- as.character(CD_for_map$year)

str(CD_for_map)

```



##1.1. All GPS locations

We will be using the GPS locations of all the points that hyenas were seen. We will use that to create KDEs and then create the random points to run the RSF.

So I have the table hyenas per session, which has a column for hyena IDs and session IDs. I will keep just these two rows, and filter to keep only serena's sessions and talek (not fig tree, pond, etc.).
I will also add the individual's clan with tlbHyenas.

```{r Sessions with hyenas}
DB_hyenas <- tblHyenasPerSession |>
  select(session, id, date) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session)) |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F)
  



# before adding the clan to the hyenas ids, I want to make sure each id have only 1 clan associated (see session S115986)

# before that, we will identify the ids that are duplicated because its hyenas born in one clan who immigrated in another one :

dupe_rows_indID <- tblClanMembership |>
  select(id, clan, status, notes) |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "unknown") |>
  filter(id %in% id[duplicated(id)]) |>
  filter(clan != "kcm") |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  ungroup() |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |>
  filter(id %in% id[duplicated(id)]) |>
  select(-notes) |>
  filter(id %in% id[duplicated(id)]) |>
  group_by(id) |>
  filter(n() > 1 & n_distinct(clan) > 1) |>
  slice(2) |>
  ungroup()


length(unique(dupe_rows_indID$id)) #20

```



Slicing the hyenas with the correct clan :

```{r Sessions with hyenas}

DB_ind_clan <- tblClanMembership |> # there are 3097 hyenas
  select(id, clan, status) |>
  filter(clan == "talek" | clan == "serena.s" | clan == "serena.n" | clan == "happy.zebra" | clan == "talek.w") |> # now we have 2061 hyenas
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |> 
  distinct() |>
  # to keep the second row of the duplicated ids
  group_by(id) |>
  filter(n() == 1 | row_number() == 2) |>
  ungroup()


length(unique(DB_ind_clan$id)) #1966 ind

dupes_diff_clan <- DB_ind_clan %>%
  group_by(id) %>%
  filter(n_distinct(clan) > 1) %>%
  ungroup()




```


```{r Sessions with hyenas}

# Then I want to add the clan for each individual 
DB_hyenas_clan <- DB_hyenas |>
  left_join(DB_ind_clan %>% select(id, clan), by = "id", relationship = "many-to-many") |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  #changing talek.w into talek
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  group_by(session, clan, year) |>
  summarise(
    hyenas = paste(unique(id), collapse = ","),
    .groups = "drop")

# then, let's get rid of the NAs in hyenas_id:
DB_hyenas_clan <- DB_hyenas_clan |>
  drop_na(hyenas) # No NA





# So, all this time I had a dataset with the sessions ID, clan, hyenas ID and date:

DB_session_hyenasID <- tblSessions |>
  select(clan, date, session, location, hyenas) |>
  filter(clan == "talek" | clan == "talek.w" | clan == "serena.n" | clan == "serena.s" | clan == "happy.zebra") |>
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-month, -day, -date) |>
  mutate(clan = str_replace_all(clan, "talek.w", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(!grepl("-mr", session)) |>
  filter(!grepl("-ft", session)) |>
  filter(!grepl("amb", session))

# then, let's get rid of the NAs in hyenas_id:
DB_session_hyenasID <- DB_session_hyenasID |>
  drop_na(hyenas) # from 151719 to 122558 (difference of 29161)


```

I have more lines in DB_session_hyenasID (122558) than in DB_hyenas_clan (128 324). The difference is 5766. Let's see what are the differences:
```{r}

extra_rows <- DB_session_hyenasID |>
  distinct(session, year, hyenas, clan) |>
  anti_join(DB_hyenas_clan |>
              distinct(session, year, hyenas, clan),
            by = c("session", "year", "hyenas", "clan"))
#This is 6038


extra_sessions <- DB_session_hyenasID |>
  distinct(session) |>
  anti_join(DB_hyenas_clan %>% distinct(session), by = "session")
#this is 1340


extra_rows <- DB_session_hyenasID |>
  distinct(session, year, clan) |>
  anti_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#1791

common_rows <- DB_session_hyenasID %>%
  distinct(session, year, clan) %>%
  inner_join(
    DB_hyenas_clan %>% distinct(session, year, clan),
    by = c("session", "year", "clan"))
#120 767

```









Now that we have the sessions with all the hyenas that were seen in each of our 3 territories, we want to add the GPS location of those sessions.
We will use tblLocationsPerSession.

```{r Adding GPS location}

DB_all_locations <- DB_session_hyenasID |>
  left_join(tblLocationsPerSession %>% select(session, landmark_id, utme, utmn), by = "session", relationship = "many-to-many") |>
  #take the first row out
  group_by(session) |>
  slice(1) |>
  ungroup()

dupe_rows <- DB_session_hyenasID %>%
  filter(session %in% session[duplicated(session)])# no duplicates


```


Now we want to fill the NAs with the landmark ID GPS point. 

```{r Filling the NAs}

# first create a dataframe with all the coordinates for the landmarks:

DF_lndmk <- tblLandmarks |>
  select(landmark_id, utme, utmn) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "Marsh_Den" & (is.na(utme) | is.na(utmn))))


# should be 75271 rows:
DB_locations_for_MCPs <- DB_all_locations |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "~NA" & is.na(utme) & is.na(utmn))) |>
  filter(!(is.na(landmark_id) & is.na(utme) & is.na(utmn))) |>
  
  #changing the names that doesn't match:
  mutate(landmark_id = case_when(
    landmark_id == "Airstrip_D" ~ "Airstrip_Den",
    landmark_id == "Maushugana_D_11" ~ "Mashugana_Den_II",
    landmark_id == "Arba_Den" ~ "ARBA_Natal_Den",  
    landmark_id == "Banana_T" ~ "Banana_Tree",
    landmark_id == "upro_d" ~ "Uproar_Den",   
    landmark_id == "Uproar_D" ~ "Uproar_Den", 
    landmark_id == "cop_c" ~ "Copulation_Culvert", #verify
    landmark_id == "Coral_Rf" ~ "Coral_Reef",
    landmark_id == "Border_Lugga" ~ "South_Border_Lugga",
    landmark_id == "btn_Mystery_Spot_&_Talek_Gate_on_basecamp_plain" ~ "Mystery_Spot",
    landmark_id == "Camel_X" ~ "Camel_Crossing",
    landmark_id == "Century_X" ~ "Century_Crossing",
    landmark_id == "Crtn_Edge_Den" ~ "Croton_Edge_Communal_Den",
    landmark_id == "Dave_D" ~ "Dave_Den",
    landmark_id == "Fever_T_ND/Cd_Cplx" ~ "Fever_Tree_Natal_Den/Communal_Den_Complex",
    landmark_id == "FFL_Wrap" ~ "Falafel_Wrap",
    landmark_id == "Hat_T" ~ "Hat_Tree",
    landmark_id == "K-6_Den" ~ "K6_Den",
    landmark_id == "Leopard_Sex_X" ~ "Leopard_Sex_Crossing",
    landmark_id == "MGTA_ND_13" ~ "MGTA_ND13",
    landmark_id == "Paula`s_Pond" ~ "Paula's_Pond",
    landmark_id == "Plant_D" ~ "Plantation_Den",
    landmark_id == "Pole_T" ~ "Pole_Tree",
    landmark_id == "Precipice_Den_3" ~ "Precipice_Den_III",
    landmark_id == "S_Lodge" ~ "Serena_Lodge",
    landmark_id == "Sbre_Gtwy_X" ~ "Sabre_Gateway_X",
    landmark_id == "Sherbert_D" ~ "Sherbert_Den",
    landmark_id == "Top_T_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "TT_L" ~ "Talek_Lugga_-_Top",
    landmark_id == "Vulture_X" ~ "Vulture_Crossing",
    landmark_id == "Central_Park_CD2" ~ "Central_Park_Communal_Den",
    landmark_id == "Rbend_D" ~ "Riverbend_Den",
    landmark_id == "Leopard_X" ~ "Leopard_Crossing",
    landmark_id == "Talek_L_X" ~ "Talek_Lugga_Crossing",
    landmark_id == "T_Marsh" ~ "Talek_Marsh",
    landmark_id == "Border_T" ~ "Border_Tree",
    landmark_id == "Stupid_T" ~ "Stupid_Tree",
    landmark_id == "Dave_L_X" ~ "Dave's_Lugga_Crossing",
    landmark_id == "Top_FT_L" ~ "Fig_Tree_Lugga_-_Top",
    landmark_id == "Aardwolf_Rd_T" ~ "Aardwolf_Trees",
    landmark_id == "Diamond_X" ~ "Diamond_Crossing",
    landmark_id == "Gomer's_D" ~ "Gomer_den",
    landmark_id == "Lion_L" ~ "Lion_Lugga",
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing", 

    TRUE ~ landmark_id   # keep everything else unchanged
  )) |>
  
  # Filling up the NAs
  left_join(DF_lndmk |>
      select(landmark_id, utme_df2 = utme, utmn_df2 = utmn) |>
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2) |>
  
   #getting rid of landmark that I have no idea what they are..
  filter(landmark_id != "Martins_Den") |>
  filter(landmark_id != "Kit_Kat_Den_2") |>
  filter(landmark_id != "ccb") |>
  filter(landmark_id != "Precipice_Den_3") |>
  filter(landmark_id != "Albatross_Den_2") |>
  filter(landmark_id != "corner_of_baboon_culdesec") |># double check for Cul_de_sac
  filter(landmark_id != "Baboon_Culdesac") |>
  filter(landmark_id != "Dave_Den") |>
  filter(landmark_id != "Jackal_Hill") |>
  filter(landmark_id != "KCM_D") |>
  filter(landmark_id != "Main_DOC_X") |># we have that location for sure --> verify
  filter(landmark_id != "MB_X") |>
  filter(landmark_id != "Mteer_D" ) |>#Maybe 3 musketeer?? 
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "No_Name_Hill_Culdesac") |>
  filter(landmark_id != "Picasso_Shrub") |>
  filter(landmark_id != "Pond_Crossing") |>
  filter(landmark_id != "Top_???") |>
  filter(landmark_id != "Top_Bane_Lugga") |>
  filter(landmark_id != "Top_Bane_L") |>
  filter(landmark_id != "Eland_Hill") |>
  filter(landmark_id != "676") |>  # no utme and utmn for this in DF_lndmk
  filter(landmark_id != "13_Lugga_South") |>
  filter(landmark_id != "Base_of_Jackel_Hill_&_top_of_Den_One_Creek") |>
  filter(landmark_id != "Den_Six") |>
  filter(landmark_id != "HML_ND") |>
  filter(landmark_id != "JH_T") |>
  filter(landmark_id != "kindt") |>
  filter(landmark_id != "NCCPA") |>
  filter(landmark_id != "NCM_D") |>
  filter(landmark_id != "Pilot's_D") |>
  filter(landmark_id != "Top_E_Fork_HS_L")

  

# important to remember those!! it's probably the sessions in 1988-2000 where the GPS locations was not associated to the sessions.
DB_NAs <- DB_all_locations |>
  filter(if_all(c(landmark_id, utme, utmn), is.na)) #51 267 obs

```

Now we have to deal with the weird coordinates that will create fuckall points.

ex: Reach_Out_and_Touch_Mound
```{r Weird coordinates}
DB_locations_for_MCPs_clean <- DB_locations_for_MCPs |>
  
  left_join(DF_lndmk %>%
      select(landmark_id, utme_new = utme, utmn_new = utmn) %>%
      distinct(landmark_id, .keep_all = TRUE),
    by = "landmark_id") |>
  
  # fix utme/utmn when they have placeholder values (11111, 111111, 1111111)
  mutate(
    utme = if_else(utme %in% c("11111", "111111", "1111111"), utme_new, utme),
    utmn = if_else(utmn %in% c("11111", "111111", "1111111"), utmn_new, utmn)) |>
  
  # fix utme/utmn when utmn is "1111111"
  mutate(
    utme = if_else(utmn == "1111111", utme_new, utme),
    utmn = if_else(utmn == "1111111", utmn_new, utmn)) |>
  
  # fix utme/utmn when both are 0 for specific landmarks
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utme_new, utme),
    utmn = if_else(
      landmark_id %in% c("Stupid_Tree", "Rhino_Tree", "Saga_Lugga_Crossing") &
        utme == 0 & utmn == 0,
      utmn_new, utmn)) |>
  
  # fix specific 98... value for Fig_Tree_Lugga_-_Top
  mutate(
    utme = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utme_new, utme),
    utmn = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == "9840057", utmn_new, utmn)) |>
  
  # drop the helper columns
  select(-utme_new, -utmn_new) |>
  
  # change the airstrip_den utme and utmn value with a comma
  mutate(
    utme = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utme)),   # take everything after the comma
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>
  
 # change the session t99585 and t60813 and swap 4 and 7 in utme
  mutate(
    utme = if_else(
      landmark_id %in% c("Stupid_Tree", "Green_Den") & grepl("^47", utme),
      sub("^47", "74", utme),
      utme)) |>
  
  #same for session t116873, t116658 and t54139 (change 5 and 7)
   mutate(
    utme = if_else(
      landmark_id %in% c("Croton_Island-East_Edge", "Cow_Tree", "Hi_Step_X") & grepl("^57", utme),
      sub("^57", "75", utme),
      utme)) |>

  #remove the last digit of a 7 digit utme
  mutate(
  utme = if_else(
    landmark_id %in% c("Big_Sky_Den", "The_Observatory", "BB_X") & nchar(utme) == 7,
    substr(utme, 1, 6),   # keep only first 6 digits
    utme)) |>
  
  #for session t117205, add a 4 after the first 7, and session t111704 add a 5 after the 7
  mutate(
    utme = if_else(
      landmark_id == "Laura_Tree" & utme == "79731",
      sub("^7", "74", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Sunshine_Tree" & utme == "70360",
      sub("^7", "75", utme),
      utme)) |>
  
  # remove the 2nd number 
  mutate(
  utme = if_else(
    session %in% c("t112413", "t51456"),
    paste0(substr(utme, 1, 1), substr(utme, 3, nchar(utme))),  # remove 2nd character
    utme)) |>
  
  # remove all that is after the coma for both utme and utmn
  mutate(
    across(
      c(utme, utmn),
      ~ if_else(
        landmark_id %in% c("Serena_Lodge", "Box_Crossing", "Surprise_X"),
        sub(",.*", "", .x),   # keep only before first comma
        .x))) |>
  
  # For this session "t49599" 
  # 745095 8906 983
  mutate(
    utme = if_else(
      session == "t49599",
      substr(utme, 1, 6),   # keep only the first 6 characters
      utme)) |>
  mutate(
    utmn = if_else(
      session == "t49599",
      paste0("9806", utmn),   # add 9806 before in utmn
      utmn)) |>
  
  # keep just after the 2nd comma (t22224)
  mutate(
    utme = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utme),   # keep everything after 2nd comma
      utme),
    utmn = if_else(
      session == "t22224",
      sub(".*?,.*?,", "", utmn),
      utmn)) |>
  
  #Get rid of the last digit
   mutate(
    utme = if_else(
      session %in% c("t47831", "t49436", "t48055", "t111641", "t72442", "t57018", "t120547", "t115642"),   # replace with your session IDs
      substr(utme, 1, nchar(utme) - 1),  # drop last character
      utme)) |>
  
  # Now these have only 5 digits in utme, i will just remove them
  filter(!session %in% c("t115558", "t49400", "t66664", "t74990")) |> #I will see later if I decide to look for the last digit or not 
  
  #take out the sessions with weird utmn (99 and 0)
  filter(session != "t99428") |>
  filter(session != "t54706") |>
  #take out the Low_Ranker's_Crossing cause utmn= utme
  filter(landmark_id != "Low_Ranker's_Crossing") |>
  
  # Keep only before the comma 
  mutate(
    utmn = if_else(
      landmark_id %in% c("Uproar_Den", "Fig_Tree_Lodge", "Virginia_Maria_Crossing", "Three_Sisters", "Green_Den"),  #t60159
      sub(",.*", "", utmn),   # keep before the first comma
      utmn)) |>
  
  # Now keep all after the comma
  mutate(
    utmn = if_else(
      landmark_id %in% c("Swakey_Tree"),
      sub("^[^,]*,", "", utmn),   # remove everything up to the first comma
      utmn)) |>
  
  # remove all utmn of 6 and 8
  filter(nchar(utmn) != 6) |>
  filter(nchar(utmn) != 8) |>

  #change the utmn starting with 6 or 7 to 9
  mutate(
    utmn = if_else(
      grepl("^[67]", utmn),   # starts with 6 or 7
      sub("^[67]", "9", utmn),
      utmn))
    
  
```





To get utme and utmn values of + or - than 6 numbers:
```{r Check}

DB_check_utme <- DB_locations_for_MCPs_clean |>
  filter(nchar(utme) != 6) # now it is 0

DB_check_utmn <- DB_locations_for_MCPs_clean |>
  filter(nchar(utmn) != 7) # now is 0

```






Now we visualize to see if there are still weird outliers points:
```{r}
# first convert in sf object :
DB_locations_for_MCPs_sf <- st_as_sf(DB_locations_for_MCPs_clean, coords = c("utme", "utmn"), crs = 32736)

# map to visualize
mapview(DB_locations_for_MCPs_sf, xcol = "utme", ycol = "utmn")
# this is to see the outliers

```
The points that looks like outliers:
- t120888
- s58
- t55969
- t112359
- s6974
- s115706
- s117340 



#2. Shapefiles

SHP file contains geometrical information (for example shape type and coordinates). DBF file contains additional attribute information (for example name or temperature)

We use St_as_sf() function from sf package to create an sf object directly from a data frame containing a column of type sfc (simple feature list-column). Here, the utme and utmn are coordinates and so can be transformed into an sfc column.

You can also create and sf object with st_sf, where you will have a sfc object stored in an r object and then its attributes and you can combien them into an sf object : "pts_sf \<- st_sf(pts_data, geometry = pts)" where pts = sfc of coordinates pts_data = attributes

```{r shp in UTM}
# Turning the points into coordinates to put in a map

#Convert the dataframe into a sf object:
sf_df <- st_as_sf(CD_for_map, coords = c("utme", "utmn"), crs = 32736) #EPSG code is 32736, UTM Zone 36 South


#Save the sf dataframe - can only do it once when it is not created before
CD_sf <- sf::st_write(sf_df, "data_files/CD_sf.shp")

CD_sf <- sf::st_read("data_files/CD_sf.shp")
CD_sf
```

Now, CD_sf is a sf object because we have the feature geometrics (POINTS here) with their related attributes (year, landmark_id, clan, etc.).

```{r}

str(CD_sf)
# our sf object is a list of POINTS

# To extract the specific item of the list of points
CD_sf$geometry[[2]] # give: POINT (717702 9849474)

class(CD_sf$geometry[[2]]) # give : [1] "XY"    "POINT" "sfg"  

#To change the name of the "geometry column": (don'T use names() function, but st_geometry)
#st_geometry(CD_sf) <- "Den_location"

```

Also, to use leaflet (to make maps), we need to change the UTM coordinates into long lat : Here we transform the CD_sf into a WGS84 to use it with leaflet.

```{r shp in WGS84}

CD_sf_wgs84 <- st_transform(CD_sf, crs = 4326)
#4326 is the EPSG code for the WGS84 system, which uses degrees of latitude and longitude.

# save the lat long data points into a shapefile 
CD_sf_wgs84 <- sf::st_write(CD_sf_wgs84, "data_files/CD_sf_wgs84.shp")

#to read the shape file just created
CD_sf_wgs84 <- sf::st_read("data_files/CD_sf_wgs84.shp")
CD_sf_wgs84

#verify it's working:
head(st_coordinates(CD_sf_wgs84))
```


```{r Data manip}

#change the clan column values: we want 1 word per clan 
# Also, here we take out 3 dens that have GPS location not in the right territory

CD_sf <- CD_sf |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(lndmrk_ != "Migration_Den") |>
  filter(lndmrk_ != "Runway_Den") |>
  filter(lndmrk_ != "Mausoleun_at_Halicarnassus_Den")


CD_sf_wgs84 <- CD_sf_wgs84 |>
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(lndmrk_ != "Migration_Den") |>
  filter(lndmrk_ != "Runway_Den") |>
  filter(lndmrk_ != "Mausoleun_at_Halicarnassus_Den")

```






#3. Covariables

Now we need to create covariables.

We have a lot of covariables to create. Here is the list:

-   Distance to human community\*

-   Distance to last and closest lion sighting\*

-   Distance/number (don't think we have that) of cattle\*

-   Number of cubs at dens

-   Land cover (bush vs plains, etc.)\*

-   Altitude\*

-   Distance to edge of territory\*

-   Distance to water\*

-   Amount of time (maybe in number of days?) that the den was used (in interaction with distance to humans?

-   Rank of the females at the den

Easily done/accessible:

-   Season/Month of year

-   Prey abundance

```{r}

```


##i) Distance to Talek

The first thing is to tell R to use python to be able to get the 20+ year historical satellite imagery data from Google Earth Engine, since it is the easiest way to get the Talek town borders for all those years. That API (Application Programming Interface) is only available in Python or JavaScript. So I have to tell R to use Python to have access to that.
  --> An API is like a menu at a restaurant - it lists the things you can ask for (functions, data) and tells the kitchen (the software) how to prepare them for you.
  --> Google Earth Engine API = the set of commands you can send to Google's satellite image servers to request data (e.g., "give me all Landsat images for Talek Town in 2005").


### rgee

Setting up a google earth engine project first.

The first time I need to install conda, create the rgee environment within the conda thing. 

I followed the steps here to have it finally worked: https://gis.stackexchange.com/questions/473730/authentification-error-when-installing-rgee-in-rstudio?newreg=3674f3f594ee4f91856aa2e5df5b204d

Here are the lines that needed to be run the first time, but now I don't need them anymore, just the ones in the code box.

- install_packages("rgee")
- reticulate::conda_install(envname = "rgee", packages = "earthengine-api=0.1.370", channel = "conda-forge", python_version = "3.9")

```{r}
#to get rgeeExtra, to use things like the gif animation
#remotes::install_github("r-earthengine/rgeeExtra")

```


```{r}
library(reticulate)

#This is to tell R to use the right Python environment
reticulate::use_condaenv("rgee", required = TRUE)

library(rgee)
library(rgeeExtra)

#Initialize EE 
ee_Initialize()
#important to initialize earth engine without the user and project argument here:

```

### Checking 
The ee_check() function will help you for checking the sanity of your rgee installation and dependencies.

- ee_check_python() - Python version
- ee_check_credentials() - Google Drive and GCS credentials
- ee_check_python_packages() - Python packages

```{r}
ee_check()

ee_check_python()
ee_check_credentials()
ee_check_python_packages()
```



### Tests
```{r}

# test to make sure all is running smoothly:
ee$Image("CGIAR/SRTM90_V4")$getInfo()
ee$String('Hello from the Earth Engine servers!')$getInfo()

```


#### Night time lights
From : https://github.com/r-spatial/rgee
To Compute the trend of night-time lights

Let's create a new band containing the image date as years since 1991 by extracting the year of the image acquisition date and subtracting it from 1991.
```{r}

createTimeBand <-function(img) {
  year <- ee$Date(img$get('system:time_start'))$get('year')$subtract(1991L) 
  ee$Image(year)$byte()$addBands(img)
}

#Use your TimeBand function to map it over the night-time lights collection.

collection <- ee$
  ImageCollection('NOAA/DMSP-OLS/NIGHTTIME_LIGHTS')$
  select('stable_lights')$
  map(createTimeBand)

# Compute a linear fit over the series of values at each pixel, so that you can visualize the y-intercept as green, and the positive/negative slopes as red/blue.
col_reduce <- collection$reduce(ee$Reducer$linearFit())
col_reduce <- col_reduce$addBands(
  col_reduce$select('scale'))
ee_print(col_reduce)

# to visualize the map: 
Map$setCenter(9.08203, 47.39835, 3)
Map$addLayer(
  eeObject = col_reduce,
  visParams = list(
    bands = c("scale", "offset", "scale"),
    min = 0,
    max = c(0.18, 20, -0.18)
  ),
  name = "stable lights trend"
)

```


#### Worldwide elevation map
From: https://cran.r-project.org/web/packages/rgee/vignettes/rgee01.html

SRTM (Shuttle Radar Topography Mission) elevation values worldwide:
```{r}

#create a ee object
srtm <- ee$Image("USGS/SRTMGL1_003")

#Define visualization parameters
viz <- list(
  max = 4000,
  min = 0,
  palette = c("#000000","#5AAD5A","#A9AD84","#FFFFFF"))

# using Map$addLayer to visualize the map interactively
Map$addLayer(
  eeObject = srtm,
  visParams =  viz,
  name = 'SRTM')

```


#### Precipitation
From: https://github.com/r-spatial/rgee

To play with some precipitation data: 

nc = North Carolina counties, 100 simple features with 14 attributes
```{r}
# to read the nc shapefile:
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

plot(nc)

# will use the Terraclimate database
terraclimate <- ee$ImageCollection("IDAHO_EPSCOR/TERRACLIMATE") %>%
  ee$ImageCollection$filterDate("2001-01-01", "2002-01-01") %>%
  ee$ImageCollection$map(function(x) x$select("pr")) %>% # Select only precipitation bands
  ee$ImageCollection$toBands() %>% # from imagecollection to image
  ee$Image$rename(sprintf("PP_%02d",1:12)) # rename the bands of an image

# to extract monthly precipitation:
ee_nc_rain <- ee_extract(x = terraclimate, y = nc["NAME"], sf = FALSE)

```

ee_extract will help you to extract monthly precipitation values from the Terraclimate ImageCollection. ee_extract works similar to raster::extract, you just need to define: the ImageCollection object (x), the geometry (y), and a function to summarize the values (fun).

```{r}
# then we visualize it using ggplot:
ee_nc_rain %>%
  pivot_longer(-NAME, names_to = "month", values_to = "pr") %>%
  mutate(month, month=gsub("PP_", "", month)) %>%
  ggplot(aes(x = month, y = pr, group = NAME, color = pr)) +
  geom_line(alpha = 0.4) +
  xlab("Month") +
  ylab("Precipitation (mm)") +
  theme_minimal()
```

#### NDVI animation

From: https://github.com/r-spatial/rgee

To create a NDVI animation map (Peru's Arequepia region)
 
```{r}

#Define the regional bounds of animation frames and a mask to clip the NDVI data by.
mask <- system.file("shp/arequipa.shp", package = "rgee") %>%
  st_read(quiet = TRUE) %>%
  sf_as_ee()
region <- mask$geometry()$bounds()


#Retrieve the MODIS Terra Vegetation Indices 16-Day Global 1km dataset as an ee.ImageCollection and then, select the NDVI band.

col <- ee$ImageCollection('MODIS/006/MOD13A2')$select('NDVI')

#Group images by composite date
col <- col$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('day', 'year')
  img$set('doy', doy)
})
distinctDOY <- col$filterDate('2013-01-01', '2014-01-01')

```

Now, let's define a filter that identifies which images from the complete collection match the DOY from the distinct DOY collection.
```{r}
filter <- ee$Filter$equals(leftField = 'doy', rightField = 'doy')

```
Define a join and convert the resulting FeatureCollection to an ImageCollection... it will take you only 2 lines of code!

```{r}
join <- ee$Join$saveAll('doy_matches')
joinCol <- ee$ImageCollection(join$apply(distinctDOY, col, filter))
```

Apply median reduction among the matching DOY collections.

```{r}
comp <- joinCol$map(function(img) {
  doyCol = ee$ImageCollection$fromImages(
    img$get('doy_matches')
  )
  doyCol$reduce(ee$Reducer$median())
})

#define RGB visualization parameters first.
visParams = list(
  min = 0.0,
  max = 9000.0,
  bands = "NDVI_median",
  palette = c(
    'FFFFFF', 'CE7E45', 'DF923D', 'F1B555', 'FCD163', '99B718', '74A901',
    '66A000', '529400', '3E8601', '207401', '056201', '004C00', '023B01',
    '012E01', '011D01', '011301'
  )
)

#Create RGB visualization images for use as animation frames.
rgbVis <- comp$map(function(img) {
  do.call(img$visualize, visParams) %>%
    ee$Image$clip(mask)
})


#GIF visualization parameters.
gifParams <- list(
  region = region,
  dimensions = 600,
  crs = 'EPSG:3857',
  framesPerSecond = 10
)

# get month names: 
dates_modis_mabbr <- distinctDOY %>%
  ee_get_date_ic %>% # Get Image Collection dates
  '[['("time_start") %>% # Select time_start column
  lubridate::month() %>% # Get the month component of the datetime
  '['(month.abb, .) # subset around month abbreviations
```

Finally, we use ee_utils_gif_* functions to render the GIF animation and add some texts.
```{r}

animation <- ee_utils_gif_creator(rgbVis, gifParams, mode = "wb")
animation %>%
  ee_utils_gif_annotate(
    text = "NDVI: MODIS/006/MOD13A2",
    size = 15, color = "white",
    location = "+10+10"
  ) %>%
  ee_utils_gif_annotate(
    text = dates_modis_mabbr,
    size = 30,
    location = "+290+350",
    color = "white",
    font = "arial",
    boxcolor = "#000000"
  ) # -> animation_wtxt

#to save it
# ee_utils_gif_save(animation_wtxt, path = "raster_as_ee.gif")

```



### TALEK 
```{r}
# 1. Location of Talek town, Maasai Mara
talek_point <- ee$Geometry$Point(35.216423, -1.442912) #talek town center
talek_roi <- talek_point$buffer(600) # 5 km radius
```

Here is a test for year 2020:
```{r}
# 2 - Function for built-up detection
get_yearly_builtup <- function(year) {
  s2 <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(talek_roi)$
    filterDate(sprintf("%d-01-01", year), sprintf("%d-12-31", year))$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
    median()
  
  # Calculate NDBI (built-up) and NDVI (vegetation)
  ndbi <- s2$normalizedDifference(c("B11", "B8"))
  ndvi <- s2$normalizedDifference(c("B8", "B4"))
  
  # Threshold: built-up AND low vegetation
  builtup_mask <- ndbi$gt(0.2)$And(ndvi$lt(0.3))
  
  # Mask original image
  builtup <- builtup_mask$selfMask()
  
  # Reduce to vectors inside ROI
  vectors <- builtup$reduceToVectors(
    geometry = talek_roi,
    geometryType = "polygon",
    reducer = ee$Reducer$countEvery(),
    scale = 10,
    maxPixels = 1e10
  )
  
  vectors$set("year", year)
}

# --- Year of interest ---
year <- 2020

# --- Generate polygons for that year ---
builtup_polygons <- get_yearly_builtup(year)

# --- Visualization parameters ---
Map$centerObject(talek_roi, zoom = 15)
Map$addLayer(talek_roi, list(color = "blue"), "Talek ROI") +
  Map$addLayer(builtup_polygons, list(color = "red"), paste("Built-up", year))


```




```{r}
# 2. Function to mask clouds and detect buildings
mask_builtup <- function(image) {
  # Calculate NDBI
  swir <- image$select("B11")
  nir <- image$select("B8")
  ndbi <- swir$subtract(nir)$divide(swir$add(nir))
  
  # Built-up mask (adjust threshold if needed)
  builtup_mask <- ndbi$gt(0.2) # try lowering to 0.1 if empty
  
  # Apply mask but keep it as a single-band image for vectorizing
  return(builtup_mask$rename("builtup"))
}

# 3. Function to get yearly composite and vector polygons
get_yearly_builtup <- function(year) {
  start <- sprintf("%d-01-01", year)
  end <- sprintf("%d-12-31", year)
  
  yearly_img <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(region)$
    filterDate(start, end)$
    map(mask_builtup)$
    max() # union of all built-up pixels in the year
  
  # Ensure projection
  yearly_img <- yearly_img$setDefaultProjection("EPSG:4326", NULL, 10)
  
  # Vectorize built-up areas
  vectors <- yearly_img$reduceToVectors(
    geometry = region,
    geometryType = "polygon",
    scale = 10,
    crs = "EPSG:4326",
    maxPixels = 1e10
  )
  
  vectors_sf <- ee_as_sf(vectors)
  vectors_sf$year <- year
  return(vectors_sf)
}

# 4. Run for a range of years
years <- 2021:2023
builtup_polygons <- do.call(rbind, lapply(years, get_yearly_builtup))

# 5. Save shapefile
st_write(builtup_polygons, "talek_builtup_yearly.shp", delete_dsn = TRUE)
```



### Distances from Talek

```{r}

talek_center <- data.frame(
  lon = 35.216423,
  lat = -1.442912
)

#Convert to sf object
talek_sf <- st_as_sf(talek_center, coords = c("lon", "lat"), crs = 4326)

# Transform to a projected CRS in meters (UTM zone for this area: 36S)
CD_sf_m <- st_transform(CD_sf, 32736)    # EPSG 32736 = WGS 84 / UTM zone 36S
talek_sf_m <- st_transform(talek_sf, 32736)

# Calculate distances in meters
CD_for_map$dist_talek_m <- as.numeric(st_distance(CD_sf_m, talek_sf_m))

```

Saving it :
```{r}

sf_df_m_to_talek <- st_as_sf(CD_for_map, coords = c("utme", "utmn"), crs = 32736) 
#Save the sf dataframe - can only do it once when it is not created before
CD_sf2 <- sf::st_write(sf_df_m_to_talek, "data_files/CD_sf_talekm.shp")

CD_sf2 <- sf::st_read("data_files/CD_sf_talekm.shp")


```



##ii) Lions

This variable is gonna be distance to the closest lion sighting, and/or it will be the number of sightings in the territory per year.

 - Number of lions/sightings
 - Distance from the closest sighting

To do so, I will use the dataframe "tblPredatorsPerSession".

```{r filtering the lions}

DF_lions <- tblPredatorsPerSession |>
  subset(predator == "lion") |>
  select(-c(3, 4, 6:8))

DF_lions$num_individuals <- as.double(DF_lions$num_individuals)
length(unique(DF_lions$session)) #6019

#I want to sum up the number of lions for each session instead of having sessions duplicated rows
DF_lions_sum <- DF_lions |>
  group_by(session) |>
  summarise(num_lion = sum(num_individuals, na.rm = TRUE))
# suppose to have 6019 rows

```

I also need to join the session ID to the location landmark dataframe.

```{r Join Session ID to Landmark location}

DF_lions_loc_data <- tblLocationsPerSession |>
  select(c(1, 5, 6, 7)) |>
  #Here, I want to change the NAs in utme and utmn for the GPS points of the landmark
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_"))



DF_lions_loc <- merge(DF_lions_sum, DF_lions_loc_data, by = "session")

# theres a difference of 302 rows between DF_lions_loc and DF_lions_sum... 
# not sure why
```

I also need to join the tblSessions to get the clan territory and date of the specific sessions ID.


```{r Clan and year}

DF_lions_sess_data <- tblSessions |>
  select(c(1, 2, 4))

# merge the lions data with the session data to get the year and everything
DF_lions_date <- merge(DF_lions_loc, DF_lions_sess_data, by = "session")


DF_lions_clans <- DF_lions_date |>
  # keep only the clans that we are interested in
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra")) |>
  # now we extract the year 
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-c(month, day)) |>
  
  mutate(landmark_id = case_when(
    landmark_id == "Lion_L" ~ "Lion_Lugga", #Lion_L to Lion_Lugga
    landmark_id == "13_Lugga_South" ~ "Culvert_of_13_Lugga_South", #Culvert_of_13_Lugga_South for 13_Lugga_South\
    landmark_id == "Border_T" ~ "Border_Tree",   #Border_T to Border_Tree
    landmark_id == "Gomer's_D" ~ "Gomer_den",   # Gomer's_D to Gomer_den
    landmark_id == "K-6_Den" ~ "K6_Den", #K-6_Den to K6_Den
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing",  #Simba_L_X to Simba_Lugga_Crossing
    TRUE ~ landmark_id   # keep everything else unchanged
  ))
```

Now we add the GPS points for the utme and utmn columns with NAs based on the landmark ID point.

```{r Filling NAs}

# first create a dataframe with all the coordinates for the landmarks:
#that was created on line #331

# this is to put the landmark coordinates where there are NAs in the utme and utmn columns
DF_lions_data <- DF_lions_clans |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2)  |>
  drop_na(utmn) # get rid of NAs for utmn column


# now we change those weird coordinates for the landmark coordinates
DF_lions_data2 <- DF_lions_data |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(
    utme = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utme_df2, utme),
    utmn = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utmn_df2, utmn)
  ) |>
  select(-utme_df2, -utmn_df2) |>
  drop_na(utmn) |>
  mutate(landmark_id = replace_na(landmark_id, "random"))

  
str(DF_lions_data2)

DF_lions_data2$utme <- as.numeric(DF_lions_data2$utme)
DF_lions_data2$utmn <- as.numeric(DF_lions_data2$utmn)

DF_lions_data2 <- DF_lions_data2 |>
  drop_na(utme)

```


I got some num of individual that are 0 :
s114919 --> change to num = 2
t61954 --> change to 1
t67315 --> change to 1
t77054 --> will put 1

remove (they all have 0 as a number of lions -- maybe we will keep them eventually ?)
s165
t101874.10
t1475.4
t53055
t57427
t60317
t62190
t75009
t77236

```{r}

DF_lions_data2 <- DF_lions_data2 |>
  mutate(num_lion = ifelse(session == "s114919", 2, num_lion)) |>
  mutate(num_lion = ifelse(session == "t61954" | 
                             session == "t67315" |
                             session == "t77054", 1, num_lion)) |>
  filter(num_lion != 0)
    

```


Outliers
Change the first 7 in utmn for a 9:
"t102747" & "s109502"

Take out "t64548" because utmn has 6 numbers --> will change later.
same for s0.58

utme miss 1 number:
and t60077

utme has 1 number too many:
t50935
t72442
t98672
t115432

wrong coordinate:  (somewhere in tanzania, so probably a number is off)
s24841
```{r}

DF_lions_data2 <- DF_lions_data2 |>
  mutate(utmn = ifelse(session == "t102747" | session == "s109502", as.numeric(paste0("9", substring(utmn, 2))),
                  utmn))  |>
  filter(session != "t64548") |>
  filter(session != "s0.58") |>
  filter(session != "t60077") |>
  filter(session != "t50935") |>
  filter(session != "t72442") |>
  filter(session != "t98672") |>
  filter(session != "t115432") |>
  filter(session != "s24841")
  

```





Saving it to shp.
```{r}

sf_df_lions <- st_as_sf(DF_lions_data2, coords = c("utme", "utmn"), crs = 32736)

#Save the sf dataframe - can only do it once when it is not created before
CD_sf_lions <- sf::st_write(sf_df_lions, "data_files/CD_sf_lions.shp")

CD_sf_lions <- sf::st_read("data_files/CD_sf_lions.shp")

```

Change to wgs84
```{r}
CD_lions_wgs84 <- st_transform(CD_sf_lions, crs = 4326)
#4326 is the EPSG code for the WGS84 system, which uses degrees of latitude and longitude.

# save the lat long data points into a shapefile 
CD_lions_wgs84 <- sf::st_write(CD_lions_wgs84, "data_files/CD_lions_wgs84.shp")

#to read the shape file just created
CD_lions_wgs84 <- sf::st_read("data_files/CD_lions_wgs84.shp")


```


###data
```{r}

lions_data <- CD_sf_lions |>
  group_by(year, clan) |>
  summarize(n_lion = sum(num_lin))

lions_data <- st_drop_geometry(lions_data)

write.csv2(lions_data, "data_files/lions_data.csv", row.names = F, quote = F)

lions_data_per_clan <- lions_data|>
  group_by(clan) |>
  summarize(n_lion = sum(n_lion))

flextable(lions_data_per_clan)

```





