---
output: html_document
title: "MHP Special Project - RSF"
author: "Anne Ju Laberge"
date: "2025"
format:
  html:
    
    code-fold: true
    output-fold: true
    code-tools: true
    code-link: true
    code-summary: "Show the code"
    theme: cosmo
    embed-resources: true
    self-contained: true
    toc-depth: 5
    toc: true
    toc-float: true
editor: visual
editor_options:
   
   chunk_output_type: console
execute:
  warning: false
  error: true
---

# 0. Data

Load packages

```{r Packages}

# for data manipulations
library(readr) # load data
library(dplyr) # data manip

#library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(viridis) # for viridis color palette
library(tibble)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt) # to calculate home ranges (KDE)

#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For RSF
library(amt) # for the RSF
library(adehabitatHR) # for mcp
library(brms)
library(glmmTMB) # to make RSF models


#install.packages("glmmTMB")

```

Load data files

```{r Files}
#shapefiles created in CD_data

#This one has the coordinates as UTM x, y
CD_sf <- sf::st_read("data_files/CD_sf.shp")
CD_sf

#This one has the coordinates in lat long, to use with leaflet
CD_sf_wgs84 <- sf::st_read("data_files/CD_sf_wgs84.shp")
CD_sf_wgs84

# FOr the lions data
CD_sf_lions <- sf::st_read("data_files/CD_sf_lions.shp")
CD_sf_lions

CD_lions_wgs84 <- sf::st_read("data_files/CD_lions_wgs84.shp")

```


# 1. Data visualisation

## 1.1. Maps

Now let's try to create a map...

### i) Base map (mapview)

Mapview is just to see real quick what the points look like.

```{r Mapviews}

#According to year
mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "year")


# According to clan 
mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "clan")


```

### ii) Basemap (leaflet)

Leaflet is a bit more complex than mapview and can get you more options for mapping.

We will create a basemap for starter: - Leaflet tile : draws the map tiles. There are numerous types of map tiles. - Layers Control: Leaflet has a nice little control that allows your users to control which layers they see on your map

```{r Leaflet per year}

pal <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$clan)

CD_leaflet_inter <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years <- unique(CD_sf_wgs84$year)

for (yr in years) {
  data_year <- CD_sf_wgs84 |> filter(year == yr)
  
CD_leaflet_inter <- CD_leaflet_inter |>
    addCircleMarkers(data = data_year,
                     color = ~pal(clan),
                     group = yr,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan))
}

#Add toggle control for years
CD_leaflet_inter <- CD_leaflet_inter |>
  
  addLayersControl(
    baseGroups = years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal, values = CD_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

CD_leaflet_inter  # display the map

```

Saving as a html file

```{r}

htmlwidgets::saveWidget(CD_leaflet_inter, file = "Maps/CD_leaflet_inter.html", selfcontained = TRUE)

```

### iii) Basemap clan (leaflet)

This is the same map, but we invert the clan and year variables

```{r Leaflet per clan}

pal_year <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$year)

CD_leaflet_year <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
clans <- unique(CD_sf_wgs84$clan)

for (cl in clans) {
  data_clan <- CD_sf_wgs84 |> filter(clan == cl)
  
CD_leaflet_year <- CD_leaflet_year |>
    addCircleMarkers(data = data_clan,
                     color = ~pal_year(year),
                     group = cl,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year,"<br>",
                                    "Clan:", clan, "<br>", 
                                    "Name:", lndmrk_))
}

#Add toggle control for years
CD_leaflet_year <- CD_leaflet_year |>
  
  addLayersControl(
    baseGroups = clans,
    overlayGroups = c("Clan"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal_year, values = CD_sf_wgs84$year,
            title = "Year", opacity = 1)
  

CD_leaflet_year  # display the map

```

Save

```{r}

htmlwidgets::saveWidget(CD_leaflet_year, file = "Maps/CD_leaflet_year.html", selfcontained = TRUE)

```

### iv) Lions map

```{r}

str(CD_lions_wgs84)

# Add lion points in red, sized by number of lions
CD_leaflet_lion <- CD_leaflet_inter |>
  addCircleMarkers(data = CD_lions_wgs84,
                   color = "red",
                   group = CD_lions_wgs84$year,
                   radius = ~num_lin,   
                   fillOpacity = 0.8,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lin, "<br>",
                                  "Session ID:", session))

CD_leaflet_lion


# not sized by number:
CD_leaflet_lion2 <- CD_leaflet_inter |>
  addCircleMarkers(data = CD_lions_wgs84,
                   color = "red",
                   group = CD_lions_wgs84$year,
                   fillOpacity = 0.8,
                   radius = 1,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lin, "<br>",
                                  "Session ID:", session))

CD_leaflet_lion2

#save
htmlwidgets::saveWidget(CD_leaflet_lion2, file = "Maps/CD_leaflet_lion2.html", selfcontained = TRUE)


```

# 2. KDE by clans

Here we will make KDE, kernel density estimation.

## 2.1. Functions

First, the functions we'll need: - Function to calculate the KDE - Function to calculate the KDE with two groups (clan per year example) - Function to extract the KDE values and then use it in maps.

```{r Function KDE for clans}
# Function to calculate 50% KDE for each clan
calculate_kde <- function(df) {
  hr_kde(df, levels = 0.95)
}
```

```{r Function KDE for yearly}

# Function to calculate 95% KDE for each group
calculate_kde_2 <- function(df) {
  if (nrow(df) < 5) return(NULL)  # Skip if not enough points
  df <- df[complete.cases(df$x_, df$y_), ]  # Remove rows with missing coords
  if (nrow(df) < 5) return(NULL)  # Skip again if filtering removed too much
  tryCatch(
    hr_kde(df, levels = 0.95)
  )
}

```

```{r Function to extract KDE}
# Extract KDE polygons and convert to sf format
extract_kde_sf <- function(kde_result) {
  kde_sf <- st_as_sf(hr_isopleths(kde_result))# Convert KDE to sf polygons
  return(kde_sf)
}
```

## 2.2. Data

```{r Organising data}
# First step : make_track
# Make_track the GPS dataset to then be able to calculate KDE

#Preparing the df for it:
CD_sf_df <- CD_sf_wgs84 |>
  # before making the make_track, we need to separate geometry column into long and lat 
  mutate(long = sf::st_coordinates(geometry)[,1],
         lat = sf::st_coordinates(geometry)[,2]) |>
  mutate(row.id = row_number()) |>
  dplyr::select(year, long, lat, clan, row.id)


#Now we make the make_track
CD_KDE_df <- CD_sf_df |>
  #Now we do the make_track function
  make_track(long, lat, crs = 4326, id = clan) |>
  #Now rename the id column just because
  rename(clan = id) |>
  mutate(row.id = CD_sf_df$row.id)
  
#Now to add year:
CD_year <- CD_sf_df |>
  dplyr::select(year, row.id) |>
  st_drop_geometry()

#merge year with the lat long df:
CD_KDE_df <- CD_KDE_df |>
  left_join(CD_year, by = "row.id")
```

```{r Split}
# Split the dataset per Clan
CD_KDE_list <- split(CD_KDE_df, CD_KDE_df$clan)

# Split the db into clan and year
CD_KDE_list_cy <- split(CD_KDE_df, CD_KDE_df$year)
```

## 2.3. Clans' KDE

```{r Calculating}
CD_clan_KDE <- lapply(CD_KDE_list, calculate_kde)
```

```{r Extract+Data manip}
CD_clan_KDE_data <- lapply(CD_clan_KDE, extract_kde_sf)

CD_clan_KDE_mapdata <- do.call(rbind, CD_clan_KDE_data)

CD_clan_KDE_mapdata <- tibble::rownames_to_column(CD_clan_KDE_mapdata, "clan")



```

```{r Leaflet}

pal_clan <- colorFactor(palette = "viridis", domain = CD_clan_KDE_mapdata$clan)


CD_clan_KDE_map <- CD_leaflet_year |>
  addPolygons(data = CD_clan_KDE_mapdata,
              fillColor = ~pal_clan(CD_clan_KDE_mapdata$clan),
              color = "black",
              weight = 1,
              fillOpacity = 0.8,
              label = ~paste("Clan:", clan)) 
  

CD_clan_KDE_map

```

```{r}
htmlwidgets::saveWidget(CD_clan_KDE_map, file = "Maps/CD_clan_KDE_map.html", selfcontained = TRUE)
```

Save KDEs data

```{r Save}

CD_clan_KDE_mapdata <- sf::st_write(CD_clan_KDE_mapdata, "data_files/CD_clan_KDE.shp")


```

# 3. MCP by clans

To make our RSF model, we need to create random points of potential locations of where the hyenas could have denned. To do that, we first need to know the extent of the clans' territory, and so we will calculate the MCP (Minimum Convex Polygon) for each clan. An MCP calculate the area that an animal uses. The MCP essentially draws a polygon around all the points where an animal has been observed.

Also, we are using the 95% MCP, so that it removes the 5% of outlier points.

```{r Individual MCPs for all clans}

#for Serena South clan
CD_sf_serenas <- CD_sf_wgs84 |>
  filter(clan == "serenas")

serenas_points <- as_Spatial(CD_sf_serenas$geometry)

serenas_MCP <- mcp(serenas_points, percent = 95)



# For Serena North clan
CD_sf_serenan <- CD_sf_wgs84 |>
  filter(clan == "serenan")

serenan_points <- as_Spatial(CD_sf_serenan$geometry)

serenan_MCP <- mcp(serenan_points, percent = 95)



# For Happy Zebra clan 
CD_sf_hz <- CD_sf_wgs84 |>
  filter(clan == "happyzebra")

hz_points <- as_Spatial(CD_sf_hz$geometry)

hz_MCP <- mcp(hz_points, percent = 95)


# For Happy Zebra clan 
CD_sf_talek <- CD_sf_wgs84 |>
  filter(clan == "talek")

talek_points <- as_Spatial(CD_sf_talek$geometry)

talek_MCP <- mcp(talek_points, percent = 95)

```

```{r Plots}
plot(serenan_MCP)
plot(serenas_MCP)
plot(hz_MCP)
plot(talek_MCP)

```

```{r MCPs grouped by clans}

#grouping the mcps in a dataframe grouped by clans
clan_MCPs <- rbind(serenan_MCP, talek_MCP, serenas_MCP, hz_MCP)

#rename the ids :
new_ids <- c("serenan", "talek", "serenas", "happyzebra")

clan_MCPs <- sp::spChFIDs(clan_MCPs, new_ids)

```

```{r}

plot(clan_MCPs)

```

```{r Plot+leaflet}

clan_MCPs_dataframe <- as.data.frame(clan_MCPs)

clan_MCPs_dataframe <- rownames_to_column(clan_MCPs_dataframe, "clan")

MCPs_clan <- unique(clan_MCPs_dataframe$clan)

MCPs_plot <- CD_leaflet_year |>
  addPolygons(data = clan_MCPs,
              color = "red",
              fillColor = "none",
              weight = 2,
              fillOpacity = 0.05)
  
MCPs_plot

```

Save

```{r}

clan_MCPs <- sf::st_write(clan_MCPs, "data_files/clan_MCPs.shp")

```

#4. Creating the RSF

ALSO, we need to control for the obs effort (some years field team would go more on obs vs other years/season). I guess with year as a random, and month as a covariable, it does control for this...

##4.1. Prepare data

The first step in doing a RSF, is to create random points on the territory. To do so, I will take the clans' KDEs as the home range of the clans, and generate random points in the polygons created. This will create a database of used and available points. The used points are the den locations known, and the available are probably all the adjacent points on the territory.

I need to determine how many points I want to sample for each points used. For example, 5 points per used points.

The `random_points()` function will create this `case\_` variable that will be use in the RSF later.

```{r Random_points}

#Convert your sf object to a data frame with coordinates:
CD_sf_wgs84_df <- CD_sf_wgs84 |>
  mutate(long = sf::st_coordinates(geometry)[,1],
         lat = sf::st_coordinates(geometry)[,2])


#Change the CD_sf_wgs84 into a track_xy object
presence_track <- make_track(CD_sf_wgs84_df, long, lat, crs = 4326, id = clan, year = year)

#serenan presence_track
presence_track_serenan <- presence_track |>
  subset(id == "serenan")

#serenan presence_track
presence_track_serenas <- presence_track |>
  subset(id == "serenas")

#serenan presence_track
presence_track_hz <- presence_track |>
  subset(id == "happyzebra")

#serenan presence_track
presence_track_talek <- presence_track |>
  subset(id == "talek")


#creating the dataset to use in RSF after (for all clans)
rp_clans_KDE <- random_points(CD_clan_KDE_mapdata, n=4000, presence = presence_track)

#Visualising
plot(rp_clans_KDE)


```

Here, I will do rp datasets for the 4 clans, to make sure the number of available points is 10 times de number of used points.

```{r RP Per clans}

#First, how much CD locations have each clans
sum(CD_sf_wgs84$clan == "serenan") # 93
sum(CD_sf_wgs84$clan == "serenas") # 92
sum(CD_sf_wgs84$clan == "happyzebra") # 83
sum(CD_sf_wgs84$clan == "talek") # 90

#serenan
rp_serenan_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "serenan") |>
  random_points(n=930, presence = presence_track_serenan) |>
  mutate(clan = "serenan")


#serenas
rp_serenas_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "serenas") |>
  random_points(n=920, presence = presence_track_serenas) |>
  mutate(clan = "serenas")

#happyzebra
rp_hz_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "happyzebra") |>
  random_points(n = 830, presence = presence_track_hz) |>
  mutate(clan = "happyzebra")

#talek
rp_talek_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "talek") |>
  random_points(n = 900, presence = presence_track_talek) |>
  mutate(clan = "talek")


# combine all of the clans:
rp_ind_clans_KDE <- rbind(rp_serenan_KDE, rp_serenas_KDE, rp_hz_KDE, rp_talek_KDE)

```

Save the rp dataset:

```{r Save}

rp_ind_clans_KDE <- st_write(rp_ind_clans_KDE, "data_files/rp_ind_clans_KDE.shp")

```

## 4.2. Plots

```{r Plot leaflet}

pal_clans_rp <- colorFactor(palette = "magma", domain = rp_ind_clans_KDE$clan)

rp_KDEs_plot <- CD_clan_KDE_map |>
  addCircleMarkers(data = rp_ind_clans_KDE,
                   lat = rp_ind_clans_KDE$y_,
                   lng = rp_ind_clans_KDE$x_,
                   group = rp_ind_clans_KDE$clan,
                   color = ~pal_clans_rp(clan),
                   radius = 1,
                   fillOpacity = 0.1)

rp_KDEs_plot

htmlwidgets::saveWidget(rp_KDEs_plot, file = "Maps/rp_KDEs_plot.html", selfcontained = TRUE)

```

## 4.3. RSF dataset

Now, we can put `rp_ind_clans_KDE` into a dataset. This will be to use in the RSF.

```{r RSF_data}

#change the case_ column to used and 0 = not used, 1 = used
RSF_data <- rp_ind_clans_KDE |>
  rename(used = case_) |>
  mutate(used = case_when(used == 'TRUE' ~ 1,
                          TRUE ~ 0))


```

```{r Save}

#RSF_data <- sf::st_write(RSF_data, "data_files/rp_ind_clans_KDE.shp")
RSF_data <- st_read("data_files/rp_ind_clans_KDE.dbf")

```


###Talek distance 

```{r}

RSF_data_sf <- st_as_sf(RSF_data[, c("x_", "y_", "clan", "used")], coords = c("x_", "y_"), crs = 4326)

mapview(RSF_data_sf)

talek_center <- data.frame(
  lon = 35.216423,
  lat = -1.442912
)

#Convert to sf object
talek_sf <- st_as_sf(talek_center, coords = c("lon", "lat"), crs = 4326)

# Transform to a projected CRS in meters (UTM zone for this area: 36S)
RSF_data_sf_m <- st_transform(RSF_data_sf, 32736)    # EPSG 32736 = WGS 84 / UTM zone 36S
talek_sf_m <- st_transform(talek_sf, 32736)

# Calculate distances in meters
RSF_data_sf$dist_talek_m <- as.numeric(st_distance(RSF_data_sf_m, talek_sf_m))

mapview(RSF_data_sf)

```


## 4.2. RSF model

I will separate the Talek points with the triangle ones, so that I can make a model with the Talek town borders as a variable.

```{r}

rsf1 <- fit_rsf(RSF_data, formula = used ~ 1)
summary(rsf1)

```
