---
output: html_document
editor_options: 
  chunk_output_type: console
---


# Resource Selection Function - Drivers of communal den location

# 0. Libraries et al.

Load packages

```{r Packages}

# for data manipulations
library(readr) # load data
library(dplyr) # data manip

#library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(viridis) # for viridis color palette
library(tibble)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt) # to calculate home ranges (KDE)

#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For RSF
library(amt) # for the RSF
library(adehabitatHR) # for mcp
library(brms)
library(glmmTMB) # to make RSF models
library(lme4)
library(sjPlot)

library(purrr)


#install.packages("glmmTMB")

```

Load data files

```{r Files}
#shapefiles created in CD_data

# To make the MCP:
DB_AllGPS_sf <- sf::st_read("data_files/DB_AllGPS_sf.shp")
DB_AllGPS_sf_wgs84 <- sf::st_read("data_files/DB_AllGPS_sf_wgs84.shp")



# Communal den GPS points:
CD_sf <- sf::st_read("data_files/CD_sf.shp")
CD_sf_wgs84 <- sf::st_read("data_files/CD_sf_wgs84.shp")


# Lions data
CD_sf_lions <- sf::st_read("data_files/CD_sf_lions.shp")
CD_lions_wgs84 <- sf::st_read("data_files/CD_lions_wgs84.shp")


RandomPointsCD_landcover <- sf::st_read("data_files/RandomPointsCD_landcover.shp")
```


```{r Files}
CD_for_map <- readr::read_csv2("data_files/CD_for_map.csv")

```


# 1. Modelling the RSF

ALSO, we need to control for the obs effort (some years field team would go more on obs vs other years/season). I guess with year as a random, and month as a covariable, it does control for this...

************
7) Quick tips for RSF modelling

Treat landcover as a categorical predictor (factor) in your logistic/conditional logistic RSF (used vs available).

If you used a very fine raster (10 m) but your hyena den location accuracy is coarse (GPS error), consider buffering points or using a majority focal window to reduce misclassification from single-pixel noise.

If you want multi-year matching (landcover year of den use), pick the dataset year accordingly (Copernicus 2019 / WorldCover 2020 / MODIS yearly).
************





## 1.1. RSF model

I will separate the Talek points with the triangle ones, so that I can make a model with the Talek town borders as a variable.


With brms:
```{r brms}

brm_RSF <- brms::brm(case ~ landcover + (1|year) + (1|clan),
                     data = RandomPointsCD_landcover, 
                     family = bernoulli(),
                     cores = 4, 
                     chains = 4)


```



## 1.2. Distance to Talek model

```{r brms tlaek model}
CD_RSF_data_tl <- CD_RSF_data |>
  filter(clan == "talek")


brm_RSF_tl <- brm(case ~ dst_tl + year + (1|year),
                  data = CD_RSF_data_tl, 
                  family = bernoulli(),
                  cores = 4, 
                  chains = 4)

brm_RSF_tl <- readRDS("brms_models/brm_RSF_tl.rds")

```

```{r}

summary(brm_RSF_tl)
plot(conditional_effects(brm_RSF_tl), points = T)


```



# 2. Lion RSF




