---
output: html_document
title: "MHP Special Project - RSF"
author: "Anne Ju Laberge"
date: "2025"
format:
  html:
    
    code-fold: true
    output-fold: true
    code-tools: true
    code-link: true
    code-summary: "Show the code"
    theme: cosmo
    embed-resources: true
    self-contained: true
    toc-depth: 5
    toc: true
    toc-float: true
editor: visual
editor_options:
   
   chunk_output_type: console
execute:
  warning: false
  error: true
---

# 0. Data

Load packages

```{r Packages}

# for data manipulations
library(readr) # load data
library(dplyr) # data manip

#library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(viridis) # for viridis color palette
library(tibble)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt) # to calculate home ranges (KDE)

#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For RSF
library(amt) # for the RSF
library(adehabitatHR) # for mcp
library(brms)
library(glmmTMB) # to make RSF models


library(purrr)


#install.packages("glmmTMB")

```

Load data files

```{r Files}

#shapefiles created in CD_data

# To make the MCP:
DB_AllGPS_sf <- sf::st_read("data_files/DB_AllGPS_sf.shp")
DB_AllGPS_sf_wgs84 <- sf::st_read("data_files/DB_AllGPS_sf_wgs84.shp")



# Communal den GPS points:
CD_sf <- sf::st_read("data_files/CD_sf.shp")
CD_sf_wgs84 <- sf::st_read("data_files/CD_sf_wgs84.shp")

CD_sf_talekm <- sf::st_read("data_files/CD_sf_talekm.shp")
CD_sf_talekm_wgs84 <- sf::st_read("data_files/CD_sf_talekm_wgs84.shp")




# Lions data
CD_sf_lions <- sf::st_read("data_files/CD_sf_lions.shp")
CD_lions_wgs84 <- sf::st_read("data_files/CD_lions_wgs84.shp")



CD_all_data <- sf::st_read("data_files/CD_all_data.shp")


```


Let's get rid of Migration den in happy zebra that is not at all at the right place:
```{r Data manip}



CD_sf_talekm_wgs84 <- CD_sf_talekm_wgs84 |> 
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(lndmrk_ != "Migration_Den")



CD_lions_wgs84 <- CD_lions_wgs84 |> 
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  select(-date)



```



# 1. Data visualisation

## 1.1. Maps

Now let's try to create a map...

### i) Base map (mapview)

Mapview is just to see real quick what the points look like.

```{r Mapviews}

#According to year
mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "year")


# According to clan 
mapview(CD_sf, xcol = "utme", ycol = "utmn", zcol = "clan")


```

### ii) Basemap (leaflet)

Leaflet is a bit more complex than mapview and can get you more options for mapping.

We will create a basemap for starter: - Leaflet tile : draws the map tiles. There are numerous types of map tiles. - Layers Control: Leaflet has a nice little control that allows your users to control which layers they see on your map

```{r Leaflet per year}

pal <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$clan)

CD_leaflet_inter <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years <- unique(CD_sf_wgs84$year)

for (yr in years) {
  data_year <- CD_sf_wgs84 |> filter(year == yr)
  
CD_leaflet_inter <- CD_leaflet_inter |>
    addCircleMarkers(data = data_year,
                     color = ~pal(clan),
                     group = yr,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan))
}

#Add toggle control for years
CD_leaflet_inter <- CD_leaflet_inter |>
  
  addLayersControl(
    baseGroups = years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal, values = CD_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

CD_leaflet_inter  # display the map

```

Saving as a html file

```{r}

htmlwidgets::saveWidget(CD_leaflet_inter, file = "Maps/CD_leaflet_inter.html", selfcontained = TRUE)

```

### iii) Basemap clan (leaflet)

This is the same map, but we invert the clan and year variables

```{r Leaflet per clan}

pal_year <- colorFactor(palette = "viridis", domain = CD_sf_wgs84$year)

CD_leaflet_year <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
clans <- unique(CD_sf_wgs84$clan)

for (cl in clans) {
  data_clan <- CD_sf_wgs84 |> filter(clan == cl)
  
CD_leaflet_year <- CD_leaflet_year |>
    addCircleMarkers(data = data_clan,
                     color = ~pal_year(year),
                     group = cl,
                     radius = 3,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year,"<br>",
                                    "Clan:", clan, "<br>", 
                                    "Name:", lndmrk_))
}

#Add toggle control for years
CD_leaflet_year <- CD_leaflet_year |>
  
  addLayersControl(
    baseGroups = clans,
    overlayGroups = c("Clan"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal_year, values = CD_sf_wgs84$year,
            title = "Year", opacity = 1)
  

CD_leaflet_year  # display the map

```

Save

```{r}

htmlwidgets::saveWidget(CD_leaflet_year, file = "Maps/CD_leaflet_year.html", selfcontained = TRUE)

```

### iv) Lions map

```{r}

str(CD_lions_wgs84)

# Add lion points in red, sized by number of lions
CD_leaflet_lion <- CD_leaflet_inter |>
  addCircleMarkers(data = CD_lions_wgs84,
                   color = "red",
                   group = CD_lions_wgs84$year,
                   radius = ~num_lin,   
                   fillOpacity = 0.8,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lin, "<br>",
                                  "Session ID:", session))

CD_leaflet_lion


# not sized by number:
CD_leaflet_lion2 <- CD_leaflet_inter |>
  addCircleMarkers(data = CD_lions_wgs84,
                   color = "red",
                   group = CD_lions_wgs84$year,
                   fillOpacity = 0.8,
                   radius = 1,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lin, "<br>",
                                  "Session ID:", session))

CD_leaflet_lion2

#save
htmlwidgets::saveWidget(CD_leaflet_lion2, file = "Maps/CD_leaflet_lion2.html", selfcontained = TRUE)


```


### v) All GPS points map

```{r}

pal2 <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

AllGPS_leaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years2 <- unique(DB_AllGPS_sf_wgs84$year)

for (yr in years2) {
  data_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
AllGPS_leaflet <- AllGPS_leaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal2(clan),
                     group = yr,
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan, "<br>", "session", "<br>", "lndmrk_"))
}

#Add toggle control for years
AllGPS_leaflet <- AllGPS_leaflet |>
  
  addLayersControl(
    baseGroups = years2,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal2, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

AllGPS_leaflet  # display the map

#save
htmlwidgets::saveWidget(AllGPS_leaflet, file = "Maps/AllGPS_leaflet.html", selfcontained = TRUE)

```

Location map:

```{r}

pal3 <- colorFactor(c("red", "blue", "yellow", "green", "pink", "black", "orange"), domain = DB_AllGPS_sf_wgs84$locatin)

location_leaflet <- leaflet() |>
  addTiles()

# Add points grouped by Year, colored by Clan
years2 <- unique(DB_AllGPS_sf_wgs84$year)

for (yr in years2) {
  data_year <- DB_AllGPS_sf_wgs84 |> filter(year == yr)
  
location_leaflet <- location_leaflet |>
    addCircleMarkers(data = data_year,
                     color = ~pal3(locatin),
                     group = yr,
                     radius = 1,
                     fillOpacity = 0.8,
                     popup = ~paste("Year:", year, "<br>", "Clan:", clan, "<br>", "Session", session, "<br>", "Landmark", lndmrk_))
}

#Add toggle control for years
location_leaflet <- location_leaflet |>
  
  addLayersControl(
    baseGroups = years2,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal3, values = DB_AllGPS_sf_wgs84$locatin,
            title = "Location", opacity = 1)
  

location_leaflet  # display the map

#save
htmlwidgets::saveWidget(location_leaflet, file = "Maps/location_leaflet.html", selfcontained = TRUE)

```


```{r}

pal_clan <- colorFactor(palette = "viridis", domain = DB_AllGPS_sf_wgs84$clan)

location_leaflet2 <- leaflet() |>
  addTiles()

loc <- unique(DB_AllGPS_sf_wgs84$locatin)

for (lc in loc) {
  data_loc <- DB_AllGPS_sf_wgs84 |> filter(locatin == lc)
  
location_leaflet2 <- location_leaflet2 |>
    addCircleMarkers(data = data_loc,
                     color = ~pal_clan(clan),
                     group = lc,
                     radius = 1,
                     fillOpacity = 1,
                     popup = ~paste("Year:", year,"<br>",
                                    "Clan:", clan, "<br>", 
                                    "Name:", lndmrk_))
}

#Add toggle control for years
location_leaflet2 <- location_leaflet2 |>
  
  addLayersControl(
    baseGroups = loc,
    overlayGroups = c("Location"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  
  addScaleBar(position = "bottomleft", options = list(metric = TRUE, imperial = FALSE)) |>
  
  #Add a legend
  addLegend("bottomright", pal = pal_clan, values = DB_AllGPS_sf_wgs84$clan,
            title = "Clan", opacity = 1)
  

location_leaflet2  # display the map


#save
htmlwidgets::saveWidget(location_leaflet2, file = "Maps/location_leaflet2.html", selfcontained = TRUE)

```



# 2. Creating the RSF

ALSO, we need to control for the obs effort (some years field team would go more on obs vs other years/season). I guess with year as a random, and month as a covariable, it does control for this...



# 3. Covariables

Now we need to create covariables.

We have a lot of covariables to create. Here is the list:

-   Distance to human community\*

-   Distance to last and closest lion sighting\*

-   Distance/number (don't think we have that) of cattle\*

-   Land cover (bush vs plains, etc.)\*

-   Altitude\*

-   Distance to edge of territory\*

-   Distance to water\*

-   Amount of time (maybe in number of days?) that the den was used (in interaction with distance to humans?

-   Rank of the females at the den

In more of a post-hoc thing:
-   Number of cubs at dens



```{r}

```


## 3.1. Distance to Talek

The first thing is to tell R to use python to be able to get the 20+ year historical satellite imagery data from Google Earth Engine, since it is the easiest way to get the Talek town borders for all those years. That API (Application Programming Interface) is only available in Python or JavaScript. So I have to tell R to use Python to have access to that.
  --> An API is like a menu at a restaurant - it lists the things you can ask for (functions, data) and tells the kitchen (the software) how to prepare them for you.
  --> Google Earth Engine API = the set of commands you can send to Google's satellite image servers to request data (e.g., "give me all Landsat images for Talek Town in 2005").


### TALEK 
```{r}
# 1. Location of Talek town, Maasai Mara
talek_point <- ee$Geometry$Point(35.216423, -1.442912) #talek town center
talek_roi <- talek_point$buffer(600) # 5 km radius
```

Here is a test for year 2020:
```{r}
# 2 - Function for built-up detection
get_yearly_builtup <- function(year) {
  s2 <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(talek_roi)$
    filterDate(sprintf("%d-01-01", year), sprintf("%d-12-31", year))$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
    median()
  
  # Calculate NDBI (built-up) and NDVI (vegetation)
  ndbi <- s2$normalizedDifference(c("B11", "B8"))
  ndvi <- s2$normalizedDifference(c("B8", "B4"))
  
  # Threshold: built-up AND low vegetation
  builtup_mask <- ndbi$gt(0.2)$And(ndvi$lt(0.3))
  
  # Mask original image
  builtup <- builtup_mask$selfMask()
  
  # Reduce to vectors inside ROI
  vectors <- builtup$reduceToVectors(
    geometry = talek_roi,
    geometryType = "polygon",
    reducer = ee$Reducer$countEvery(),
    scale = 10,
    maxPixels = 1e10
  )
  
  vectors$set("year", year)
}

# --- Year of interest ---
year <- 2020

# --- Generate polygons for that year ---
builtup_polygons <- get_yearly_builtup(year)

# --- Visualization parameters ---
Map$centerObject(talek_roi, zoom = 15)
Map$addLayer(talek_roi, list(color = "blue"), "Talek ROI") +
  Map$addLayer(builtup_polygons, list(color = "red"), paste("Built-up", year))


```




```{r}
# 2. Function to mask clouds and detect buildings
mask_builtup <- function(image) {
  # Calculate NDBI
  swir <- image$select("B11")
  nir <- image$select("B8")
  ndbi <- swir$subtract(nir)$divide(swir$add(nir))
  
  # Built-up mask (adjust threshold if needed)
  builtup_mask <- ndbi$gt(0.2) # try lowering to 0.1 if empty
  
  # Apply mask but keep it as a single-band image for vectorizing
  return(builtup_mask$rename("builtup"))
}

# 3. Function to get yearly composite and vector polygons
get_yearly_builtup <- function(year) {
  start <- sprintf("%d-01-01", year)
  end <- sprintf("%d-12-31", year)
  
  yearly_img <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(region)$
    filterDate(start, end)$
    map(mask_builtup)$
    max() # union of all built-up pixels in the year
  
  # Ensure projection
  yearly_img <- yearly_img$setDefaultProjection("EPSG:4326", NULL, 10)
  
  # Vectorize built-up areas
  vectors <- yearly_img$reduceToVectors(
    geometry = region,
    geometryType = "polygon",
    scale = 10,
    crs = "EPSG:4326",
    maxPixels = 1e10
  )
  
  vectors_sf <- ee_as_sf(vectors)
  vectors_sf$year <- year
  return(vectors_sf)
}

# 4. Run for a range of years
years <- 2021:2023
builtup_polygons <- do.call(rbind, lapply(years, get_yearly_builtup))

# 5. Save shapefile
st_write(builtup_polygons, "talek_builtup_yearly.shp", delete_dsn = TRUE)
```




### Distance to talek calculation

```{r}

talek_center <- data.frame(
  lon = 35.216423,
  lat = -1.442912
)

#Convert to sf object
talek_sf <- st_as_sf(talek_center, coords = c("lon", "lat"), crs = 4326)

# Transform to a projected CRS in meters (UTM zone for this area: 36S)
CD_all_data_tl <- st_transform(CD_all_data, 32736)    # EPSG 32736 = WGS 84 / UTM zone 36S
talek_sf_m <- st_transform(talek_sf, 32736)


# Calculate distances in meters
CD_all_data$dst_tl <- as.numeric(st_distance(CD_all_data_tl, talek_sf_m))


```



## 5.2. Lions

I want to calculate the distance from the den location points to the closest lion sighting.

```{r}



```





# 6. RSF

Now, we can use `CD_all_data` in the RSF.

```{r RSF data}

CD_RSF_data <- CD_all_data |>
  rename(case = case_)
  
```


```{r Save}

CD_RSF_data <- sf::st_write(CD_RSF_data, "data_files/CD_RSF_data.shp")
CD_RSF_data <- st_read("data_files/CD_RSF_data.shp")

```



## 6.1. RSF model

I will separate the Talek points with the triangle ones, so that I can make a model with the Talek town borders as a variable.

```{r}

rsf1 <- fit_rsf(CD_RSF_data, formula = used ~ 1 + (1|year) + (1|clan), 
                family = binomial)



summary(rsf1)
plot(rsf1)

```

With brms:
```{r brms}


brm_RSF <- brm(case_ ~ 1 + (1|clan) + (1|year),
               data = CD_RSF_data, 
               family = bernoulli(),
               cores = 4, 
               chains = 4)


```



## 6.2. Distance to Talek model
```{r brms tlaek model}
CD_RSF_data_tl <- CD_RSF_data |>
  filter(clan == "talek")


brm_RSF_tl <- brm(case ~ dst_tl + year + (1|year),
                  data = CD_RSF_data_tl, 
                  family = bernoulli(),
                  cores = 4, 
                  chains = 4)

brm_RSF_tl <- readRDS("brms_models/brm_RSF_tl.rds")

```


```{r}

summary(brm_RSF_tl)
plot(conditional_effects(brm_RSF_tl), points = T)


```









