---
output: html_document
editor_options: 
  chunk_output_type: console
---


# Resource Selection Function - Drivers of communal den location

# 0. Libraries et al.

Load packages

```{r Packages}

# for data manipulations
library(readr) # load data
library(dplyr) # data manip

#library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(viridis) # for viridis color palette
library(tibble)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt) # to calculate home ranges (KDE)

#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For RSF
library(amt) # for the RSF
library(adehabitatHR) # for mcp
library(brms)
library(glmmTMB) # to make RSF models


library(purrr)


#install.packages("glmmTMB")

```

Load data files

```{r Files}

#shapefiles created in CD_data

# To make the MCP:
DB_AllGPS_sf <- sf::st_read("data_files/DB_AllGPS_sf.shp")
DB_AllGPS_sf_wgs84 <- sf::st_read("data_files/DB_AllGPS_sf_wgs84.shp")



# Communal den GPS points:
CD_sf <- sf::st_read("data_files/CD_sf.shp")
CD_sf_wgs84 <- sf::st_read("data_files/CD_sf_wgs84.shp")

CD_sf_talekm <- sf::st_read("data_files/CD_sf_talekm.shp")
CD_sf_talekm_wgs84 <- sf::st_read("data_files/CD_sf_talekm_wgs84.shp")




# Lions data
CD_sf_lions <- sf::st_read("data_files/CD_sf_lions.shp")
CD_lions_wgs84 <- sf::st_read("data_files/CD_lions_wgs84.shp")



CD_all_data <- sf::st_read("data_files/CD_all_data.shp")


```

Let's get rid of Migration den in happy zebra that is not at all at the right place:

```{r Data manip}



CD_sf_talekm_wgs84 <- CD_sf_talekm_wgs84 |> 
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  filter(lndmrk_ != "Migration_Den")



CD_lions_wgs84 <- CD_lions_wgs84 |> 
  mutate(clan = str_replace_all(clan, "\\.", "")) |>
  select(-date)



```

# 1. Creating the RSF

ALSO, we need to control for the obs effort (some years field team would go more on obs vs other years/season). I guess with year as a random, and month as a covariable, it does control for this...



## 1.1. RSF model

I will separate the Talek points with the triangle ones, so that I can make a model with the Talek town borders as a variable.

```{r}

rsf1 <- fit_rsf(CD_RSF_data, formula = used ~ 1 + (1|year) + (1|clan), 
                family = binomial)



summary(rsf1)
plot(rsf1)

```

With brms:

```{r brms}


brm_RSF <- brm(case_ ~ 1 + (1|clan) + (1|year),
               data = CD_RSF_data, 
               family = bernoulli(),
               cores = 4, 
               chains = 4)


```

## 1.2. Distance to Talek model

```{r brms tlaek model}
CD_RSF_data_tl <- CD_RSF_data |>
  filter(clan == "talek")


brm_RSF_tl <- brm(case ~ dst_tl + year + (1|year),
                  data = CD_RSF_data_tl, 
                  family = bernoulli(),
                  cores = 4, 
                  chains = 4)

brm_RSF_tl <- readRDS("brms_models/brm_RSF_tl.rds")

```

```{r}

summary(brm_RSF_tl)
plot(conditional_effects(brm_RSF_tl), points = T)


```



# 2. Lion RSF




