---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Creating the covariables 

# 0. Libraries et al.

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

```

```{r Load MHP tables}
#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```


```{r Load dataframe}

RandomPointsCD_year_data <- sf::st_read("data_files/RandomPointsCD_year_data.shp")

```




# 1. Covariables 

Now we need to create covariables to put in the RSF models. Those will be our potential drivers for communal den locations.

We have a lot of covariables to create. Here is the list:

-   Distance to last and closest lion sighting\*

-   Distance to human community (Talek) \*

-   Distance/number (don't think we have that) of cattle\*

-   Land cover (bush vs plains, etc.)\*

-   Elevation\*

-   Distance to edge of territory\*

-   Distance to water\*


In more of a post-hoc thing: - Number of cubs at dens



## 1.1. Lions

This variable is gonna be distance to the closest lion sighting, and/or it will be the number of sightings in the territory per year.

 - Number of lions/sightings
 - Distance from the closest sighting

To do so, I will use the dataframe "tblPredatorsPerSession".

```{r filtering the lions}

DF_lions <- tblPredatorsPerSession |>
  subset(predator == "lion") |>
  select(-c(3, 4, 6:8))

DF_lions$num_individuals <- as.double(DF_lions$num_individuals)
length(unique(DF_lions$session)) #6019

#I want to sum up the number of lions for each session instead of having sessions duplicated rows
DF_lions_sum <- DF_lions |>
  group_by(session) |>
  summarise(num_lion = sum(num_individuals, na.rm = TRUE))
# suppose to have 6019 rows

```

```{r}
DF_lndmk <- tblLandmarks |>
  select(landmark_id, utme, utmn) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "Marsh_Den" & (is.na(utme) | is.na(utmn))))

```


I also need to join the session ID to the location landmark dataframe.

```{r Join Session ID to Landmark location}

DF_lions_loc_data <- tblLocationsPerSession |>
  select(c(1, 5, 6, 7)) |>
  #Here, I want to change the NAs in utme and utmn for the GPS points of the landmark
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_"))



DF_lions_loc <- merge(DF_lions_sum, DF_lions_loc_data, by = "session")

# theres a difference of 302 rows between DF_lions_loc and DF_lions_sum... 
# not sure why
```

I also need to join the tblSessions to get the clan territory and date of the specific sessions ID.

```{r Clan and year}

DF_lions_sess_data <- tblSessions |>
  select(c(1, 2, 4))

# merge the lions data with the session data to get the year and everything
DF_lions_date <- merge(DF_lions_loc, DF_lions_sess_data, by = "session")


DF_lions_clans <- DF_lions_date |>
  # keep only the clans that we are interested in
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra")) |>
  # now we extract the year 
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-c(month, day)) |>
  
  mutate(landmark_id = case_when(
    landmark_id == "Lion_L" ~ "Lion_Lugga", #Lion_L to Lion_Lugga
    landmark_id == "13_Lugga_South" ~ "Culvert_of_13_Lugga_South", #Culvert_of_13_Lugga_South for 13_Lugga_South\
    landmark_id == "Border_T" ~ "Border_Tree",   #Border_T to Border_Tree
    landmark_id == "Gomer's_D" ~ "Gomer_den",   # Gomer's_D to Gomer_den
    landmark_id == "K-6_Den" ~ "K6_Den", #K-6_Den to K6_Den
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing",  #Simba_L_X to Simba_Lugga_Crossing
    TRUE ~ landmark_id   # keep everything else unchanged
  ))
```

Now we add the GPS points for the utme and utmn columns with NAs based on the landmark ID point.

```{r Filling NAs}
# first create a dataframe with all the coordinates for the landmarks:
#that was created on line #331

# this is to put the landmark coordinates where there are NAs in the utme and utmn columns
DF_lions_data <- DF_lions_clans |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2)  |>
  drop_na(utmn) # get rid of NAs for utmn column
```


### 1.1.1. Cleaning

```{r Cleaning}
# now we change those weird coordinates for the landmark coordinates
DF_lions_data_clean <- DF_lions_data |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(
    utme = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utme_df2, utme),
    utmn = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utmn_df2, utmn)
  ) |>
  select(-utme_df2, -utmn_df2) |>
  drop_na(utmn) |>
  mutate(landmark_id = replace_na(landmark_id, "random"))

  
str(DF_lions_data_clean)

DF_lions_data_clean$utme <- as.numeric(DF_lions_data_clean$utme)
DF_lions_data_clean$utmn <- as.numeric(DF_lions_data_clean$utmn)

DF_lions_data_clean <- DF_lions_data_clean |>
  drop_na(utme)

```


I got some num of individual that are 0 :
s114919 --> change to num = 2
t61954 --> change to 1
t67315 --> change to 1
t77054 --> will put 1

remove (they all have 0 as a number of lions -- maybe we will keep them eventually ?)
s165
t101874.10
t1475.4
t53055
t57427
t60317
t62190
t75009
t77236

```{r Cleaning}

DF_lions_data_clean <- DF_lions_data_clean |>
  mutate(num_lion = ifelse(session == "s114919", 2, num_lion)) |>
  mutate(num_lion = ifelse(session == "t61954" | 
                             session == "t67315" |
                             session == "t77054", 1, num_lion)) |>
  filter(num_lion != 0)
    

```


### 1.1.2. Outliers
Change the first 7 in utmn for a 9:
"t102747" & "s109502"

Take out "t64548" because utmn has 6 numbers --> will change later.
same for s0.58

utme miss 1 number:
and t60077

utme has 1 number too many:
t50935
t72442
t98672
t115432

wrong coordinate:  (somewhere in tanzania, so probably a number is off)
s24841
```{r}

DF_lions_data_clean <- DF_lions_data_clean |>
  mutate(utmn = ifelse(session == "t102747" | session == "s109502", as.numeric(paste0("9", substring(utmn, 2))),
                  utmn))  |>
  filter(session != "t64548") |>
  filter(session != "s0.58") |>
  filter(session != "t60077") |>
  filter(session != "t50935") |>
  filter(session != "t72442") |>
  filter(session != "t98672") |>
  filter(session != "t115432") |>
  filter(session != "s24841") |>
   #removing the "." in the clan names
  mutate(clan = str_replace_all(clan, "talekw", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", ""))


  

```



### 1.1.3. Shapefiles

Saving it to shp.
```{r}

DB_lions_sf <- st_as_sf(DF_lions_data_clean, coords = c("utme", "utmn"), crs = 32736)

#Save the sf dataframe - can only do it once when it is not created before
DB_lions_sf <- sf::st_write(DB_lions_sf, "data_files/DB_lions_sf.shp", delete_dsn = TRUE)
```

Change to wgs84
```{r}
DB_lions_wgs84 <- st_transform(DB_lions_sf, crs = 4326)
#4326 is the EPSG code for the WGS84 system, which uses degrees of latitude and longitude.

# save the lat long data points into a shapefile 
DB_lions_wgs84 <- sf::st_write(DB_lions_wgs84, "data_files/DB_lions_wgs84.shp", delete_dsn = TRUE)

```


### 1.1.4. Summarize
```{r}

# number of lions in each territory for the different years
lions_data <- DB_lions_sf |>
  st_drop_geometry(lions_data) |>
  group_by(year, clan) |>
  summarize(n_lion = sum(num_lion)) |>
  arrange(year, clan)

lions_data_tab <- flextable(lions_data)
save_as_html(lions_data_tab, path = "Tables/lions_data_tab.html")


#Save
write.csv2(lions_data, "data_files/lions_data.csv", row.names = F, quote = F)


# Lions per clan over the years
lions_data_per_clan <- lions_data |>
  group_by(clan) |>
  summarize(n_lion = sum(n_lion))



lions_data_perclan_tab <- flextable(lions_data_per_clan)
save_as_html(lions_data_perclan_tab, path = "Tables/lions_data_perclan_tab.html")


#Save
write.csv2(lions_data_per_clan, "data_files/lions_data_per_clan.csv", row.names = F, quote = F)


```


### 1.1.5. Map
```{r}

# Reload our base map CD_peryear_mapleaflet
CD_peryear_mapleaflet <- readRDS("Maps/CD_peryear_mapleaflet.rds")


CD_lion_mapleaflet <- CD_peryear_mapleaflet |>
  addCircleMarkers(data = DB_lions_wgs84,
                   color = "red",
                   group = DB_lions_wgs84$year,
                   fillOpacity = 0.8,
                   radius = 1,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lion, "<br>",
                                  "Session ID:", session))

CD_lion_mapleaflet

#save
htmlwidgets::saveWidget(CD_lion_mapleaflet, file = "Maps/CD_lion_mapleaflet.html", selfcontained = TRUE)


```

### 1.1.6. Distance to closest lion sighting

I want to calculate the distance from the den location points to the closest lion sighting.

```{r}



```
















## 1.2. Distance to Talek

The first thing is to tell R to use python to be able to get the 20+ year historical satellite imagery data from Google Earth Engine, since it is the easiest way to get the Talek town borders for all those years. That API (Application Programming Interface) is only available in Python or JavaScript. So I have to tell R to use Python to have access to that. --\> An API is like a menu at a restaurant - it lists the things you can ask for (functions, data) and tells the kitchen (the software) how to prepare them for you. --\> Google Earth Engine API = the set of commands you can send to Google's satellite image servers to request data (e.g., "give me all Landsat images for Talek Town in 2005").

```{r}
# 1. Location of Talek town, Maasai Mara
talek_point <- ee$Geometry$Point(35.216423, -1.442912) #talek town center
talek_roi <- talek_point$buffer(600) # 5 km radius
```

Here is a test for year 2020:

```{r}
# 2 - Function for built-up detection
get_yearly_builtup <- function(year) {
  s2 <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(talek_roi)$
    filterDate(sprintf("%d-01-01", year), sprintf("%d-12-31", year))$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
    median()
  
  # Calculate NDBI (built-up) and NDVI (vegetation)
  ndbi <- s2$normalizedDifference(c("B11", "B8"))
  ndvi <- s2$normalizedDifference(c("B8", "B4"))
  
  # Threshold: built-up AND low vegetation
  builtup_mask <- ndbi$gt(0.2)$And(ndvi$lt(0.3))
  
  # Mask original image
  builtup <- builtup_mask$selfMask()
  
  # Reduce to vectors inside ROI
  vectors <- builtup$reduceToVectors(
    geometry = talek_roi,
    geometryType = "polygon",
    reducer = ee$Reducer$countEvery(),
    scale = 10,
    maxPixels = 1e10
  )
  
  vectors$set("year", year)
}

# --- Year of interest ---
year <- 2020

# --- Generate polygons for that year ---
builtup_polygons <- get_yearly_builtup(year)

# --- Visualization parameters ---
Map$centerObject(talek_roi, zoom = 15)
Map$addLayer(talek_roi, list(color = "blue"), "Talek ROI") +
  Map$addLayer(builtup_polygons, list(color = "red"), paste("Built-up", year))


```

```{r}
# 2. Function to mask clouds and detect buildings
mask_builtup <- function(image) {
  # Calculate NDBI
  swir <- image$select("B11")
  nir <- image$select("B8")
  ndbi <- swir$subtract(nir)$divide(swir$add(nir))
  
  # Built-up mask (adjust threshold if needed)
  builtup_mask <- ndbi$gt(0.2) # try lowering to 0.1 if empty
  
  # Apply mask but keep it as a single-band image for vectorizing
  return(builtup_mask$rename("builtup"))
}

# 3. Function to get yearly composite and vector polygons
get_yearly_builtup <- function(year) {
  start <- sprintf("%d-01-01", year)
  end <- sprintf("%d-12-31", year)
  
  yearly_img <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(region)$
    filterDate(start, end)$
    map(mask_builtup)$
    max() # union of all built-up pixels in the year
  
  # Ensure projection
  yearly_img <- yearly_img$setDefaultProjection("EPSG:4326", NULL, 10)
  
  # Vectorize built-up areas
  vectors <- yearly_img$reduceToVectors(
    geometry = region,
    geometryType = "polygon",
    scale = 10,
    crs = "EPSG:4326",
    maxPixels = 1e10
  )
  
  vectors_sf <- ee_as_sf(vectors)
  vectors_sf$year <- year
  return(vectors_sf)
}

# 4. Run for a range of years
years <- 2021:2023
builtup_polygons <- do.call(rbind, lapply(years, get_yearly_builtup))

# 5. Save shapefile
st_write(builtup_polygons, "talek_builtup_yearly.shp", delete_dsn = TRUE)
```

### 1.2.1. Calculation

```{r}

talek_center <- data.frame(
  lon = 35.216423,
  lat = -1.442912
)

#Convert to sf object
talek_sf <- st_as_sf(talek_center, coords = c("lon", "lat"), crs = 4326)

# Transform to a projected CRS in meters (UTM zone for this area: 36S)
RandomPointsCD_year_data_talek <- st_transform(RandomPointsCD_year_data , 32736)    # EPSG 32736 = WGS 84 / UTM zone 36S
talek_sf_m <- st_transform(talek_sf, 32736)


# Calculate distances in meters
RandomPointsCD_year_data$dst_tl <- as.numeric(st_distance(RandomPointsCD_year_data_talek, talek_sf_m))


```



## 1.3. Land cover (LULC)

Now we will use the landcover raster from Spagnuolo et al. 2018. It has way more precise categories of land cover by using multi-date radar and optical imagery in a highly accurate way. More importantly, it was validated with field observations.

```{r}
# Load your raster land cover from LULC (ZL file)
Mara_LULC <- terra::rast("Mara_LULC/Mara_LULC_2018.tif")

# Check it
Mara_LULC

plot(Mara_LULC, col = terrain.colors(20))
```


### 1.3.1. Extract land cover

We want to extract the land cover for my specific CD and ramdom points created.
To do this we will use the raster directly, so first we need to transfer my dataframe of Random/used points into a vector (terra format) and then extract the land cover respective to each location.

```{r}
# Convert sf → SpatVector (terra format)
RandomPointsCD_vect <- vect(RandomPointsCD_year_data)

# Extract land cover values at each point
RandomPointsCD_year_data$landcover <- terra::extract(Mara_LULC, RandomPointsCD_vect)[,2]
```

Now we change the code to what it is:

The type_id translates to land cover class as follows (classes defined in Table 1 of the attached paper):

type_id 2 = Barren
type_id 5 = Riverine forest
type_id 6 = Upland forest
type_id 7 = Grass Acacia
type_id 8 = Grass Balanites
type_id 10 = Tall grass
type_id 11 = Medium grass
type_id 13 = Short grass
type_id 14 = Shrub
type_id 17 = Water
type_id 18 = Emergent wetland
type_id 20 = Wet meadow


```{r}
RandomPointsCD_data <- RandomPointsCD_year_data |>
  mutate(landcover = dplyr::recode(landcover,
    `2` = "Barren",
    `5` = "Riverine_forest",
    `6` = "Upland_forest",
    `7` = "Grass_Acaciad",
    `8` = "Grass_Balanites",
    `10` = "Tall_grass",
    `11` = "Medium_grass",
    `13` = "Short_grass",
    `14` = "Shrub",
    `17` = "Water", 
    `18` = "Emergent_wetland", 
    `20` = "Wet_meadow"))
```

### 1.3.2. Map

```{r}
# Reproject to WGS84 (Leaflet needs EPSG:4326)
Mara_LULC_wgs84 <- project(Mara_LULC, "EPSG:4326")

Mara_LULC_wgs84_small <- terra::aggregate(Mara_LULC_wgs84, fact = 20)


```



```{r}

# Create color palette:
pal_LULC <- colorNumeric(palette = terrain.colors(20), 
                    domain = values(Mara_LULC_wgs84), 
                    na.color = "transparent")


Mara_LULC_CD_leaflet <- leaflet() |>
  addTiles() |>
  addRasterImage(Mara_LULC_wgs84_small, colors = pal_LULC, opacity = 0.6) |>
  addCircleMarkers(data = CD_landcover,
                   radius = 2, stroke = FALSE, fillOpacity = 0.9,
                   popup = ~paste("Clan:", clan, "<br>", 
                                  "LC:", landcover, "<br>",
                                  "Landmark:", lndmrk_))  |>
  addLegend(pal = pal_LULC, values = values(Mara_LULC_wgs84_small),
            title = "Landcover", position = "bottomright")

Mara_LULC_CD_leaflet

```

```{r}
#save
htmlwidgets::saveWidget(Mara_LULC_CD_leaflet, file = "Maps/Mara_LULC_CD_leaflet.html", selfcontained = TRUE)
```


```{r}

Mara_LULC_CDRandom_leaflet <- leaflet() |>
  addTiles() |>
  addRasterImage(Mara_LULC_wgs84_small, colors = pal_LULC, opacity = 0.6) |>
  addCircleMarkers(data = RandomPointsCD_data,
                   color = ~ifelse(case=="1", "red", "blue"),
                   radius = 2, stroke = FALSE, fillOpacity = 0.9,
                   popup = ~paste("Clan:", clan, "<br>", 
                                  "LC:", landcover, "<br>",
                                  "Landmark:", lndmrk_))  |>
  addLegend(pal = pal_LULC, values = values(Mara_LULC_wgs84_small),
            title = "Landcover", position = "bottomright")

Mara_LULC_CDRandom_leaflet
```

```{r}
#save
htmlwidgets::saveWidget(Mara_LULC_CDRandom_leaflet, file = "Maps/Mara_LULC_CDRandom_leaflet.html", selfcontained = TRUE)
```




## 1.4. Distance to water

To get the distance to the closest water source, here are the steps we need to do :

1. Identify water sources in the Masai Mara (type 17 of LULC is water.
2. Convert that into a raster mask (water = 1, non-water = 0).
3. Compute distance to nearest water pixel.
4. Sample that distance at your points.







## 1.5. Distance from edge of the territory


```{r}

```





## 1.6. Distance to roads

```{r}

```
















