---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Creating the covariables 

# 0. Libraries et al.

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

```

```{r Load MHP tables}
#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```


```{r Load dataframe}

RandomPointsCD_year_data <- sf::st_read("data_files/RandomPointsCD_year_data.shp")

```

# 1. Installing rgee

I will use rgee package to get different spatial covariables : 

-   Land cover (bush vs plains, etc.)\*

-   Altitude\*

-   Distance to edge of territory\*

-   Distance to water\*

-   NDVI


rgee is a package that is use to call Google Earth Engine API in R. Google Earth Engine is a cloud-based platform designed for planetary-scale environmental data analysis, offering access to a vast archive of remote sensing data and the ability to perform geospatial analysis on Google's infrastructure. While Google Earth Engine officially supports Python and JavaScript client libraries, rgee bridges this gap for R users by utilizing the reticulate package to wrap the Earth Engine Python API. This allows R users to interact with GEE and leverage its capabilities using a familiar R interface, facilitating rapid development and integration with other open-source R packages for spatial data analysis.


Setting up a google earth engine project first.

The first time I need to install conda, create the rgee environment within the conda thing. 

I followed the steps here to have it finally worked: https://gis.stackexchange.com/questions/473730/authentification-error-when-installing-rgee-in-rstudio?newreg=3674f3f594ee4f91856aa2e5df5b204d


```{r}
#to get rgeeExtra, to use things like the gif animation
#remotes::install_github("r-earthengine/rgeeExtra")

```

Here are the lines that needed to be run the first time, but now I don't need them anymore. Might need to rerun them when ee_Initialize doesn't work.
```{r }
#library(reticulate)

#install.packages("rgee")
#reticulate::conda_install(envname = "rgee", packages = "earthengine-api=0.1.370", channel = "conda-forge", python_version = "3.9")
```

```{r}
library(reticulate)

#This is to tell R to use the right Python environment
reticulate::use_condaenv("rgee", required = TRUE)

library(rgee)
library(rgeeExtra)

```


Initialize rgee
```{r}
#Initialize EE 
ee_Initialize()
#important to initialize earth engine without the user and project argument here
```


This will need to be run when ee_Initialize is not working.
```{r}
#To get credentials: (need to run it once in a while when ee_Initialize doesn't work)
ee_Authenticate()

ee_clean_user_credentials()
#toforce it:
ee_Authenticate(auth_mode = "notebook")

```

## 1.1. Checking 
The ee_check() function will help for checking the sanity of the rgee installation and dependencies.

- ee_check_python() - Python version
- ee_check_credentials() - Google Drive and GCS credentials
- ee_check_python_packages() - Python packages

```{r}
ee_check()

ee_check_python()
ee_check_credentials()
ee_check_python_packages()
```



## 1.2. Tests
```{r}

# test to make sure all is running smoothly:
ee$Image("CGIAR/SRTM90_V4")$getInfo()
ee$String('Hello from the Earth Engine servers!')$getInfo()

```


### i) Night time lights
From : https://github.com/r-spatial/rgee
To Compute the trend of night-time lights

Let's create a new band containing the image date as years since 1991 by extracting the year of the image acquisition date and subtracting it from 1991.
```{r}

createTimeBand <-function(img) {
  year <- ee$Date(img$get('system:time_start'))$get('year')$subtract(1991L) 
  ee$Image(year)$byte()$addBands(img)
}

#Use your TimeBand function to map it over the night-time lights collection.

collection <- ee$
  ImageCollection('NOAA/DMSP-OLS/NIGHTTIME_LIGHTS')$
  select('stable_lights')$
  map(createTimeBand)

# Compute a linear fit over the series of values at each pixel, so that you can visualize the y-intercept as green, and the positive/negative slopes as red/blue.
col_reduce <- collection$reduce(ee$Reducer$linearFit())
col_reduce <- col_reduce$addBands(
  col_reduce$select('scale'))
ee_print(col_reduce)

# to visualize the map: 
Map$setCenter(9.08203, 47.39835, 3)
Map$addLayer(
  eeObject = col_reduce,
  visParams = list(
    bands = c("scale", "offset", "scale"),
    min = 0,
    max = c(0.18, 20, -0.18)
  ),
  name = "stable lights trend"
)

```


### ii) Worldwide elevation map
From: https://cran.r-project.org/web/packages/rgee/vignettes/rgee01.html

SRTM (Shuttle Radar Topography Mission) elevation values worldwide:
```{r}

#create a ee object
srtm <- ee$Image("USGS/SRTMGL1_003")

#Define visualization parameters
viz <- list(
  max = 4000,
  min = 0,
  palette = c("#000000","#5AAD5A","#A9AD84","#FFFFFF"))

# using Map$addLayer to visualize the map interactively
Map$addLayer(
  eeObject = srtm,
  visParams =  viz,
  name = 'SRTM')

```


### iii) Precipitation
From: https://github.com/r-spatial/rgee

To play with some precipitation data: 

nc = North Carolina counties, 100 simple features with 14 attributes
```{r}
# to read the nc shapefile:
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

plot(nc)

# will use the Terraclimate database
terraclimate <- ee$ImageCollection("IDAHO_EPSCOR/TERRACLIMATE") %>%
  ee$ImageCollection$filterDate("2001-01-01", "2002-01-01") %>%
  ee$ImageCollection$map(function(x) x$select("pr")) %>% # Select only precipitation bands
  ee$ImageCollection$toBands() %>% # from imagecollection to image
  ee$Image$rename(sprintf("PP_%02d",1:12)) # rename the bands of an image

# to extract monthly precipitation:
ee_nc_rain <- ee_extract(x = terraclimate, y = nc["NAME"], sf = FALSE)

```

ee_extract will help you to extract monthly precipitation values from the Terraclimate ImageCollection. ee_extract works similar to raster::extract, you just need to define: the ImageCollection object (x), the geometry (y), and a function to summarize the values (fun).

```{r}
# then we visualize it using ggplot:
ee_nc_rain %>%
  pivot_longer(-NAME, names_to = "month", values_to = "pr") %>%
  mutate(month, month=gsub("PP_", "", month)) %>%
  ggplot(aes(x = month, y = pr, group = NAME, color = pr)) +
  geom_line(alpha = 0.4) +
  xlab("Month") +
  ylab("Precipitation (mm)") +
  theme_minimal()
```

### iv) NDVI animation

From: https://github.com/r-spatial/rgee

To create a NDVI animation map (Peru's Arequepia region)
 
```{r}

#Define the regional bounds of animation frames and a mask to clip the NDVI data by.
mask <- system.file("shp/arequipa.shp", package = "rgee") %>%
  st_read(quiet = TRUE) %>%
  sf_as_ee()
region <- mask$geometry()$bounds()


#Retrieve the MODIS Terra Vegetation Indices 16-Day Global 1km dataset as an ee.ImageCollection and then, select the NDVI band.

col <- ee$ImageCollection('MODIS/006/MOD13A2')$select('NDVI')

#Group images by composite date
col <- col$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('day', 'year')
  img$set('doy', doy)
})
distinctDOY <- col$filterDate('2013-01-01', '2014-01-01')

```

Now, let's define a filter that identifies which images from the complete collection match the DOY from the distinct DOY collection.
```{r}
filter <- ee$Filter$equals(leftField = 'doy', rightField = 'doy')

```
Define a join and convert the resulting FeatureCollection to an ImageCollection... it will take you only 2 lines of code!

```{r}
join <- ee$Join$saveAll('doy_matches')
joinCol <- ee$ImageCollection(join$apply(distinctDOY, col, filter))
```

Apply median reduction among the matching DOY collections.

```{r}
comp <- joinCol$map(function(img) {
  doyCol = ee$ImageCollection$fromImages(
    img$get('doy_matches')
  )
  doyCol$reduce(ee$Reducer$median())
})

#define RGB visualization parameters first.
visParams = list(
  min = 0.0,
  max = 9000.0,
  bands = "NDVI_median",
  palette = c(
    'FFFFFF', 'CE7E45', 'DF923D', 'F1B555', 'FCD163', '99B718', '74A901',
    '66A000', '529400', '3E8601', '207401', '056201', '004C00', '023B01',
    '012E01', '011D01', '011301'
  )
)

#Create RGB visualization images for use as animation frames.
rgbVis <- comp$map(function(img) {
  do.call(img$visualize, visParams) %>%
    ee$Image$clip(mask)
})


#GIF visualization parameters.
gifParams <- list(
  region = region,
  dimensions = 600,
  crs = 'EPSG:3857',
  framesPerSecond = 10
)

# get month names: 
dates_modis_mabbr <- distinctDOY %>%
  ee_get_date_ic %>% # Get Image Collection dates
  '[['("time_start") %>% # Select time_start column
  lubridate::month() %>% # Get the month component of the datetime
  '['(month.abb, .) # subset around month abbreviations
```

Finally, we use ee_utils_gif_* functions to render the GIF animation and add some texts.
```{r}

animation <- ee_utils_gif_creator(rgbVis, gifParams, mode = "wb")
animation %>%
  ee_utils_gif_annotate(
    text = "NDVI: MODIS/006/MOD13A2",
    size = 15, color = "white",
    location = "+10+10"
  ) %>%
  ee_utils_gif_annotate(
    text = dates_modis_mabbr,
    size = 30,
    location = "+290+350",
    color = "white",
    font = "arial",
    boxcolor = "#000000"
  ) # -> animation_wtxt

#to save it
# ee_utils_gif_save(animation_wtxt, path = "raster_as_ee.gif")

```


# 2. Covariables 

Now we need to create covariables to put in the RSF models. Those will be our potential drivers for communal den locations.

We have a lot of covariables to create. Here is the list:

-   Distance to last and closest lion sighting\*

-   Distance to human community (Talek) \*

-   Distance/number (don't think we have that) of cattle\*

-   Land cover (bush vs plains, etc.)\*

-   Altitude\*

-   Distance to edge of territory\*

-   Distance to water\*

- Time of year ? --> I really want to find a way to get this and have a 


In more of a post-hoc thing: - Number of cubs at dens



## 1.1. Lions

This variable is gonna be distance to the closest lion sighting, and/or it will be the number of sightings in the territory per year.

 - Number of lions/sightings
 - Distance from the closest sighting

To do so, I will use the dataframe "tblPredatorsPerSession".

```{r filtering the lions}

DF_lions <- tblPredatorsPerSession |>
  subset(predator == "lion") |>
  select(-c(3, 4, 6:8))

DF_lions$num_individuals <- as.double(DF_lions$num_individuals)
length(unique(DF_lions$session)) #6019

#I want to sum up the number of lions for each session instead of having sessions duplicated rows
DF_lions_sum <- DF_lions |>
  group_by(session) |>
  summarise(num_lion = sum(num_individuals, na.rm = TRUE))
# suppose to have 6019 rows

```

```{r}
DF_lndmk <- tblLandmarks |>
  select(landmark_id, utme, utmn) |>
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_")) |>
  filter(!(landmark_id == "Marsh_Den" & (is.na(utme) | is.na(utmn))))

```


I also need to join the session ID to the location landmark dataframe.

```{r Join Session ID to Landmark location}

DF_lions_loc_data <- tblLocationsPerSession |>
  select(c(1, 5, 6, 7)) |>
  #Here, I want to change the NAs in utme and utmn for the GPS points of the landmark
  mutate(landmark_id = str_replace_all(landmark_id, " ", "_"))



DF_lions_loc <- merge(DF_lions_sum, DF_lions_loc_data, by = "session")

# theres a difference of 302 rows between DF_lions_loc and DF_lions_sum... 
# not sure why
```

I also need to join the tblSessions to get the clan territory and date of the specific sessions ID.

```{r Clan and year}

DF_lions_sess_data <- tblSessions |>
  select(c(1, 2, 4))

# merge the lions data with the session data to get the year and everything
DF_lions_date <- merge(DF_lions_loc, DF_lions_sess_data, by = "session")


DF_lions_clans <- DF_lions_date |>
  # keep only the clans that we are interested in
  subset(str_detect(clan, "talek") | str_detect(clan, "serena") | str_detect(clan, "happy.zebra")) |>
  # now we extract the year 
  separate(date, c("year", "month", "day"), sep = "-", remove = F) |>
  select(-c(month, day)) |>
  
  mutate(landmark_id = case_when(
    landmark_id == "Lion_L" ~ "Lion_Lugga", #Lion_L to Lion_Lugga
    landmark_id == "13_Lugga_South" ~ "Culvert_of_13_Lugga_South", #Culvert_of_13_Lugga_South for 13_Lugga_South\
    landmark_id == "Border_T" ~ "Border_Tree",   #Border_T to Border_Tree
    landmark_id == "Gomer's_D" ~ "Gomer_den",   # Gomer's_D to Gomer_den
    landmark_id == "K-6_Den" ~ "K6_Den", #K-6_Den to K6_Den
    landmark_id == "Simba_L_X" ~ "Simba_Lugga_Crossing",  #Simba_L_X to Simba_Lugga_Crossing
    TRUE ~ landmark_id   # keep everything else unchanged
  ))
```

Now we add the GPS points for the utme and utmn columns with NAs based on the landmark ID point.

```{r Filling NAs}
# first create a dataframe with all the coordinates for the landmarks:
#that was created on line #331

# this is to put the landmark coordinates where there are NAs in the utme and utmn columns
DF_lions_data <- DF_lions_clans |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(utme = ifelse(is.na(utme), utme_df2, utme), 
         utmn = ifelse(is.na(utmn), utmn_df2, utmn)) |>
  select(-utme_df2, -utmn_df2)  |>
  drop_na(utmn) # get rid of NAs for utmn column
```


### 1.1.1. Cleaning

```{r Cleaning}
# now we change those weird coordinates for the landmark coordinates
DF_lions_data_clean <- DF_lions_data |>
  left_join(DF_lndmk, by = "landmark_id", suffix = c("", "_df2"), relationship = "many-to-many") |>
  mutate(
    utme = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utme_df2, utme),
    utmn = if_else(session == "s13727" | 
                     session == "s22671" |
                     session == "s22901" |
                     session == "s6586" |
                     session == "t105678" |
                     session == "t88078" |
                     session == "s10853", utmn_df2, utmn)
  ) |>
  select(-utme_df2, -utmn_df2) |>
  drop_na(utmn) |>
  mutate(landmark_id = replace_na(landmark_id, "random"))

  
str(DF_lions_data_clean)

DF_lions_data_clean$utme <- as.numeric(DF_lions_data_clean$utme)
DF_lions_data_clean$utmn <- as.numeric(DF_lions_data_clean$utmn)

DF_lions_data_clean <- DF_lions_data_clean |>
  drop_na(utme)

```


I got some num of individual that are 0 :
s114919 --> change to num = 2
t61954 --> change to 1
t67315 --> change to 1
t77054 --> will put 1

remove (they all have 0 as a number of lions -- maybe we will keep them eventually ?)
s165
t101874.10
t1475.4
t53055
t57427
t60317
t62190
t75009
t77236

```{r Cleaning}

DF_lions_data_clean <- DF_lions_data_clean |>
  mutate(num_lion = ifelse(session == "s114919", 2, num_lion)) |>
  mutate(num_lion = ifelse(session == "t61954" | 
                             session == "t67315" |
                             session == "t77054", 1, num_lion)) |>
  filter(num_lion != 0)
    

```


### 1.1.2. Outliers
Change the first 7 in utmn for a 9:
"t102747" & "s109502"

Take out "t64548" because utmn has 6 numbers --> will change later.
same for s0.58

utme miss 1 number:
and t60077

utme has 1 number too many:
t50935
t72442
t98672
t115432

wrong coordinate:  (somewhere in tanzania, so probably a number is off)
s24841
```{r}

DF_lions_data_clean <- DF_lions_data_clean |>
  mutate(utmn = ifelse(session == "t102747" | session == "s109502", as.numeric(paste0("9", substring(utmn, 2))),
                  utmn))  |>
  filter(session != "t64548") |>
  filter(session != "s0.58") |>
  filter(session != "t60077") |>
  filter(session != "t50935") |>
  filter(session != "t72442") |>
  filter(session != "t98672") |>
  filter(session != "t115432") |>
  filter(session != "s24841") |>
   #removing the "." in the clan names
  mutate(clan = str_replace_all(clan, "talekw", "talek")) |>
  mutate(clan = str_replace_all(clan, "\\.", ""))


  

```



### 1.1.3. Shapefiles

Saving it to shp.
```{r}

DB_lions_sf <- st_as_sf(DF_lions_data_clean, coords = c("utme", "utmn"), crs = 32736)

#Save the sf dataframe - can only do it once when it is not created before
DB_lions_sf <- sf::st_write(DB_lions_sf, "data_files/DB_lions_sf.shp", delete_dsn = TRUE)
```

Change to wgs84
```{r}
DB_lions_wgs84 <- st_transform(DB_lions_sf, crs = 4326)
#4326 is the EPSG code for the WGS84 system, which uses degrees of latitude and longitude.

# save the lat long data points into a shapefile 
DB_lions_wgs84 <- sf::st_write(DB_lions_wgs84, "data_files/DB_lions_wgs84.shp", delete_dsn = TRUE)

```


### 1.1.4. Summarize
```{r}

# number of lions in each territory for the different years
lions_data <- DB_lions_sf |>
  st_drop_geometry(lions_data) |>
  group_by(year, clan) |>
  summarize(n_lion = sum(num_lion)) |>
  arrange(year, clan)

lions_data_tab <- flextable(lions_data)
save_as_html(lions_data_tab, path = "Tables/lions_data_tab.html")


#Save
write.csv2(lions_data, "data_files/lions_data.csv", row.names = F, quote = F)


# Lions per clan over the years
lions_data_per_clan <- lions_data |>
  group_by(clan) |>
  summarize(n_lion = sum(n_lion))



lions_data_perclan_tab <- flextable(lions_data_per_clan)
save_as_html(lions_data_perclan_tab, path = "Tables/lions_data_perclan_tab.html")


#Save
write.csv2(lions_data_per_clan, "data_files/lions_data_per_clan.csv", row.names = F, quote = F)


```


### 1.1.5. Map
```{r}

# Reload our base map CD_peryear_mapleaflet
CD_peryear_mapleaflet <- readRDS("Maps/CD_peryear_mapleaflet.rds")


CD_lion_mapleaflet <- CD_peryear_mapleaflet |>
  addCircleMarkers(data = DB_lions_wgs84,
                   color = "red",
                   group = DB_lions_wgs84$year,
                   fillOpacity = 0.8,
                   radius = 1,
                   popup = ~paste("Year:", year, "<br>", 
                                  "Number of lions:", num_lion, "<br>",
                                  "Session ID:", session))

CD_lion_mapleaflet

#save
htmlwidgets::saveWidget(CD_lion_mapleaflet, file = "Maps/CD_lion_mapleaflet.html", selfcontained = TRUE)


```

### 1.1.6. Calculation of distance

I want to calculate the distance from the den location points to the closest lion sighting.

```{r}



```
















## 1.2. Distance to Talek

The first thing is to tell R to use python to be able to get the 20+ year historical satellite imagery data from Google Earth Engine, since it is the easiest way to get the Talek town borders for all those years. That API (Application Programming Interface) is only available in Python or JavaScript. So I have to tell R to use Python to have access to that. --\> An API is like a menu at a restaurant - it lists the things you can ask for (functions, data) and tells the kitchen (the software) how to prepare them for you. --\> Google Earth Engine API = the set of commands you can send to Google's satellite image servers to request data (e.g., "give me all Landsat images for Talek Town in 2005").

```{r}
# 1. Location of Talek town, Maasai Mara
talek_point <- ee$Geometry$Point(35.216423, -1.442912) #talek town center
talek_roi <- talek_point$buffer(600) # 5 km radius
```

Here is a test for year 2020:

```{r}
# 2 - Function for built-up detection
get_yearly_builtup <- function(year) {
  s2 <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(talek_roi)$
    filterDate(sprintf("%d-01-01", year), sprintf("%d-12-31", year))$
    filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 20))$
    median()
  
  # Calculate NDBI (built-up) and NDVI (vegetation)
  ndbi <- s2$normalizedDifference(c("B11", "B8"))
  ndvi <- s2$normalizedDifference(c("B8", "B4"))
  
  # Threshold: built-up AND low vegetation
  builtup_mask <- ndbi$gt(0.2)$And(ndvi$lt(0.3))
  
  # Mask original image
  builtup <- builtup_mask$selfMask()
  
  # Reduce to vectors inside ROI
  vectors <- builtup$reduceToVectors(
    geometry = talek_roi,
    geometryType = "polygon",
    reducer = ee$Reducer$countEvery(),
    scale = 10,
    maxPixels = 1e10
  )
  
  vectors$set("year", year)
}

# --- Year of interest ---
year <- 2020

# --- Generate polygons for that year ---
builtup_polygons <- get_yearly_builtup(year)

# --- Visualization parameters ---
Map$centerObject(talek_roi, zoom = 15)
Map$addLayer(talek_roi, list(color = "blue"), "Talek ROI") +
  Map$addLayer(builtup_polygons, list(color = "red"), paste("Built-up", year))


```

```{r}
# 2. Function to mask clouds and detect buildings
mask_builtup <- function(image) {
  # Calculate NDBI
  swir <- image$select("B11")
  nir <- image$select("B8")
  ndbi <- swir$subtract(nir)$divide(swir$add(nir))
  
  # Built-up mask (adjust threshold if needed)
  builtup_mask <- ndbi$gt(0.2) # try lowering to 0.1 if empty
  
  # Apply mask but keep it as a single-band image for vectorizing
  return(builtup_mask$rename("builtup"))
}

# 3. Function to get yearly composite and vector polygons
get_yearly_builtup <- function(year) {
  start <- sprintf("%d-01-01", year)
  end <- sprintf("%d-12-31", year)
  
  yearly_img <- ee$ImageCollection("COPERNICUS/S2_SR")$
    filterBounds(region)$
    filterDate(start, end)$
    map(mask_builtup)$
    max() # union of all built-up pixels in the year
  
  # Ensure projection
  yearly_img <- yearly_img$setDefaultProjection("EPSG:4326", NULL, 10)
  
  # Vectorize built-up areas
  vectors <- yearly_img$reduceToVectors(
    geometry = region,
    geometryType = "polygon",
    scale = 10,
    crs = "EPSG:4326",
    maxPixels = 1e10
  )
  
  vectors_sf <- ee_as_sf(vectors)
  vectors_sf$year <- year
  return(vectors_sf)
}

# 4. Run for a range of years
years <- 2021:2023
builtup_polygons <- do.call(rbind, lapply(years, get_yearly_builtup))

# 5. Save shapefile
st_write(builtup_polygons, "talek_builtup_yearly.shp", delete_dsn = TRUE)
```

### 1.2.1. Calculation

```{r}

talek_center <- data.frame(
  lon = 35.216423,
  lat = -1.442912
)

#Convert to sf object
talek_sf <- st_as_sf(talek_center, coords = c("lon", "lat"), crs = 4326)

# Transform to a projected CRS in meters (UTM zone for this area: 36S)
RandomPointsCD_yearly_data_talek <- st_transform(RandomPointsCD_yearly_data, 32736)    # EPSG 32736 = WGS 84 / UTM zone 36S
talek_sf_m <- st_transform(talek_sf, 32736)


# Calculate distances in meters
RandomPointsCD_yearly_data$dst_tl <- as.numeric(st_distance(RandomPointsCD_yearly_data_talek, talek_sf_m))


```





## 1.3. Land cover


### 1.3.1. Masai Mara geometry

I will use the package rgee (Google Earth Engine for R), which was installed in section 1.

First, we will get the Masai MAra polygon, which will be use to extract the land cover for our used and available points in `RandomPointsCD_yearly_data`.
Here, we’ll query the WDPA protected-areas layer in GEE and filter for "Masai Mara" to avoid manually drawing or copying coordinates.
```{r Masai Mara}

# WDPA featurecollection (server-side)
wdpa <- ee$FeatureCollection("WCMC/WDPA/current/polygons")

# filter for features whose NAME contains "Masai Mara"
mara_fc <- wdpa$filter(ee$Filter$stringContains('NAME', 'Masai Mara'))

mara_geom <- mara_fc$geometry()   # keep geometry in EE

#Quick map to see if we got the right place:
Map$centerObject(mara_geom, zoom = 10)
Map
```



### 1.3.2. Extract land cover

Now we extract the land cover. We will use the ESA WorldCover dataset, which is a 10m global land cover with 11 classes. It is supposedely great for small-scale features like our den locations. This is going to be for 2020 landcover.

See: 
- https://developers.google.com/earth-engine/datasets/catalog/ESA_WorldCover_v100?hl=fr
- https://esa-worldcover.org/en
- https://worldcover2020.esa.int/viewer
- https://www.arcgis.com/home/item.html?id=e28b7e1da5414010ba4f47dd5a3c3ebb (for the classes code)


```{r}
# Load land cover (WorldCover 10m)
WorldCover <- ee$ImageCollection("ESA/WorldCover/v100")$first()$select("Map")

# Clip to Mara (still server-side)
MasaiMara_landcover_WC <- WorldCover$clip(mara_geom)

# quick map to see (rgee interactive map):
Map$addLayer(MasaiMara_landcover_WC, list(bands = "Map"), "WorldCover 10m (Map)")
Map

```


Now we sample the landcover at our used and random points :
```{r}
# Sample at your points (points_sf is your sf dataframe of den + available points)
RandomPointsCD_year_data_ee <- sf_as_ee(RandomPointsCD_year_data)

RandomPointsCD_data_sampled <- MasaiMara_landcover_WC$sampleRegions(
  collection = RandomPointsCD_year_data_ee,
  scale = 10,
  geometries = TRUE
)

# Bring only the sampled results to R (much smaller)
RandomPointsCD_landcover <- ee_as_sf(RandomPointsCD_data_sampled)


```

```{r}

RandomPointsCD_landcover <- RandomPointsCD_landcover |>
  mutate(landcover = dplyr::recode(Map,
    `10` = "Tree",
    `20` = "Shrub",
    `30` = "Grass",
    `40` = "Cropland",
    `50` = "Built-up",
    `60` = "Bare",
    `80` = "Water",
    `90` = "Wetland"))

```


### 1.3.3. Map

```{r}
# Color palette for land cover
pal_lc <- colorFactor(palette = "Set2",  # or "viridis", or custom c("green","yellow",...)
                      domain = RandomPointsCD_landcover$landcover)

# keeping only the used points
CD_landcover <- RandomPointsCD_landcover |>
  filter(case == 1)


# make a base map
CD_landcover_mapleaflet <- leaflet(CD_landcover) |>
  addTiles()

# loop over clans and add each as a separate group
for (cl in unique(CD_landcover$clan)) {
  CD_landcover_mapleaflet <- CD_landcover_mapleaflet |>
    addCircleMarkers(
      data = filter(CD_landcover, clan == cl),
      color = ~pal_lc(landcover),
      fillOpacity = 0.8,
      group = cl,
      radius = 3,
      stroke = FALSE,
      popup = ~paste("Clan:", clan, "<br>", 
                     "LC:", landcover, "<br>",
                     "Landmark:", lndmrk_))
}

# add toggle control
CD_landcover_mapleaflet <- CD_landcover_mapleaflet |>
  addLayersControl(
    overlayGroups = unique(CD_landcover$clan),
    options = layersControlOptions(collapsed = FALSE)) |>
  addLegend("bottomright", pal = pal_lc, values = ~landcover,
            title = "Land cover", opacity = 1)

CD_landcover_mapleaflet

```

```{r Save}

#save
htmlwidgets::saveWidget(CD_landcover_mapleaflet, file = "Maps/CD_landcover_mapleaflet.html", selfcontained = TRUE)


```


### 1.3.4. Table

```{r}

landcover_table_sum <- CD_landcover |>
  st_drop_geometry() |>
  group_by(landcover) |>
  summarise(n_used = n()) |>
  arrange(desc(n_used))

landcover_table_sum

landcover_tab_sum <- flextable(landcover_table_sum)
save_as_html(landcover_tab_sum, path = "Tables/landcover_tab_sum.html")


```




## 1.4. Altitude
```{r}

```






































