---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Den Locations - Fiona's script

## 0. Libraries

```{r}

install.packages("terra")

library(readr)
library(dplyr)
library(ggplot2)
library(sf)
library(janitor)
library(leaflet)
library(terra) # to upload raster data
library(tidyterra)
```

## 1. Import Data

data_all \<- list.files(path = "E:\\MHP\\Data\\MHP\\ActiveCD_data", pattern = "\*.csv", full.names = TRUE) colnames(den_data)

df \<- readr::read_csv2("E:\\MHP\\Data\\MHP\\ActiveCD_data.csv")

head(df)

```{r}


den_data <- read_delim("data_files/ActiveCD_data.csv")

# Inspect structure
str(den_data)
den_data <- filter(den_data,utmn<9859000)

den_data <- den_data %>%
  mutate(clan = case_when(
    clan %in% c("talek,", "talek") ~ "talek",
    TRUE ~ clan
  ))
den_data <- den_data %>%
  janitor::clean_names() %>%    # standardizes names (lowercase, no spaces)
  mutate(
    utme  = as.numeric(gsub(",", "", utme)),
    utmn  = as.numeric(gsub(",", "", utmn)),
    year  = as.integer(year),
    month = as.factor(month),
    clan  = as.factor(clan) 
  )

den_summary <- den_data %>%
  group_by(landmark_id, utme, utmn, clan) %>%
  summarise(total_sessions = n(), .groups = "drop")
```

## 2. Map


Uploading landcover grassland raster.
```{r}
Mara_LULC <- terra::rast("Mara_LULC/Mara_LULC_2018.tif")
```


```{r}

Mara_LULC_df <- as.data.frame(Mara_LULC, xy = TRUE)

Mara_LULC_df <- Mara_LULC_df |>
  rename(landcover = Mara_LULC_2018)


landcover_map <- ggplot(Mara_LULC_df, aes(x, y, fill = landcover))+
  geom_raster()+
  coord_fixed()+
  tidyterra::scale_fill_terrain_c()

landcover_map
```


```{r}

CD_plot <-
ggplot(den_summary, aes(x = utme, y = utmn, 
                        size = total_sessions, 
                        color = clan)) +
  geom_point(alpha = 0.7) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Hyena Den Activity",
       x = "Easting (UTM)", 
       y = "Northing (UTM)",
       size = "Total Sessions")

CD_plot

```


Put lancover map in background:
```{r}

CD_landcover_map <- ggplot()+
  #landcover map
  geom_raster(data=Mara_LULC_df, aes(x=x, y=y, fill = landcover))+
  coord_fixed()+
  scale_fill_viridis_c()+
  
  #CD plot
  geom_point(data=den_summary, aes(x = utme, y = utmn, 
                        size = total_sessions, 
                        color = clan),alpha = 0.7) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Hyena Den Activity",
       x = "Easting (UTM)", 
       y = "Northing (UTM)",
       size = "Total Sessions")

CD_landcover_map

```





