---
title: "rgee"
---


# 0. Libraries et al.

```{r Packages}

#install.packages("rgee")


# for data manipulations
library(readr) # load data
library(dplyr) # data manip

library(hyenadata) # MHP hyena data

library(tidyr) # # for data manip
library(ggplot2) # data visualization and graphs
library(stringr) 
library(lubridate) # to create my year column and month one
library(ggspatial) # to make easy maps with ggplot
library(basemaps) # for basemap function 
library(tidyverse)

#to make maps and plotting:
library(sf) 
library(mapview)
library(leaflet)
library(leaflet.extras)
library(amt)


#from Guélat
library(leaflet.extras2)
library(terra) #to work with raster data
library(lwgeom)
library(spatstat.random)
library(spdep)
library(httr2)
library(tmap)
library(tmaptools)
library(classInt)
library(leafsync)
library(qgisprocess)


# for more stuff 
library(dbscan)
library(openrouteservice)
library(geosphere)
library(magrittr)


# For rgee stuff:
library(geojsonio)
library(magick)

library(flextable)

```

```{r Load MHP tables}
#load all tables:
load_all_tables <- function(){
  data(list = data(package = 'hyenadata')$results[,3], package = 'hyenadata')
}

load_all_tables()

```


```{r Load dataframe}

RandomPointsCD_year_data <- sf::st_read("data_files/RandomPointsCD_year_data.shp")

```






# 1. Installing rgee

I will use rgee package to get different spatial covariables : 

-   Land cover (bush vs plains, etc.)\*

-   Elevation\*

-   Distance to edge of territory\*

-   Distance to water\*

-   NDVI


rgee is a package that is use to call Google Earth Engine API in R. Google Earth Engine is a cloud-based platform designed for planetary-scale environmental data analysis, offering access to a vast archive of remote sensing data and the ability to perform geospatial analysis on Google's infrastructure. While Google Earth Engine officially supports Python and JavaScript client libraries, rgee bridges this gap for R users by utilizing the reticulate package to wrap the Earth Engine Python API. This allows R users to interact with GEE and leverage its capabilities using a familiar R interface, facilitating rapid development and integration with other open-source R packages for spatial data analysis.


Setting up a google earth engine project first.

The first time I need to install conda, create the rgee environment within the conda thing. 

I followed the steps here to have it finally worked: https://gis.stackexchange.com/questions/473730/authentification-error-when-installing-rgee-in-rstudio?newreg=3674f3f594ee4f91856aa2e5df5b204d


```{r}
#to get rgeeExtra, to use things like the gif animation
#remotes::install_github("r-earthengine/rgeeExtra")

```

Here are the lines that needed to be run the first time, but now I don't need them anymore. Might need to rerun them when ee_Initialize doesn't work.
```{r }
#library(reticulate)

#install.packages("rgee")
#reticulate::conda_install(envname = "rgee", packages = "earthengine-api=0.1.370", channel = "conda-forge", python_version = "3.9")
```

```{r}
library(reticulate)

#This is to tell R to use the right Python environment
use_condaenv("rgee", required = TRUE)

# verify puthon version
py_config()
```

```{r}

#Load rgee:
library(rgee)
library(rgeeExtra)

#check ee API version
ee_check()
```

```{r}

```



Initialize rgee
```{r}
#Initialize EE 
ee_Initialize()
#important to initialize earth engine without the user and project argument here
```


This will need to be run when ee_Initialize is not working.
```{r}
#To get credentials: (need to run it once in a while when ee_Initialize doesn't work)
ee_Authenticate()

ee_clean_user_credentials()
#toforce it:
ee_Authenticate(auth_mode = "notebook")

```

## 1.1. Checking 
The ee_check() function will help for checking the sanity of the rgee installation and dependencies.

- ee_check_python() - Python version
- ee_check_credentials() - Google Drive and GCS credentials
- ee_check_python_packages() - Python packages

```{r}
ee_check()

ee_check_python()
ee_check_credentials()
ee_check_python_packages()
```



## 1.2. Tests
```{r}

# test to make sure all is running smoothly:
ee$Image("CGIAR/SRTM90_V4")$getInfo()
ee$String('Hello from the Earth Engine servers!')$getInfo()

```


### i) Night time lights
From : https://github.com/r-spatial/rgee
To Compute the trend of night-time lights

Let's create a new band containing the image date as years since 1991 by extracting the year of the image acquisition date and subtracting it from 1991.
```{r}

createTimeBand <-function(img) {
  year <- ee$Date(img$get('system:time_start'))$get('year')$subtract(1991L) 
  ee$Image(year)$byte()$addBands(img)
}

#Use your TimeBand function to map it over the night-time lights collection.

collection <- ee$
  ImageCollection('NOAA/DMSP-OLS/NIGHTTIME_LIGHTS')$
  select('stable_lights')$
  map(createTimeBand)

# Compute a linear fit over the series of values at each pixel, so that you can visualize the y-intercept as green, and the positive/negative slopes as red/blue.
col_reduce <- collection$reduce(ee$Reducer$linearFit())
col_reduce <- col_reduce$addBands(
  col_reduce$select('scale'))
ee_print(col_reduce)

# to visualize the map: 
Map$setCenter(9.08203, 47.39835, 3)
Map$addLayer(
  eeObject = col_reduce,
  visParams = list(
    bands = c("scale", "offset", "scale"),
    min = 0,
    max = c(0.18, 20, -0.18)
  ),
  name = "stable lights trend"
)

```


### ii) Worldwide elevation map
From: https://cran.r-project.org/web/packages/rgee/vignettes/rgee01.html

SRTM (Shuttle Radar Topography Mission) elevation values worldwide:
```{r}

#create a ee object
srtm <- ee$Image("USGS/SRTMGL1_003")

#Define visualization parameters
viz <- list(
  max = 4000,
  min = 0,
  palette = c("#000000","#5AAD5A","#A9AD84","#FFFFFF"))

# using Map$addLayer to visualize the map interactively
Map$addLayer(
  eeObject = srtm,
  visParams =  viz,
  name = 'SRTM')

```


### iii) Precipitation
From: https://github.com/r-spatial/rgee

To play with some precipitation data: 

nc = North Carolina counties, 100 simple features with 14 attributes
```{r}
# to read the nc shapefile:
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

plot(nc)

# will use the Terraclimate database
terraclimate <- ee$ImageCollection("IDAHO_EPSCOR/TERRACLIMATE") %>%
  ee$ImageCollection$filterDate("2001-01-01", "2002-01-01") %>%
  ee$ImageCollection$map(function(x) x$select("pr")) %>% # Select only precipitation bands
  ee$ImageCollection$toBands() %>% # from imagecollection to image
  ee$Image$rename(sprintf("PP_%02d",1:12)) # rename the bands of an image

# to extract monthly precipitation:
ee_nc_rain <- ee_extract(x = terraclimate, y = nc["NAME"], sf = FALSE)

```

ee_extract will help you to extract monthly precipitation values from the Terraclimate ImageCollection. ee_extract works similar to raster::extract, you just need to define: the ImageCollection object (x), the geometry (y), and a function to summarize the values (fun).

```{r}
# then we visualize it using ggplot:
ee_nc_rain %>%
  pivot_longer(-NAME, names_to = "month", values_to = "pr") %>%
  mutate(month, month=gsub("PP_", "", month)) %>%
  ggplot(aes(x = month, y = pr, group = NAME, color = pr)) +
  geom_line(alpha = 0.4) +
  xlab("Month") +
  ylab("Precipitation (mm)") +
  theme_minimal()
```

### iv) NDVI animation

From: https://github.com/r-spatial/rgee

To create a NDVI animation map (Peru's Arequepia region)
 
```{r}

#Define the regional bounds of animation frames and a mask to clip the NDVI data by.
mask <- system.file("shp/arequipa.shp", package = "rgee") %>%
  st_read(quiet = TRUE) %>%
  sf_as_ee()
region <- mask$geometry()$bounds()


#Retrieve the MODIS Terra Vegetation Indices 16-Day Global 1km dataset as an ee.ImageCollection and then, select the NDVI band.

col <- ee$ImageCollection('MODIS/006/MOD13A2')$select('NDVI')

#Group images by composite date
col <- col$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('day', 'year')
  img$set('doy', doy)
})
distinctDOY <- col$filterDate('2013-01-01', '2014-01-01')

```

Now, let's define a filter that identifies which images from the complete collection match the DOY from the distinct DOY collection.
```{r}
filter <- ee$Filter$equals(leftField = 'doy', rightField = 'doy')

```
Define a join and convert the resulting FeatureCollection to an ImageCollection... it will take you only 2 lines of code!

```{r}
join <- ee$Join$saveAll('doy_matches')
joinCol <- ee$ImageCollection(join$apply(distinctDOY, col, filter))
```

Apply median reduction among the matching DOY collections.

```{r}
comp <- joinCol$map(function(img) {
  doyCol = ee$ImageCollection$fromImages(
    img$get('doy_matches')
  )
  doyCol$reduce(ee$Reducer$median())
})

#define RGB visualization parameters first.
visParams = list(
  min = 0.0,
  max = 9000.0,
  bands = "NDVI_median",
  palette = c(
    'FFFFFF', 'CE7E45', 'DF923D', 'F1B555', 'FCD163', '99B718', '74A901',
    '66A000', '529400', '3E8601', '207401', '056201', '004C00', '023B01',
    '012E01', '011D01', '011301'
  )
)

#Create RGB visualization images for use as animation frames.
rgbVis <- comp$map(function(img) {
  do.call(img$visualize, visParams) %>%
    ee$Image$clip(mask)
})


#GIF visualization parameters.
gifParams <- list(
  region = region,
  dimensions = 600,
  crs = 'EPSG:3857',
  framesPerSecond = 10
)

# get month names: 
dates_modis_mabbr <- distinctDOY %>%
  ee_get_date_ic %>% # Get Image Collection dates
  '[['("time_start") %>% # Select time_start column
  lubridate::month() %>% # Get the month component of the datetime
  '['(month.abb, .) # subset around month abbreviations
```

Finally, we use ee_utils_gif_* functions to render the GIF animation and add some texts.
```{r}

animation <- ee_utils_gif_creator(rgbVis, gifParams, mode = "wb")
animation %>%
  ee_utils_gif_annotate(
    text = "NDVI: MODIS/006/MOD13A2",
    size = 15, color = "white",
    location = "+10+10"
  ) %>%
  ee_utils_gif_annotate(
    text = dates_modis_mabbr,
    size = 30,
    location = "+290+350",
    color = "white",
    font = "arial",
    boxcolor = "#000000"
  ) # -> animation_wtxt

#to save it
# ee_utils_gif_save(animation_wtxt, path = "raster_as_ee.gif")

```



## 1.3. Masai Mara geometry

Here, we will get the Masai Mara polygon, which will be use to extract the land cover, elevation, water, etc. for our used and available points in `RandomPointsCD_yearly_data`.

Here, we’ll query the WDPA protected-areas layer in GEE and filter for "Masai Mara" to avoid manually drawing or copying coordinates.


```{r Masai Mara}
# WDPA featurecollection (server-side)
wdpa <- ee$FeatureCollection("WCMC/WDPA/current/polygons")

# filter for features whose NAME contains "Masai Mara"
mara_fc <- wdpa$filter(ee$Filter$stringContains('NAME', 'Masai Mara'))

mara_geom <- mara_fc$geometry()   # keep geometry in EE
```







# 2. Land cover (rgee)

I will use the package rgee (Google Earth Engine for R), which was installed in section 1.

## 2.1. Extract land cover

Now we extract the land cover. We will use the ESA WorldCover dataset, which is a 10m global land cover with 11 classes. It is supposedely great for small-scale features like our den locations. This is going to be for 2020 landcover.

See: 
- https://developers.google.com/earth-engine/datasets/catalog/ESA_WorldCover_v100?hl=fr
- https://esa-worldcover.org/en
- https://worldcover2020.esa.int/viewer
- https://www.arcgis.com/home/item.html?id=e28b7e1da5414010ba4f47dd5a3c3ebb (for the classes code)


```{r}
# Load land cover (WorldCover 10m)
WorldCover <- ee$ImageCollection("ESA/WorldCover/v100")$first()$select("Map")

# Clip to Mara (still server-side)
MasaiMara_landcover_WC <- WorldCover$clip(mara_geom)

# quick map to see (rgee interactive map):
MasaiMara_landcover_map <- Map$addLayer(MasaiMara_landcover_WC, list(bands = "Map"), "WorldCover 10m (Map)")
MasaiMara_landcover_map

```

```{r}
# Save the map
htmlwidgets::saveWidget(MasaiMara_landcover_map, file = "Maps/MasaiMara_landcover_map.html", selfcontained = TRUE)
```


Now we sample the landcover at our used and random points :
```{r}
# Sample at your points (points_sf is your sf dataframe of den + available points)
RandomPointsCD_year_data_ee <- sf_as_ee(RandomPointsCD_year_data)

RandomPointsCD_data_sampled <- MasaiMara_landcover_WC$sampleRegions(
  collection = RandomPointsCD_year_data_ee,
  scale = 10,
  geometries = TRUE)

# Bring only the sampled results to R (much smaller)
RandomPointsCD_landcover <- ee_as_sf(RandomPointsCD_data_sampled, maxFeatures = 20000)


```

```{r}

RandomPointsCD_data <- RandomPointsCD_landcover |>
  mutate(landcover = dplyr::recode(Map,
    `10` = "Tree",
    `20` = "Shrub",
    `30` = "Grass",
    `40` = "Cropland",
    `50` = "Built-up",
    `60` = "Bare",
    `80` = "Water",
    `90` = "Wetland")) |>
  select(-Map)

```


## 2.2. Map

```{r}
# Color palette for land cover
pal_lc <- colorFactor(palette = "Set2",  # or "viridis", or custom c("green","yellow",...)
                      domain = RandomPointsCD_data$landcover)

# keeping only the used points
CD_landcover <- RandomPointsCD_data |>
  filter(case == 1)


# make a base map
CD_landcover_mapleaflet <- leaflet(CD_landcover) |>
  addTiles()

# loop over clans and add each as a separate group
for (cl in unique(CD_landcover$clan)) {
  CD_landcover_mapleaflet <- CD_landcover_mapleaflet |>
    addCircleMarkers(
      data = filter(CD_landcover, clan == cl),
      color = ~pal_lc(landcover),
      fillOpacity = 0.8,
      group = cl,
      radius = 3,
      stroke = FALSE,
      popup = ~paste("Clan:", clan, "<br>", 
                     "LC:", landcover, "<br>",
                     "Landmark:", lndmrk_))
}

# add toggle control
CD_landcover_mapleaflet <- CD_landcover_mapleaflet |>
  addLayersControl(
    overlayGroups = unique(CD_landcover$clan),
    options = layersControlOptions(collapsed = FALSE)) |>
  addLegend("bottomright", pal = pal_lc, values = ~landcover,
            title = "Land cover", opacity = 1)

CD_landcover_mapleaflet

```

```{r Save}
#save
htmlwidgets::saveWidget(CD_landcover_mapleaflet, file = "Maps/CD_landcover_mapleaflet.html", selfcontained = TRUE)

```



##### Overlay landcover with Den location:
```{r Masai Mara + CD Map}

# Visualization parameters (classes + colors)
landcover_palette <- c(
  "#006400", # Tree cover
  "#ffbb22", # Shrubland
  "#ffff4c", # Grassland
  "#f096ff", # Cropland
  "#fa0000", # Built-up
  "#b4b4b4", # Bare / sparse vegetation
  "#f0f0f0", # Snow / ice
  "#0064c8", # Permanent water bodies
  "#0096a0", # Herbaceous wetland
  "#00cf75", # Mangroves
  "#fae6a0", # Moss & lichen
  "#000000"  # No data
)

landcover_classes <- c(
  "Tree cover", "Shrubland", "Grassland", "Cropland", "Built-up",
  "Bare / sparse", "Snow / Ice", "Water", "Wetland",
  "Mangroves", "Moss/Lichen", "No data")

MasaiMara_CD_map <- MasaiMara_landcover_map |>
  # to save it with esri world imagery:
  addProviderTiles(providers$Esri.WorldImagery) |>
  addLegend(
    position = "bottomright",
    colors = landcover_palette,
    labels = landcover_classes,
    title = "ESA WorldCover 2020") |>
  addCircleMarkers(
    data = CD_landcover, 
    color = "red",
    radius = 1,
    fillOpacity = 0.9,
    popup = ~paste("Clan:", clan, "<br>Year:", year))


MasaiMara_CD_map

```

```{r}
#save
htmlwidgets::saveWidget(MasaiMara_CD_map, file = "Maps/MasaiMara_CD_map.html", selfcontained = TRUE)

```




## 2.3. Table

```{r}

landcover_table_sum <- CD_landcover |>
  st_drop_geometry() |>
  group_by(landcover) |>
  summarise(n_used = n()) |>
  arrange(desc(n_used))

landcover_table_sum

landcover_tab_sum <- flextable(landcover_table_sum)
save_as_html(landcover_tab_sum, path = "Tables/landcover_tab_sum.html")


```

Save dataframe:
```{r}

RandomPointsCD_data <- sf::st_write(RandomPointsCD_data, "data_files/RandomPointsCD_data.shp", delete_dsn = TRUE)

```




## 1.4. Elevation

We will use the rgee package again to extrat elevation of the Masai Mara. The dataset is SRTM Digital Elevation (30 m). 

We will use the `mara_geom` to extract the elevation just like I did with the land cover.

```{r}

# 1. Load SRTM DEM
srtm <- ee$Image("USGS/SRTMGL1_003")


# 2. Clip DEM to Masai Mara area
srtm_mara <- srtm$clip(mara_geom)

# 3. Extract elevation for your points
RandomPointsCD_elevation <- sf_as_ee(RandomPointsCD_data)

sampled_elevation <- srtm_mara$sampleRegions(
  collection = RandomPointsCD_elevation,
  scale = 30,      # resolution in meters
  geometries = TRUE)

# 4. Bring back to R
sampled_elevation_sf <- ee_as_sf(sampled_elevation, maxFeatures = 20000)

# 5. Merge back into your dataframe
RandomPointsCD_data <- sampled_elevation_sf

```


### 1.5.1. Map 

```{r}

elev_vis <- list(
  min = 1200,  # lowest elevation in Mara (adjust if needed)
  max = 2200,  # highest elevation in Mara
  palette = c("#e5f5e0", "#a1d99b", "#31a354"))  # light → medium → dark green


Map$centerObject(mara_geom, zoom = 9)

Elevation_map <- Map$addLayer(srtm_mara, elev_vis, "Elevation (m)")
Elevation_map
```

```{r}
#save
htmlwidgets::saveWidget(Elevation_map, file = "Maps/Elevation_map.html", selfcontained = TRUE)
```


```{r}

CD_elevation <- RandomPointsCD_data |>
  filter(case == 1)

CD_elevation_map <- Elevation_map |>
  # to save it with esri world imagery:
  addProviderTiles(providers$Esri.WorldImagery) |>
  
  addCircleMarkers(
    data = CD_landcover, 
    color = "red",
    radius = 1,
    fillOpacity = 0.9,
    popup = ~paste("Clan:", clan, "<br>Year:", year))


CD_elevation_map

```

```{r}
#save
htmlwidgets::saveWidget(CD_elevation_map, file = "Maps/CD_elevation_map.html", selfcontained = TRUE)
```



## 1.6. NDVI

```{r}
# Load MODIS NDVI (MOD13Q1, 250 m)
ndvi <- ee$ImageCollection("MODIS/061/MOD13Q1")$
  filterDate("2020-01-01", "2020-12-31")$  # adjust year if needed
  select("NDVI")$
  mean()  # average over the year

# Clip to Mara
ndvi_mara <- ndvi$clip(mara_geom)

# Visualize NDVI
ndvi_vis <- list(min = 0, max = 9000, palette = c("white","yellow","green"))


Map$centerObject(mara_geom, zoom = 9)
ndvi_map <- Map$addLayer(ndvi_mara, ndvi_vis, "NDVI")
ndvi_map
```

```{r}
htmlwidgets::saveWidget(ndvi_map, file = "Maps/ndvi_map.html", selfcontained = TRUE)
```



```{r}

# Convert your sf points to ee objects
RandomPointsCD_ndvi <- sf_as_ee(RandomPointsCD_data)

# Sample NDVI at points
sampled_ndvi <- ndvi_mara$sampleRegions(
  collection = RandomPointsCD_ndvi,
  scale = 250,       # MODIS resolution
  geometries = TRUE)

# Bring to R
sampled_ndvi_sf <- ee_as_sf(sampled_ndvi, maxFeatures = 20000)

RandomPointsCD_data <- sampled_ndvi_sf
```


```{r}
RandomPointsCD_data <- sf::st_write(RandomPointsCD_data, "data_files/RandomPointsCD_data.shp", delete_dsn = TRUE)

```



### 1.6.1. Map

```{r}

CD_ndvi <- RandomPointsCD_data |>
  filter(case == 1)

CD_ndvi_map <- ndvi_map |>
  # to save it with esri world imagery:
  addProviderTiles(providers$Esri.WorldImagery) |>
  
  addCircleMarkers(
    data = CD_ndvi, 
    color = "red",
    radius = 1,
    fillOpacity = 0.9,
    popup = ~paste("Clan:", clan, "<br>Year:", year))


CD_ndvi_map

```

```{r}
#save
htmlwidgets::saveWidget(CD_ndvi_map, file = "Maps/CD_ndvi_map.html", selfcontained = TRUE)
```


## 3. Water

### 1.7.1. Identidy water source

We will use the JRC Global Surface Water dataset. This makes polygons where there is water detected.

```{r Water Source}
# Load JRC Global Surface Water
gsw <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("occurrence")

# Permanent water mask (>90% occurrence)
water_mask <- gsw$gt(90)

# Convert water pixels to vector polygons
water_fc <- water_mask$reduceToVectors(
  geometry = mara_geom,
  geometryType = "polygon",
  scale = 30,          # MODIS/SRTM resolution is fine
  bestEffort = TRUE,
  maxPixels = 1e13)

```

```{r}
# Visualize Masai Mara boundary
Map$centerObject(mara_geom, zoom = 10)
Map$addLayer(mara_fc, list(color = "blue"), "Masai Mara Boundary")

# Add water polygons
Map$addLayer(water_fc, list(color = "cyan"), "Water Polygons")
```
Now, this gives me only the Mara River, not the Talek river. I need to find a way to get all the water sources, maybe even seasonnally? Like every month of the year or something? and also identify the areas that are marsh and use it too.



```{r}

RandomPointsCD_water <- sf_as_ee(RandomPointsCD_data)

sampled_dist_water <- dist_to_water$sampleRegions(
  collection = RandomPointsCD_water,
  scale = 10,       # match your land cover resolution
  geometries = TRUE)

sampled_dist_water_sf <- ee_as_sf(sampled_dist_water)
View(sampled_dist_water_sf)


RandomPointsCD_data <- sampled_dist_water_sf

```

























