---
output: html_document
title: "MHP Special Project - trash/unused"
author: "Anne Ju Laberge"
date: "2025"
format:
  html:
    
    code-fold: true
    output-fold: true
    code-tools: true
    code-link: true
    code-summary: "Show the code"
    theme: cosmo
    embed-resources: true
    self-contained: true
    toc-depth: 5
    toc: true
    toc-float: true
editor: visual
editor_options:
   
   chunk_output_type: console
execute:
  warning: false
  error: true
---

### OSM

OSM = OpenStreet Map it uses Overpass which is an online API. If someone mapped Talek borders in OSM over the years, I will be able to extract it yearly. After doing this function, it says that there are no boundaries for Talek, so I will just extract satellite imagery for each year and then draw my own polygons over talek town using pixels with buildings and stuff

```{r}

#install.packages("osmdata")
#install.packages("osmextract")

library(osmdata)
library(osmextract)

```

```{r}
# Install needed packages (run once)
#install.packages(c("httr2", "jsonlite"))

library(httr2)
library(sf)
library(jsonlite)

get_talek_border <- function(year) {
  date_str <- sprintf('%d-06-30T00:00:00Z', year)
  
  query <- sprintf(
    '[out:json][timeout:60][date:"%s"];
    relation["name"="Talek"]["boundary"="administrative"];
    out geom;',
    date_str
  )
  
  resp <- request("https://overpass-api.de/api/interpreter") |>
    req_body_raw(charToRaw(query)) |>
    req_perform()
  
  content_txt <- resp_body_string(resp)
  content_json <- fromJSON(content_txt, simplifyVector = FALSE)
  
  if (length(content_json$elements) == 0) {
    message("No boundary found for Talek in ", year)
    return(NULL)
  }
  
  # Build polygon
  coords_list <- lapply(content_json$elements[[1]]$geometry, function(pt) c(pt$lon, pt$lat))
  coords_mat <- do.call(rbind, coords_list)
  poly <- st_polygon(list(coords_mat))
  sfobj <- st_sf(year = year, geometry = st_sfc(poly, crs = 4326))
  return(sfobj)
}

# Test
talek_2020 <- get_talek_border(2020)
print(talek_2020)

```

## 2.4. Yearly KDE

```{r Calculating}

CD_cy_KDE <- lapply(CD_KDE_list_cy, calculate_kde_2)


# To remove the NULL element : 
CD_cy_KDE <- CD_cy_KDE[!sapply(CD_cy_KDE, is.null)]


```

To make maps with the KDE now:

```{r Extract+Data manip}

# extracting the KDE
CD_KDE_yearly <- lapply(CD_cy_KDE, extract_kde_sf)

# to R bind all the list from the extract function
CD_KDE_yearly_data <- do.call(rbind, CD_KDE_yearly)


CD_KDE_yearly_map <- tibble::rownames_to_column(CD_KDE_yearly_data, "year")

```

```{r Leaflet map}

# Add points grouped by Year
KDE_years <- unique(CD_KDE_yearly_map$year)

CD_yearly_KDE <- CD_leaflet_inter


for (yr in KDE_years) {
  KDE_year_data <- CD_KDE_yearly_map |> filter(year == yr)
  
CD_yearly_KDE <- CD_yearly_KDE |>
  addPolygons(data = KDE_year_data,
              color = "red",
              fillColor = "none",
              group = yr,
              weight = 2,
              fillOpacity = 0.05)
}


CD_yearly_KDE <- CD_yearly_KDE |>
  addLayersControl(
    baseGroups = KDE_years,
    overlayGroups = c("Year"),
    options = layersControlOptions(collapsed = FALSE))
  

CD_yearly_KDE




```

```{r spsample function}
#With MCPs
#spsample is used with sp object, to sample random points from a spatial object
rp_clans_MCPs <- spsample(clan_MCPs, n=1000, "random")
plot(rp_clans_MCPs)


# With KDEs
#st_sample is the same as spsample, but we use it with sf object, like the KDEs previously calculated are sf objects.
rp_clans_KDE <- st_sample(CD_clan_KDE, size = 1000, type ="random")

plot(rp_clans_KDE)
```

```{r}
# with MCPs
rp_MCPs_plot <- MCPs_plot |>
  addCircleMarkers(data = rp_clans_MCPs,
                   color = "red",
                   radius = 1,
                   fillOpacity = 0.1)

rp_MCPs_plot
```

```{r}
#trying to troubleshoot the 18 extra rows.

#anti_join() is used to compare 2 datasets and finding the rows that differ 
extra_rows <- anti_join(RSF_data, rp_ind_clans_KDE, by = colnames(rp_ind_clans_KDE)[1:4])

hz_essai_RSFdata <- RSF_data |>
  filter(case_ == "TRUE") |>
  filter(clan == "happyzebra")

hz_essai_presencetrack <- presence_track |>
  filter(clan == "happyzebra")

# Full join both datasets based on the first 4 columns
merged_df <- full_join(hz_essai_presencetrack, hz_essai_RSFdata, by = colnames(hz_essai_presencetrack)[1:4])


```

# Conda bullshit...

```{r}
# Tell reticulate where to install miniconda
#Sys.setenv(RETICULATE_MINICONDA_PATH = "C:/r-miniconda")

#force python.exe path into c:/r-miniconda
#use_python("C:/r-miniconda/python.exe", required = TRUE)

#see python configuration
```

To make a r-reticulate environment

```{r}
#to tell R where conda.exe is:
#conda <- "C:/r-miniconda/Scripts/conda.exe"

# to create dedicated conda environment for R-Python work
#conda_create("r-reticulate", packages = "python", conda = conda)
```

```{r}
#use_condaenv("r-reticulate", required = TRUE)

```

```{r}
# Install earthengine-api into that env
#conda_install(envname = "r-reticulate", packages = "earthengine-api", pip = TRUE, conda = conda)

```

numpy:

```{r}
reticulate::py_install("numpy", envname = "r-reticulate", method = "conda")
```

Setting up a google earth engine project first:

Installing mini_conda:

```{r}
library(reticulate)

reticulate::conda_install(envname = "rgee", packages = "earthengine-api=0.1.370", channel = "conda-forge", python_version = "3.9")

#Tell R to use the right python in the right environment 
use_condaenv("r-reticulate", 
             conda = "C:/r-miniconda/Scripts/conda.exe", 
             required = TRUE)

# to see the python configuration and make sure all is good 
py_config()

#make sure we have earth engine installed and all
py_module_available("ee")
#TRUE
```

```{r}

#install.packages("rgee")
library(rgee)

Sys.setenv(EARTHENGINE_CREDENTIALS = file.path(Sys.getenv("CONDA_PREFIX"), ".config/earthengine/credentials"))

#ee_install_set_pyenv(py_env = "YOUR_ENV", py_path = "C:/r-miniconda/envs/r-reticulate/python.exe")

#copying my credentials :
#file.copy(
#  from = "C:/Users/Anne Ju Laberge/.config/earthengine/credentials",
#  to   = "C:/r-miniconda/envs/r-reticulate/.config/earthengine/credentials",
#  overwrite = TRUE)


```

```{r}
#initialize earth engine so that it is linked to my google cloud account in the right project 
ee_Initialize(user = 'anneju.laberge21@gmail.com', drive = TRUE, project = 'ee-annejulaberge21')

ee_Initialize()

ee_clean_user_credentials()
ee_Authenticate()

#rgee::ee_install_upgrade()

ee_check_credentials()
ee_check()

```

```{r}
# to see where the credentials are stored:
ee_get_earthengine_path()


```

```{r}


#in cmd:
#C:\Users\Anne Ju Laberge\rgee-env\Scripts\activate"

# Point to the exact Python binary in your working env
use_python("C:/Users/Anne Ju Laberge/rgee-env/Scripts/python.exe", required = TRUE)


ee_Initialize(user = 'anneju.laberge21@gmail.com', drive = TRUE, project = 'ee-annejulaberge21')
```

```{r}
#check credentials, all look ok
ee_check_credentials()

# Now we make sure that all Earth engine python packages are installed
ee_check()
ee_check_python_packages()

```

### rstac

connect to STAC APIs (like Microsoft Planetary Computer) and download imagery locally.

This was not working, I used the rgee package instead...

```{r}
library(rstac)
library(terra)
library(sf)

```

Step1: define Talek region:

```{r}
# Talek town borders : 
talek_bbox <- c(35.0, -1.42, 35.05, -1.39) 
# xmin, ymin, xmax, ymax
talek_bbox
```

Step2: Connect to a STAC endpoint:

```{r}
stac_url <- "https://planetarycomputer.microsoft.com/api/stac/v1"
s <- stac(stac_url)
```

Step3: Search for one year of Sentinel-2 imagery

```{r}
search_res <- s |>
  stac_search(
    collections = "sentinel-2-l2a",
    bbox = talek_bbox,
    datetime = "2020-01-01/2020-12-31",
    limit = 10
  ) |>
  get_request()
```

Step4: Get asset URLs and download

```{r}

items <- items_fetch(search_res)
assets <- assets_select(items, asset_names = "visual") # true color composite

asset_urls <- assets_url(assets)
#Download first image
stac_download(
  assets,
  output_dir = "talek_images",   # creates folder if missing
  overwrite = TRUE
)
```

Step5: View & digitize

```{r}
img <- rast("talek_2020.tif")
plotRGB(img, r = 1, g = 2, b = 3, stretch = "lin")

```

##i) Prey abundance

Will use the Prey census data. We have 2 prey transects per month. Maybe we can do that per month, and just count the total number of preys per month. That would give us an estimate of the monthly abundance of prey, which is good because prey abundance does vary according to season, which in our case is represented by the month.

```{r Prey census data}
str(tblPreyCensus)

#transforming all the prey columns as numeric
prey_data <- tblPreyCensus |>
  mutate_at(vars(9: 39), as.numeric)

str(prey_data)
unique(prey_data$region)
```

Now we want to keep months and the total number of preys per the month.

```{r PREY_dat}

prey_data_sum <- prey_data |>
  #Create a column with the sum of all the prey columns (9 - 27 are preys)
  mutate(prey = rowSums(prey_data[9:27])) |>
  mutate(lstock = rowSums(prey_data[38:39])) |>
  # keeping only the talek clans
  filter(region == "Narok") |>
  #adding a month columns
  mutate(month = month(date)) |>
  #adding a year column
  mutate(year = year(date)) |>
  #keeping only year >=1988
  filter(year >= 1988) |>
  #now we can get rid of all the prey columns 
  select(-c(4:8, 9:36, 38, 39))

#now we want to create a new dataset with monthly prey abundance
#So we want to group by year and transect
PREY_dat <- prey_data_sum |>
  arrange(year, month) |>
  group_by(year, month, clan, transect) |>
  summarise(prey_abun = sum(prey), 
            lstock_abun = sum(lstock)) |>
  ungroup()


PREY_dat$year <- as.character(PREY_dat$year)
PREY_dat$month <- as.numeric(PREY_dat$month)


```

Visualisation

```{r Plot per year}

#the total amount for all the years combined 
PREY_plot_year <- 
  ggplot(aes(x=transect, y=prey_abun), data = PREY_dat) +
  geom_col(fill = "#F8766D", width = 0.6) +
  labs(x = "Transect", y = "Prey abundance") +
  facet_wrap(~month) +
  theme_bw() +
  theme(legend.position = "none")

PREY_plot_year


# if we select specific years
PREY_plot <- PREY_dat |>
  #Choose a year
  filter(year == 2022) |>
  
  #Now plot: 
  ggplot(aes(x=transect, y=prey_abun)) +
  geom_col(aes(fill = as.factor(month)), width = 0.6) +
  labs(x = "Transect", y = "Prey abundance") +
  facet_wrap(~month) +
  theme_bw() + 
  theme(legend.position = "none")
  

PREY_plot



```

```{r Plot all}

#I will combine all transect together...
PREY_dat_all <- PREY_dat |>
  group_by(year, month) |>
  summarise(prey_abun_tot = sum(prey_abun),
            lstock_abun_tot = sum (lstock_abun)) |>
  ungroup()

# now we plot:
PREY_plot_all <- 
  ggplot(aes(x=year, y=prey_abun_tot), data = PREY_dat_all) +
  geom_col(aes(fill = as.factor(month)), width = 0.6) +
  labs(x = "Year", y = "Prey abundance") +
  facet_wrap(~month) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust=0.7, size = 7), 
        legend.position = "none")

PREY_plot_all

```

Combine the datasets

```{r Merge DB}

CD_data_prey <- merge(CD_for_map, PREY_dat_all, by = c("month", "year")) 

```

## Data cleaning (GPS points):

```{r}
DB_locations_for_KDEs_clean <- DB_locations_for_KDEs |>
  
  #Change the utme coordinates that are 111111 and 1111111
  left_join(DF_lndmk %>% select(landmark_id, utme_new = utme, utmn_new = utmn),
            by = "landmark_id", relationship = "many-to-many") |>
  mutate(
    utme = if_else(utme == "11111", utme_new, utme),
    utmn = if_else(utme == "11111", utmn_new, utmn)) |>
  mutate(
    utme = if_else(utme == "111111", utme_new, utme),
    utmn = if_else(utme == "111111", utmn_new, utmn)) |>
  mutate(
    utme = if_else(utme == "1111111", utme_new, utme),
    utmn = if_else(utme == "1111111", utmn_new, utmn)) |>
  # now same for the 1111111 in utmn:
  mutate(
    utme = if_else(utmn == "1111111", utme_new, utme),
    utmn = if_else(utmn == "1111111", utmn_new, utmn)) |>
  # change when I have 0s in utme and/or utmn
  mutate(
    utme = if_else(landmark_id == "Stupid_Tree" & utme == 0 & utmn == 0,
                   utme_new, utme),
    utmn = if_else(landmark_id == "Stupid_Tree" & utme == 0 & utmn == 0,
                   utmn_new, utmn)) |>
  mutate( # session t99428
    utme = if_else(landmark_id == "Stupid_Tree" & utmn == "0", utme_new, utme),
    utmn = if_else(landmark_id == "Stupid_Tree" & utmn == "0", utmn_new, utmn)) |>
  #change the utme value of 98... of Fig_Tree_Lugga_-_Top
  mutate(
    utme = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == 9840057, utme_new, utme),
    utmn = if_else(landmark_id == "Fig_Tree_Lugga_-_Top" & utme == 9840057, utmn_new, utmn)) |>
  # now we remove the utme and utmn new columns
  select(-utme_new, -utmn_new) # |>
  
  #remove the 0 in from of Uwanja_Den_2
  mutate(
    utme = if_else(
      landmark_id == "Uwanja_Den_2" & grepl("^0", utme),
      sub("^0", "", utme),
      utme)) |>
  #remove the last digit of a 7 digit utme
  mutate(
    utme = if_else(
      landmark_id == "Stupid_Tree" & nchar(as.character(utme)) == 7,
      substr(as.character(utme), 1, 6),   # keep only first 6 digits
      as.character(utme))) |>
  # change the airstrip_den utme and utmn value with a comma
  mutate(
    utme = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utme)),   # take everything after the comma
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Airstrip_Den",
      sub(".*,", "", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>
  # same for the TWin_trees values
  mutate(
    utme = if_else(
      landmark_id == "Twin_Trees",
      sub(".*,(.*)$", "\\1", as.character(utme)),   # keep last number
      as.character(utme)),
    utmn = if_else(
      landmark_id == "Twin_Trees",
      sub(".*,(.*)$", "\\1", as.character(utmn)),   # same for utmn
      as.character(utmn))) |>
  #for session t117205, add a 4 after the first 7, and session t111704 add a 5 after the 7
  mutate(
    utme = if_else(
      landmark_id == "Laura_Tree" & utme == "79731",
      sub("^7", "74", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Sunshine_Tree" & utme == "70360",
      sub("^7", "75", utme),
      utme)) |>
  # change the session t99585 and t60813 and swap 4 and 7 in utme
  mutate(
    utme = if_else(
      landmark_id == "Stupid_Tree" & utme == "476966",
      sub("^47", "74", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Green_Den" & utme == "477420",
      sub("^47", "74", utme),
      utme)) |>
  #same for session and change 5 and 7
   mutate(
    utme = if_else(
      landmark_id == "Croton_Island-East_Edge" & utme == "570527",
      sub("^57", "75", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Cow_Tree" & utme == "570804",
      sub("^57", "75", utme),
      utme)) |>
  mutate(
    utme = if_else(
      landmark_id == "Hi_Step_X" & utme == "573282",
      sub("^57", "75", utme),
      utme)) |>
  
  #take out the Low_Ranker's_Crossing cause utmn= utme
  filter(landmark_id != "Low_Ranker's_Crossing") |>
  
  #change the utmn starting with 6 to 9
  mutate(utmn = if_else(
      grepl("^6", utmn),
      sub("^6", "9", utmn),
      utmn)) |>
  mutate(utmn = if_else(
      grepl("^7", utmn),
      sub("^7", "9", utmn),
      utmn)) |>
  #remove the 8th digit if there is one
  mutate(utmn = if_else(
      nchar(utmn) == 8,
      substr(utmn, 1, 7),  # keep only the first 7 characters
      utmn)) |>
  #take out the session t54706 because utmn = 99 and the point is ~700m from the landmark
  filter(session != "t54706") |>
  filter(nchar(utmn) != 6)
```

## MCPs with only the dens locations (we decided to use all the GPS locations anytime a hyena was seen on their respective territory)

```{r Individual MCPs for all clans}

#for Serena South clan
CD_sf_serenas <- CD_sf_wgs84 |>
  filter(clan == "serenas")

serenas_points <- as_Spatial(CD_sf_serenas$geometry)

serenas_MCP <- mcp(serenas_points, percent = 95)



# For Serena North clan
CD_sf_serenan <- CD_sf_wgs84 |>
  filter(clan == "serenan")

serenan_points <- as_Spatial(CD_sf_serenan$geometry)

serenan_MCP <- mcp(serenan_points, percent = 95)



# For Happy Zebra clan 
CD_sf_hz <- CD_sf_wgs84 |>
  filter(clan == "happyzebra")

hz_points <- as_Spatial(CD_sf_hz$geometry)

hz_MCP <- mcp(hz_points, percent = 95)


# For Talek clan 
CD_sf_talek <- CD_sf_wgs84 |>
  filter(clan == "talek")

talek_points <- as_Spatial(CD_sf_talek$geometry)

talek_MCP <- mcp(talek_points, percent = 95)

```

```{r Plots}
plot(serenan_MCP)
plot(serenas_MCP)
plot(hz_MCP)
plot(talek_MCP)

```

Long way to make the MCPs

```{r Clan MCPs}

#for Serena South clan
CD_sf_serenas <- DB_AllGPS_sf_wgs84 |>
  filter(clan == "serenas")

serenas_points <- as_Spatial(CD_sf_serenas$geometry)

serenas_MCP <- mcp(serenas_points, percent = 95)



# For Serena North clan
CD_sf_serenan <- DB_AllGPS_sf_wgs84 |>
  filter(clan == "serenan")

serenan_points <- as_Spatial(CD_sf_serenan$geometry)

serenan_MCP <- mcp(serenan_points, percent = 95)



# For Happy Zebra clan 
CD_sf_hz <- DB_AllGPS_sf_wgs84 |>
  filter(clan == "happyzebra")

hz_points <- as_Spatial(CD_sf_hz$geometry)

hz_MCP <- mcp(hz_points, percent = 95)


# For Talek clan 
CD_sf_talek <- DB_AllGPS_sf_wgs84 |>
  filter(clan == "talek")

talek_points <- as_Spatial(CD_sf_talek$geometry)

talek_MCP <- mcp(talek_points, percent = 95)

```

```{r Visualization}

plot(serenan_MCP)
plot(serenas_MCP)
plot(hz_MCP)
plot(talek_MCP)

```

```{r}
#grouping the mcps in a dataframe grouped by clans
clan_MCPs <- rbind(serenan_MCP, talek_MCP, serenas_MCP, hz_MCP)

#rename the ids :
new_ids <- c("serenan", "talek", "serenas", "happyzebra")

clan_MCPs <- sp::spChFIDs(clan_MCPs, new_ids)

```

Here, I will do rp datasets for the 4 clans, to make sure the number of available points is 10 times de number of used points.

```{r RP Per clans}

#First, how much CD locations have each clans
sum(CD_sf_wgs84$clan == "serenan") # 93
sum(CD_sf_wgs84$clan == "serenas") # 92
sum(CD_sf_wgs84$clan == "happyzebra") # 83
sum(CD_sf_wgs84$clan == "talek") # 90

#serenan
rp_serenan_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "serenan") |>
  random_points(n=930, presence = presence_track_serenan) |>
  mutate(clan = "serenan")


#serenas
rp_serenas_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "serenas") |>
  random_points(n=920, presence = presence_track_serenas) |>
  mutate(clan = "serenas")

#happyzebra
rp_hz_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "happyzebra") |>
  random_points(n = 830, presence = presence_track_hz) |>
  mutate(clan = "happyzebra")

#talek
rp_talek_KDE <- CD_clan_KDE_mapdata |>
  subset(clan == "talek") |>
  random_points(n = 900, presence = presence_track_talek) |>
  mutate(clan = "talek")


# combine all of the clans:
rp_ind_clans_KDE <- rbind(rp_serenan_KDE, rp_serenas_KDE, rp_hz_KDE, rp_talek_KDE)

```

Save the rp dataset:

```{r Save}

rp_ind_clans_KDE <- st_write(rp_ind_clans_KDE, "data_files/rp_ind_clans_KDE.shp")

```

```{r Plot leaflet}

pal_clans_rp <- colorFactor(palette = "magma", domain = rp_ind_clans_KDE$clan)

rp_KDEs_plot <- CD_clan_KDE_map |>
  addCircleMarkers(data = rp_ind_clans_KDE,
                   lat = rp_ind_clans_KDE$y_,
                   lng = rp_ind_clans_KDE$x_,
                   group = rp_ind_clans_KDE$clan,
                   color = ~pal_clans_rp(clan),
                   radius = 1,
                   fillOpacity = 0.1)

rp_KDEs_plot

htmlwidgets::saveWidget(rp_KDEs_plot, file = "Maps/rp_KDEs_plot.html", selfcontained = TRUE)

```

# 2. KDE by clans

Here we will make KDE, kernel density estimation.

## 2.1. Functions

First, the functions we'll need: - Function to calculate the KDE - Function to calculate the KDE with two groups (clan per year example) - Function to extract the KDE values and then use it in maps.

```{r Function KDE for clans}
# Function to calculate 50% KDE for each clan
calculate_kde <- function(df) {
  hr_kde(df, levels = 0.95)
}
```

```{r Function KDE for yearly}

# Function to calculate 95% KDE for each group
calculate_kde_2 <- function(df) {
  if (nrow(df) < 5) return(NULL)  # Skip if not enough points
  df <- df[complete.cases(df$x_, df$y_), ]  # Remove rows with missing coords
  if (nrow(df) < 5) return(NULL)  # Skip again if filtering removed too much
  tryCatch(
    hr_kde(df, levels = 0.95)
  )
}

```

```{r Function to extract KDE}
# Extract KDE polygons and convert to sf format
extract_kde_sf <- function(kde_result) {
  kde_sf <- st_as_sf(hr_isopleths(kde_result))# Convert KDE to sf polygons
  return(kde_sf)
}
```

## 2.2. Data

```{r Organising data}
# First step : make_track
# Make_track the GPS dataset to then be able to calculate KDE

#Preparing the df for it:
CD_sf_df <- CD_sf_wgs84 |>
  # before making the make_track, we need to separate geometry column into long and lat 
  mutate(long = sf::st_coordinates(geometry)[,1],
         lat = sf::st_coordinates(geometry)[,2]) |>
  mutate(row.id = row_number()) |>
  dplyr::select(year, long, lat, clan, row.id)


#Now we make the make_track
CD_KDE_df <- CD_sf_df |>
  #Now we do the make_track function
  make_track(long, lat, crs = 4326, id = clan) |>
  #Now rename the id column just because
  rename(clan = id) |>
  mutate(row.id = CD_sf_df$row.id)
  
#Now to add year:
CD_year <- CD_sf_df |>
  dplyr::select(year, row.id) |>
  st_drop_geometry()

#merge year with the lat long df:
CD_KDE_df <- CD_KDE_df |>
  left_join(CD_year, by = "row.id")
```

```{r Split}
# Split the dataset per Clan
CD_KDE_list <- split(CD_KDE_df, CD_KDE_df$clan)

# Split the db into clan and year
CD_KDE_list_cy <- split(CD_KDE_df, CD_KDE_df$year)
```

## 2.3. Clans' KDE

```{r Calculating}
CD_clan_KDE <- lapply(CD_KDE_list, calculate_kde)
```

```{r Extract+Data manip}
CD_clan_KDE_data <- lapply(CD_clan_KDE, extract_kde_sf)

CD_clan_KDE_mapdata <- do.call(rbind, CD_clan_KDE_data)

CD_clan_KDE_mapdata <- tibble::rownames_to_column(CD_clan_KDE_mapdata, "clan")



```

```{r Leaflet}

pal_clan <- colorFactor(palette = "viridis", domain = CD_clan_KDE_mapdata$clan)


CD_clan_KDE_map <- CD_leaflet_year |>
  addPolygons(data = CD_clan_KDE_mapdata,
              fillColor = ~pal_clan(CD_clan_KDE_mapdata$clan),
              color = "black",
              weight = 1,
              fillOpacity = 0.8,
              label = ~paste("Clan:", clan)) 
  

CD_clan_KDE_map

```

```{r}
htmlwidgets::saveWidget(CD_clan_KDE_map, file = "Maps/CD_clan_KDE_map.html", selfcontained = TRUE)
```

Save KDEs data

```{r Save}

CD_clan_KDE_mapdata <- sf::st_write(CD_clan_KDE_mapdata, "data_files/CD_clan_KDE.shp")


```

## MCP (old way)

```{r Random_points; old way}
#serenan 
presence_track_serenan <- presence_track |> subset(clan == "serenan") 

#serenas 
presence_track_serenas <- presence_track |> subset(clan == "serenas") 

#hz 
presence_track_hz <- presence_track |> subset(clan == "happyzebra") 

#talek 
presence_track_talek <- presence_track |> subset(clan == "talek") 


# make MCP valid:
mcp95_clan_sf <- st_make_valid(mcp95_clan_sf)


#creating the dataset to use in RSF after (for all clans) 
rp_clans_MCP <- random_points(mcp95_clan_sf, n=4000, presence = presence_track) 

#Visualising 
plot(rp_clans_MCP)

```
